### 2018-03-08 cuckoo_plots.Rmd
Initial update to project

---
title: "Cuckoo movements"
author: "MM"
date: "7 February 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cuckoos, echo=FALSE, warning=FALSE, message=FALSE}
########################

rm(list=ls())
library(maps)
library(sp)
library(maptools)
library(rgdal)
library(ggplot2)
library(ggmap)

# using best of day data
dat2<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_stopovers.csv", h=T)

# hack to put months in correct numeric order
dat2$month2<-ifelse(dat2$month>9, dat2$month, paste(0, dat2$month, sep=""))
dat2$yrmo<-as.numeric(paste(dat2$year.x, dat2$month2, sep="."))

m1 <- get_map(location = "Nigeria", zoom=4)

for(i in unique(dat2$ptt))
{

  g<-ggmap(m1)+
    geom_line(data=dat2[dat2$ptt==i,], aes(x=long, y=lat,
                                             colour=factor(yrmo)))+
    geom_jitter(data=dat2[dat2$ptt==i,], aes(x=long, y=lat,
                colour=factor(yrmo)), shape=1, width = 1, height= 1)+ggtitle(i)

  print(g)
}

```

### 2018-03-08 cuckoo_compile_clean.R

@@ -1,0 +1,309 @@
+##########################################################
+#
+#  Cuckoo ORIGINAL data - tidying source file, adding new variables (distance moved between points, bearing, movement groups, and movement types)
+#
+#  Samantha Franks
+#	15 Nov 2013
+# 4 Dec 2013
+# 23 Dec 2013
+# 6 May 2015 - revised to include 2014 cohort and working from original raw data messages
+# August 2017 - modified by Jenni Border
+
+
+# Jan 2018 - script lifted by MM from above paper trail
+
+##########################################################
+
+
+
+
+library(lubridate)
+library(plyr)
+library(ggplot2)
+library(ggmap)
+library(maps)
+
+
+library(rgdal)
+library(maptools)
+library(shape)
+library(splancs)
+library(geosphere)
+library(adehabitatHR) # check this version works ok for what I need to use it for
+###sam used adehabitats but that package is now depreciated and so use one of the new versions (adehabitatHR, adehabitatLT, adehabitatMA)
+library(sp) # converts R dataframes into spatially explicit dta
+library(rgeos)
+library(raster)
+library(chron)
+library(lubridate)
+library(plyr)
+
+
+####======================== LOAD DATA ======================================
+
+# cuckoo coordinate locations are in Lat-Long and UTM coordinates and WGS84 datum
+dat<-read.csv("N:/cuckoo_tracking/data/cuckoos.csv", h=T)
+
+head(dat)
+
+#---- remove lines with blank location_dates ----
+d1 <- dat[-which(is.na(dat$location_date)),]
+
+#-------------- convert UNIX timestamp to R date format timestamp --------------
+d1$timestamp <- as.POSIXct(as.numeric(as.character(d1$location_date)),origin="1970-01-01",tz="UTC")
+
+#-------------- Add julian date, year, month, day data to locations ----------------
+d1$year <- year(d1$timestamp)
+d1$month <- month(d1$timestamp)
+d1$day <- mday(d1$timestamp)
+d1$julian <- yday(d1$timestamp)
+
+table(d1$year)
+
+
+
+##remove all records from 2017 and 2018
+d2a<-subset(d1, year!=2017)
+d2a<-subset(d1, year!=2018)
+##check this has worked
+summary(d1)
+summary(d2a)
+
+# load the reference spreadsheet with individual cuckoo specs, includes up to the 2014 cohort
+# merge with raw dataset
[…]

### 2018-03-08 cuckoo_compile_clean_movebank.R

@@ -1,0 +1,339 @@
+##########################################################
+#
+#  Cuckoo ORIGINAL data - tidying source file, adding new variables (distance moved between points, bearing, movement groups, and movement types)
+#
+#  Samantha Franks
+#	15 Nov 2013
+# 4 Dec 2013
+# 23 Dec 2013
+# 6 May 2015 - revised to include 2014 cohort and working from original raw data messages
+# August 2017 - modified by Jenni Border
+
+
+# Jan 2018 - script lifted by MM from above paper trail
+
+##########################################################
+
+
+
+
+library(lubridate)
+library(plyr)
+library(ggplot2)
+library(ggmap)
+library(maps)
+
+
+library(rgdal)
+library(maptools)
+library(shape)
+library(splancs)
+library(geosphere)
+library(adehabitatHR) # check this version works ok for what I need to use it for
+###sam used adehabitats but that package is now depreciated and so use one of the new versions (adehabitatHR, adehabitatLT, adehabitatMA)
+library(sp) # converts R dataframes into spatially explicit dta
+library(rgeos)
+library(raster)
+library(chron)
+library(lubridate)
+library(plyr)
+
+
+####======================== LOAD DATA ======================================
+
+# cuckoo coordinate locations are in Lat-Long and UTM coordinates and WGS84 datum
+#dat<-read.csv("N:/cuckoo_tracking/data/movebank_cuckoos_hybridfilter.csv", h=T)
+dat<-read.csv("~/BTO/cuckoo_tracking/data/movebank_cuckoos_hybridfilter.csv", h=T)
+
+
+head(dat)
+
+#---- remove lines with blank location_dates ---- NONE in movebank
+#d1 <- dat[-which(is.na(dat$timestamp)),]
+
+#-------------- convert UNIX timestamp to R date format timestamp --------------
+#d1$timestamp <- as.POSIXct(as.numeric(as.character(d1$location_date)),origin="1970-01-01",tz="UTC")
+
+d1<-dat
+d1$timestamp <- as.POSIXct(strptime(substr(dat$timestamp,1,19), "%Y-%m-%d %H:%M:%S"), "UTC")
+
+#-------------- Add julian date, year, month, day data to locations ----------------
+d1$year <- year(d1$timestamp)
+d1$month <- month(d1$timestamp)
+d1$day <- mday(d1$timestamp)
+d1$julian <- yday(d1$timestamp)
+
+table(d1$year)
+
+# have a look
+
+
[…]

### 2018-03-08 cuckoo_compile_clean_movebank_30Jan.R

@@ -1,0 +1,512 @@
+##########################################################
+#
+#  Cuckoo ORIGINAL data - tidying source file, adding new variables (distance moved between points, bearing, movement groups, and movement types)
+#
+#  Samantha Franks
+#	15 Nov 2013
+# 4 Dec 2013
+# 23 Dec 2013
+# 6 May 2015 - revised to include 2014 cohort and working from original raw data messages
+# August 2017 - modified by Jenni Border
+
+
+# Jan 2018 - script lifted by MM from above paper trail
+
+##########################################################
+
+rm(list=ls())
+
+
+library(lubridate)
+library(plyr)
+library(ggplot2)
+library(ggmap)
+library(maps)
+
+
+library(rgdal)
+library(maptools)
+library(shape)
+library(splancs)
+library(geosphere)
+library(adehabitatHR) # check this version works ok for what I need to use it for
+###sam used adehabitats but that package is now depreciated and so use one of the new versions (adehabitatHR, adehabitatLT, adehabitatMA)
+library(sp) # converts R dataframes into spatially explicit dta
+library(rgeos)
+library(raster)
+library(chron)
+library(lubridate)
+library(plyr)
+
+
+####======================== LOAD DATA ======================================
+
+# cuckoo coordinate locations are in Lat-Long and UTM coordinates and WGS84 datum
+
+#dat<-read.csv("N:/cuckoo_tracking/data/movebank_cuckoos_hybridfilter.csv", h=T)
+dat<-read.csv("~/BTO/cuckoo_tracking/data/movebank_cuckoos_hybridfilter.csv", h=T)
+
+#dat<-read.csv("N:/cuckoo_tracking/data/movebank_cuckoos_hybridfilter_bestofday.csv", h=T)
+
+head(dat)
+
+#---- remove lines with blank location_dates ---- NONE in movebank
+#d1 <- dat[-which(is.na(dat$timestamp)),]
+
+#-------------- convert UNIX timestamp to R date format timestamp --------------
+#d1$timestamp <- as.POSIXct(as.numeric(as.character(d1$location_date)),origin="1970-01-01",tz="UTC")
+
+d1<-dat
+d1$timestamp <- as.POSIXct(strptime(substr(dat$timestamp,1,19), "%Y-%m-%d %H:%M:%S"), "UTC")
+
[…]

### 2018-03-08 cuckoo_stopovers.R

@@ -1,0 +1,211 @@
+########################
+# 31/01/18
+# Defining cuckoo stopover sites
+# using convex hull method
+########################
+
+rm(list=ls())
+
+library(ggplot2)
+library(ggmap)
+library(rgeos)
+library(maps)
+library(maptools)
+
+
+# Read in processed cuckoo data with stopover regions defined
+# using Sam & Jenni's code
+
+dat<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_stopovers.csv", h=T)
+
+# FORMAT!!
+dat$timestamp <- as.POSIXct(strptime(dat$timestamp, "%Y-%m-%d %H:%M:%S"), "UTC")
+
+
+# OK so stopover information provided in this dataset is:
+# tcycle, which gives an identifier for the the 10 hr on transmission cycle, per bird
+# days.btwn.trans, which gives the time (in days) between transmissions (rows)
+# distance, which gives the distance (km) between the fix (row) and the last
+# bearing, which gives the bearing (deg) between the fix (row) and the last
+# mtype, which is movement type. if more than 50km from previous fix, then
+# type is M (migration) otherwise it is S (stationary).
+# mgroup and m_group_replicate, which is the movement group. Fixes are grouped
+# if they do not have a >50km distance between points
+# LOS, which is the length of stopover (days) assigned to each mgroup.
+
+m1 <- get_map(location = c(lon = mean(dat$long),
+                           lat = mean(dat$lat)), zoom=3)
+g <- ggmap(m1) +geom_point(data =dat, aes(x=long, y=lat, colour=mtype))+
+  facet_wrap(~year.x)
+g
+
+# have a look at West African region
+
+m1 <- get_map(location = "Ghana", zoom=5)
+g <- ggmap(m1) +geom_point(data =dat, aes(x=long, y=lat, colour=mtype))+
+  facet_wrap(~year.x)
+g
+
+# Ok so lets build a loop that makes a convex hull around each group of stopover points
+# attributes will be: ptt, start timestamp, end timestamp, ndays, month, year
+
+stopovers_spdf<-NULL
+
+birds<-unique(dat$ptt)
+
+for(i in birds)
+  {
[…]

### 2018-03-08 cuckoo_winter_2_summer_subset.R

@@ -1,0 +1,161 @@
+########################
+# 02/02/18
+# Subsetting data to West Africa region using
+# dates from wintering to summer return to UK
+# Test hypotheses about timing of spring migration
+# movement, W. African stopovers and return to UK
+########################
+
+rm(list=ls())
+library(maps)
+library(sp)
+library(maptools)
+library(rgdal)
+library(ggplot2)
+library(ggmap)
+
+# using best of day data
+dat<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_stopovers.csv", h=T)
+
+# FORMAT!!
+dat$timestamp <- as.POSIXct(strptime(dat$timestamp, "%Y-%m-%d %H:%M:%S"), "UTC")
+
+# subset from wintering period to first landfall in UK,
+# first limit is a time boundary, second is a geographical overlap
+
+# first subset to points in UK
+setwd("~/BTO/cuckoo_tracking/sourced_data/")
+world<-readOGR(layer="TM_WORLD_BORDERS-0.3",
+            dsn="country_borders") # different linux/windows
+
+head(world@data)
+
+uk<-world[world$NAME=="United Kingdom",]
+
+plot(world);plot(uk, add=T, col=2)
+
+datSP<-SpatialPointsDataFrame(SpatialPoints(cbind(dat$long, dat$lat), 
+      CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")),
+      data=dat)
+
+plot(uk); plot(datSP, add=T)
+plot(datSP[uk,], col=2, add=T)
+
+# bit of a hack to get rows that sit in uk
+datSP$rownum<-1:nrow(dat)
+datSP$inUK<-"N"
+ukptz<-datSP[uk,]
+# back to data.frame
+dat2<-datSP@data
+dat2[ukptz$rownum,]$inUK<-"Y"
+#make sure it worked
+
+plot(lat~long, dat2[dat2$inUK=="Y",])
+map("world", add=T)
+points(lat~long, dat2[dat2$inUK=="N",], col=2)
+# ok
+dat2$rownum<-NULL
+
+# Now to figure out when wintering starts and ends
+
+qplot(data=dat2[dat2$lat<20,], x=lat, geom="histogram", binwidth=2)
[…]

### 2018-03-08 west_africa_timing.R

@@ -1,0 +1,391 @@
+########################
+# 13/02/18
+# Investigating wintering and west africa stopover movements
+########################
+
+rm(list=ls())
+
+library(ggplot2)
+
+library(readxl)
+
+# read the xlsx file as it has manually corrected country column
+# and has the mig_cohort column added
+dat<-read_excel("~/BTO/cuckoo_tracking/data/stopover_table_bestofday_biomes_Chris.xlsx")
+
+
+# these cuckoos complete at least one annual migration
+gd_cucks<-c(62608,62688,115586,115591,115594,
+            115602,121792,128297,
+            128300,128302,134951,134952,
+            134955,134956,134957,134963,
+            146754,146757,146758,146759,146760,
+            161318,161321,161324)
+
+gd_cucks %in% unique(dat$ptt)  
+
+dat<-subset(dat, dat$ptt %in% gd_cucks)
+
+dat$stage<-"migration"
+dat[dat$country=="United Kingdom",]$stage<-"breeding"
+
+# remove migration cohorts from multi-year birds 
+# that do not complete the full cycle
+
+dat$complete_mig<-"Y"
+
+dat[dat$ptt==62608 & dat$mig_cohort==2015,]$complete_mig<-"N"
+dat[dat$ptt==62688 & dat$mig_cohort==2012,]$complete_mig<-"N"
+dat[dat$ptt==115586 & dat$mig_cohort==2015,]$complete_mig<-"N"
+dat[dat$ptt==115591 & dat$mig_cohort==2014,]$complete_mig<-"N"
+dat[dat$ptt==115594 & dat$mig_cohort==2016,]$complete_mig<-"N"
+dat[dat$ptt==115602 & dat$mig_cohort==2013,]$complete_mig<-"N"
+dat[dat$ptt==128297 & dat$mig_cohort==2014,]$complete_mig<-"N"
+dat[dat$ptt==128300 & dat$mig_cohort==2014,]$complete_mig<-"N"
+dat[dat$ptt==128302 & dat$mig_cohort==2014,]$complete_mig<-"N"
+dat[dat$ptt==134952 & dat$mig_cohort==2015,]$complete_mig<-"N"
+dat[dat$ptt==134955 & dat$mig_cohort==2015,]$complete_mig<-"N"
+dat[dat$ptt==134956 & dat$mig_cohort==2015,]$complete_mig<-"N"
[…]

### 2018-03-09 west_africa_timing.R
Added loop to calc arrival on breeding grounds

@@ -121,10 +121,67 @@ UK_travel$unreliable_exit<-unlist(by(UK_travel, UK_travel$ptt, FUN=function(x) {
 
 write.csv(UK_travel, "~/BTO/cuckoo_tracking/data/complete_cycles_UK_timing.csv", quote=F, row.names=F)
 
+# OK we need better than just arrival into the UK, no add arrival at breeding ground
+# this is defined as a 10km buffer around breeding stopovers from the best of day data
+# these data are read in from shapefile created in qgis
+
+# read in necessary datatsets
+library(rgdal)
+
+setwd("~/BTO/cuckoo_tracking/data/")
+breeding_site<-readOGR(layer="complete_cycle_cuckoo_breeding_buffers",
+               dsn="spatial") # different linux/windows
+
+UK_travel<-read.csv("~/BTO/cuckoo_tracking/data/complete_cycles_UK_timing.csv", h=T)
+hires<-read.csv("~/BTO/cuckoo_tracking/data/movebank_cuckoos_hybridfilter_clean.csv", h=T)
+
+# loop will run per ptt and isolate points within breeding buffers
+# then, per year pull the first and last points
+
+breeding_timing<-NULL
+for(i in gd_cucks)
+{
+  pttSP<-SpatialPointsDataFrame(
+    SpatialPoints(cbind(hires[hires$ptt==i,]$location.long,
+                        hires[hires$ptt==i,]$location.lat), 
+                        CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")),
+                                data=hires[hires$ptt==i,]) 
+  
+  breedingpts<-pttSP[breeding_site[breeding_site$ptt==i,],]@data
+  
+  breedingpts$timestamp <- as.POSIXct(strptime(breedingpts$timestamp, "%Y-%m-%d %H:%M:%S"), "UTC")
+
+  m1<-data.frame(ptt=i,year=aggregate(timestamp~year.x, breedingpts, min)[,1],
+                 breeding_entry=aggregate(timestamp~year.x, breedingpts, min)[,2],
+                 breeding_exit=aggregate(timestamp~year.x, breedingpts, max)[,2])
+  
+  breeding_timing<-rbind(breeding_timing, m1)
+}
+
+
+library(plyr)
+
[…]

### 2018-03-12 west_africa_timing.R
Changed UK entry to Breeding entry and added stats

@@ -173,15 +173,20 @@ travel2[1,7:8]<-travel2[1,3:4]
 #write out
 write.csv(travel2,"~/BTO/cuckoo_tracking/data/complete_cycles_UK_all_timing.csv", quote=F, row.names=F)
 
+#Read in again but this time call UK_travel
+UK_travel<-read.csv("~/BTO/cuckoo_tracking/data/complete_cycles_UK_all_timing.csv", h=T)
+
 # have a look
 
-travel2$UK_entryHACK<-as.Date(paste("2000",
-  substr(travel2$UK_entry,5,11), sep=""))
+UK_travel$UK_entryHACK<-as.Date(paste("2000",
+  substr(UK_travel$UK_entry,5,11), sep=""))
 
-travel2$breeding_entryHACK<-as.Date(paste("2000",
-  substr(travel2$breeding_entry,5,11), sep=""))
+UK_travel$breeding_entryHACK<-as.Date(paste("2000",
+  substr(UK_travel$breeding_entry,5,11), sep=""))
 
-travel2$UK_entry-travel2$breeding_entry
+# difference between arrival in UK and arrival in
+UK_travel$UK_br_diff_days<-
+  UK_travel$breeding_entryHACK-UK_travel$UK_entryHACK
 
 #d_limit<-c(min(UK_travel[UK_travel$deployment_entry==FALSE,]$UK_entryHACK),
 #          max(UK_travel[UK_travel$deployment_entry==FALSE,]$UK_entryHACK))
@@ -189,28 +194,30 @@ travel2$UK_entry-travel2$breeding_entry
 # need to specify limits bit further from the data otherwise it
 # cuts bits off the plot
 
-d_limit<-as.Date(c("2000-04-11", "2000-05-22"))
+summary(UK_travel[UK_travel$deployment_entry
+                  ==FALSE,]$breeding_entryHACK)
+
+d_limit<-as.Date(c("2000-04-12", "2000-05-24"))
 
 p<-ggplot(data=UK_travel[UK_travel$deployment_entry==FALSE,], 
-          aes(x=UK_entryHACK))
+          aes(x=breeding_entryHACK))
 
 p1<-p+geom_histogram(binwidth=1, col='black' )+facet_wrap(~year)+
           scale_x_date(limits=d_limit, 
           date_breaks = "1 week", 
           date_minor_breaks = "1 day",
           date_labels = "%b %d")
-          
 
[…]

### 2018-03-13 west_africa_timing.R
Some stats for arrival date

@@ -307,12 +307,9 @@ aggregate(UK_entryHACK~ptt,
 
 # join stopover data and UK_travel summary table
 
-UK_travel<-read.csv("~/BTO/cuckoo_tracking/data/complete_cycles_UK_timing.csv", h=T)
+UK_travel<-read.csv("~/BTO/cuckoo_tracking/data/complete_cycles_UK_all_timing.csv", h=T)
 
 dat<-read.csv("~/BTO/cuckoo_tracking/data/stopover_bestofday_complete_cycles.csv", h=T)
-
-UK_travel$UK_entryHACK<-as.Date(paste("2000",
-              substr(UK_travel$UK_entry,5,11), sep=""))
 
 
 library(plyr)
@@ -357,10 +354,9 @@ d3$SO_startHACK<-as.Date(paste("2000",
 library(ggplot2)
 
 p<-ggplot(data=d3[d3$deployment_entry!=TRUE & d3$stage=="migration" ,],
-          aes(x=SO_endHACK, y=UK_entryHACK,
+          aes(x=SO_endHACK, y=breeding_entryHACK,
               colour=factor(ptt), shape=region))
-p1<-p+geom_point()+geom_line()+facet_wrap(~year)
-
+p1<-p+geom_point(aes(size=SO_days))+geom_line()+facet_wrap(~year)
 
 
 # so do birds go further west within west africa in relation to date
@@ -395,20 +391,54 @@ by_obj<-by(d3[d3$region=="tropical_Africa",],
 d4<-do.call(rbind, by_obj) # whay!
 # alternate apprach using ddply from plyr? could get to work
 
-p<-ggplot(data=d4,aes(x=SO_median_long, y=SO_startHACK))
-          
-p1<-p+geom_point(aes(colour=factor(year)))+scale_x_date(date_breaks = "1 week", 
-                           date_minor_breaks = "1 day",
-                           date_labels = "%b %d")+
-  scale_y_date(date_breaks = "1 week", 
-               date_minor_breaks = "1 day",
-               date_labels = "%b %d")+geom_smooth(method="lm")
+library(lubridate)
 
-jpeg("~/BTO/cuckoo_tracking/outputs/Africa_UK_spring_timing.jpg",
+d4$breeding_entryDOY<-yday(
+  d4$breeding_entry) 
+
+d4$SO_startDOY<-yday(
+  d4$SO_start) 
+
[…]

### 2018-03-14 gd_cuck_plots.Rmd
Visualising cuckoo spring migration

---
title: "Good Cuck movements"
author: "MM"
date: "13/03/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cuckoos, echo=FALSE, warning=FALSE, message=FALSE}
########################

rm(list=ls())
library(maps)
library(sp)
library(maptools)
library(rgdal)
library(ggplot2)
library(ggmap)
library(ggrepel)

# using best of day data

dat<-read.csv("~/BTO/cuckoo_tracking/data/stopover_bestofday_spring_mig.csv", h=T)

dat2<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_clean_stopovers.csv", h=T)

UK_travel<-read.csv("~/BTO/cuckoo_tracking/data/complete_cycles_UK_all_timing.csv", h=T)

UK_travel<-UK_travel[UK_travel$deployment_entry==FALSE,]

dat2$key<-paste(dat2$ptt, dat2$year.x)

dat3<-dat2[dat2$key %in% unique(paste(UK_travel$ptt, UK_travel$year)),]

dat3<-dat3[dat3$month<6,]

m1 <- get_map(location = c(5, 30), zoom=4)

for(i in unique(dat3$ptt))
{
  cols <- c("2012" = "red", "2013" = "blue", "2014" = "darkgreen", "2015" = "orange", "2016"= "cyan", "2017"= "black")

  g<-ggmap(m1)+
    geom_path(data=dat3[dat3$ptt==i,], aes(x=long, y=lat,
                                             colour=factor(year.x)))+
    geom_jitter(data=dat3[dat3$ptt==i,], aes(x=long, y=lat,
                colour=factor(year.x)), shape=1, width = 0.5, height= 0.5)+ggtitle(i)+

  geom_label_repel(data=dat[dat$ptt==i & dat$deployment_entry==FALSE,], aes(x=SO_median_long, y=SO_median_lat,
                colour=factor(year), label=substr(SO_start, 6,10)))+
  geom_label_repel(data=aggregate(year~breeding_entry, data=dat[dat$ptt==i & dat$deployment_entry==FALSE,], FUN=function(x){unique(x)})
                   , aes(x=-18, y=45,
                colour=factor(year), label=unique(substr(breeding_entry, 6,10)))) + scale_colour_manual(values = cols)

  print(g)

}

```

### 2018-03-14 migration_strategies_and_arrival_time.R

@@ -1,0 +1,43 @@
+### 13/03/18
+
+### Outputs for Chris H 
+
+### 1) Create summary master table for hypotheses testing
+
+d3<-read.csv("~/BTO/cuckoo_tracking/data/stopover_bestofday_spring_mig.csv", h=T)
+
+d3<-na.omit(d3)
+
+library(lubridate)
+
+d3$breeding_entryDOY<-yday(
+  d3$breeding_entry) 
+
+d3$SO_startDOY<-yday(
+  d3$SO_start) 
+
+d3$SO_endDOY<-yday(
+  d3$SO_end) 
+
+
+# sort to regions
+table(d3$country)
+library(maps)
+plot(SO_median_lat~SO_median_long, d3)
+map("world", add=T,  col=3)
+
+d3$region<-"N_Africa_Europe"
+d3[d3$country=="United Kingdom",]$region<-"UK"
+d3[d3$SO_median_lat<20,]$region<-"tropical_Africa"
+
+# OK now just want last stopover in west africa before bird flew
+
+by_obj<-by(d3[d3$region=="tropical_Africa",], 
+           list(d3[d3$region=="tropical_Africa",]$ptt, 
+                d3[d3$region=="tropical_Africa",]$year),
+           FUN=function(x){x[which(x$SO_endDOY==max(x$SO_endDOY)),]})
+
+d4<-do.call(rbind, by_obj) # whay!
+# alternate apprach using ddply from plyr? could get to work
+
+

### 2018-03-14 cuckoo_stopovers.R
Expoted stopovers for non best-of-day data , defined based on >= 1 day stopover duration

@@ -108,9 +108,10 @@ writeOGR(obj=stopovers_spdf, dsn="spatial",
          layer="movebank_cuckoos_bestofday_stopovers", driver="ESRI Shapefile",
          overwrite_layer = T)
   
+
 # To create master spredsheet with stopovers for Chris
 
-dat<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_stopovers.csv", h=T)
+dat<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_clean_stopovers.csv", h=T)
 
 # FORMAT!!
 dat$timestamp <- as.POSIXct(strptime(dat$timestamp, "%Y-%m-%d %H:%M:%S"), "UTC")
@@ -125,16 +126,24 @@ for(i in birds)
   # for each bird give the mgroups which are not migratory ones
   bird_out<-NULL
   
-  mgz<-na.omit(unique(dat[dat$ptt==i & dat$mtype!="M",]$mgroup))
+  mgz<-na.omit(unique(dat[dat$ptt==i & dat$mtype=="S",]$mgroup))
+  # removes 'M' migratory and 'C' deployment point
   
   for (j in mgz)
   {
     ptt_mgroup<-dat[dat$ptt==i & dat$mgroup==j,]
     
     ########## Drops stopovers with only two locations ########
-    if(nrow(ptt_mgroup)<3){next}
+    #if(nrow(ptt_mgroup)<3){next}
     ###########################################################
 
+    ########## Allows a stopover to be defined based on time #
+    # Here we use 1 hour - we're doing this on non duty-cycle data 
+    # as it gives more accurate timings 
+    
+    if(unique(ptt_mgroup$LOS)<1){next}
+    
+    ###########################################################
     
     ptt_mgroup_out<-data.frame(ptt=i, name=unique(ptt_mgroup$name), mgroup=j, 
                                 SO_start=min(ptt_mgroup$timestamp),
@@ -156,7 +165,7 @@ for(i in birds)
   
 }  
 
-write.csv(stopovers_tab, "~/BTO/cuckoo_tracking/data/stopover_table_bestofday.csv", quote=F, row.names=F)
+write.csv(stopovers_tab, "~/BTO/cuckoo_tracking/data/stopover_table_1daymin.csv", quote=F, row.names=F)
 
[…]

### 2018-03-14 interp_visualise.R
Visualisation and hypotheses addressing phase

@@ -1,0 +1,107 @@
+#### 14/03/18
+
+## Interpolate tracking data to gap fill duty cycling 
+## Make animation of spring migration using time-step
+## approach. Could facet by year but could be hectic.
+
+# best of day data
+
+dat<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_stopovers.csv", h=T)
+
+dat$timestamp <- as.POSIXct(strptime(dat$timestamp, "%Y-%m-%d %H:%M:%S"), "UTC")
+dat$TrackTime<-as.double(dat$timestamp)
+names(dat)[names(dat)=="lat"]<-"Latitude"
+names(dat)[names(dat)=="long"]<-"Longitude"
+
+# Interpolation using old seabird code.
+
+####@@@@ RESAMPLE @@@@####
+
+source("~/BTO/cuckoo_tracking/sourced_data/code/Resample.r")
+# note I hased out the plot part of function to speed up
+
+## set up to not interpolate between points > 1 hr apart
+
+results<-NULL
+for(i in unique(dat$ptt))
+{
+  Track<-dat[dat$ptt == i,]
+  
+  #Track<-Track[-which(duplicated(Track$TrackTime)==TRUE),]
+  #print(paste(length(which(duplicated(Track$TrackTime)==TRUE)), "duplicates removed", sep=" "))
+  
+  resample_output<-resample(Track, timeStep=1)  
+  ## timeStep set for 1 hour interval as anything above makes weird
+  
+  #readline("ok")
+  results<-rbind(results,resample_output)
+  print(i)
+}
+
+plot(Latitude~Longitude, results, pch=16, cex=0.4, col=Bird_ID)
+map("worldHires", add=T, col=3)
+
+results$timestamp<-
+paste(as.Date(as.POSIXlt(results$TrackTime, origin="1970-01-01", "GMT")),
+      format((as.POSIXlt(results$TrackTime, origin="1970-01-01", "GMT")), "%H:%M:%S"))
+
+# maybe re-run with ID for stopover or migratory pioint to
+# tell if point is interpolated (majorly) or not
+
+write.csv(results, "~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_1hrINTERP.csv", quote=F, row.names=F)
+
+#### Visualisation
+
[…]

### 2018-03-14 migration_strategies_and_arrival_time.R

@@ -4,6 +4,13 @@
 
 ### 1) Create summary master table for hypotheses testing
 
+# First we need to use the non-best of day dataset and re-define 
+# what we are calling a stopovr. previously we used => 3 days, as per
+# the nature paper so need something smaller
+
+
+#### OLD
+
 d3<-read.csv("~/BTO/cuckoo_tracking/data/stopover_bestofday_spring_mig.csv", h=T)
 
 d3<-na.omit(d3)

### 2018-04-17 migration_strategies_and_arrival_time.R
Write out summary table with resid and other useful columns

@@ -549,10 +549,14 @@ d4_naeu<-d4 %>% filter(region=='Europe' | region=='North Africa')
 
 g1<-d4_naeu %>% group_by(ptt, year)
 
-lc1<-g1 %>% summarise(fc=first(country), minlat=min(SO_median_lat))
+lc1<-g1 %>% dplyr::summarise(first_SO_afterWA=first(country), first_SO_afterWAlat=min(SO_median_lat))
 lc1$year<-as.factor(lc1$year)
 
-dat2<-left_join(dat, lc1, by=c('ptt', 'year'))
+dat2<-dplyr::left_join(dat, lc1, by=c('ptt', 'year'))
+
+# write out data
+write.csv(dat2, "data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv", quote=F, row.names=F)
+
 
 qplot(data=dat2, x=fc, y=m4b_resid, geom='boxplot')
 

### 2018-04-17 cuckoo_stopovers.R
Fixed error that meant script was not actually correcting stopover end date to account for LOS.recalc

@@ -141,14 +141,21 @@ for(i in birds)
     # Here we use 1 hour - we're doing this on non duty-cycle data 
     # as it gives more accurate timings 
     
-    if(unique(ptt_mgroup$LOS)<1){next}
+    if(unique(ptt_mgroup$LOS.recalc)<1){next}
     
     ###########################################################
     
+    # new stopover end datetime from LOS.recalc
+    
+    endTT<-as.double(min(ptt_mgroup$timestamp))+(unique(ptt_mgroup$LOS.recalc)*(24*60*60))
+    endtimestamp<-paste(as.Date(as.POSIXlt(endTT, origin="1970-01-01", "UTC")), 
+                        format((as.POSIXlt(endTT, origin="1970-01-01", "UTC")), "%H:%M:%S"))
+    ###########################################
+    
     ptt_mgroup_out<-data.frame(ptt=i, name=unique(ptt_mgroup$name), mgroup=j, 
                                 SO_start=min(ptt_mgroup$timestamp),
-                                SO_end=max(ptt_mgroup$timestamp),
-                                SO_days=unique(ptt_mgroup$LOS),
+                                SO_end=endtimestamp,
+                                SO_days=unique(ptt_mgroup$LOS.recalc),
                                 SO_median_long=median(ptt_mgroup$long),
                                 SO_median_lat=median(ptt_mgroup$lat),
                                 SO_month=round(median(ptt_mgroup$month)),

### 2018-04-17 stopover_rainfall_vis.Rmd
Creation of knitr file to visualise each stopover and rain pattern. will eventually have rainfall_compile without plot code

---
title: "West Africa cuckoo stopover rainfall patterns"
author: "Mark Miller"
date: "17 April 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pressure, echo=FALSE, message=F, warning=F}

#### work around for different laptop/desktop directories ####
#### !!!! RUN THIS !!!! ####
 if(Sys.info()['nodename']=="D9L5812"){
   setwd("C:/cuckoo_tracking")}else{
    setwd("N:/cuckoo_tracking")}

int_out<-read.csv('data/spring_rainfall_by_stopover.csv', h=T)

dat<-read.csv("data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv", h=T)

### visualisastion

library(ggplot2)

# order country factors by longitude

int_out$country<-factor(int_out$country, levels= c("Angola", "Democratic Republic of the Congo",
                                                   "Congo", "Gabon", "Central African Republic", 
                                                   "Cameroon", "Nigeria", "Togo", "Burkina Faso",
                                                   "Ghana", "Cote d'Ivoire", "Liberia", "Guinea", 
                                                   "Sierra Leone"))
for( i in 2012:2017)
  {
  int1<-int_out[int_out$year==i,]

  for(j in unique(int1$ptt))
      {
dtemp<-dat[dat$year==i & dat$ptt==j,]

p<-ggplot(data=int_out[int_out$ptt==j & int_out$year==i,], 
          aes(x=variable, y=value))+
geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=100), fill='grey')+
  geom_bar(stat='identity', colour='black')+
  geom_smooth(span=0.5, se=FALSE, linetype='dashed', col='red')+
  facet_wrap(~SO_startDOY+country)+theme_classic()+
  ylab('daily rainfall (mm)')+ xlab('Day of year')+
          ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", dtemp$arrive_breeding, "   Mig speed:", round(dtemp$m4c_resid, 2)))

  print(p)

  }
}

```

### 2018-04-17 rainfall_compile.R

@@ -45,6 +45,11 @@ map('world', add=T)
 
 dat<-read.csv('data/stopover_bestofday_1daymin_recalc_spring_mig.csv', h=T)
 
+#remove non- west african rows
+
+dat<-dat[which( ! dat$country%in% c('Algeria', 'France', 'Italy', 'Morocco' ,'Spain',
+                            'United Kingdom', 'Portugal')),]
+
 # Make are of WA extraction
 
 data("wrld_simpl", package = 'maptools')
@@ -72,6 +77,7 @@ names(ras)
 
 print(raster(ras, layer=1))
 
+
 # get annual rasters
 
 # 124 comes from 1;124 max DOY in west africa (4th May)
@@ -88,6 +94,116 @@ r2017<-subset(ras, 2040:(2040+124))
 # I THINK that it is actually pulling the rainfall DOY before the one in
 # question, so 24hr time lag.. which is kinda what we want
 
+
+# Remove - values from rasters? should do !!! 
+
+library(reshape2)
+library(dplyr)
+
+
+int_out<-NULL
+for( i in 2012:2017) # or should I do per row
+{
+ 
+  indat<-dat[dat$year==i,]
+  
+    # cool the 'get' command grabs the collect raster set from the memory
+    ras_temp<-get(paste0('r', i))
+    
+    # add 50 km buffer as per stopover definition. buffer=50000
+    ext<-extract(ras_temp, indat[,7:8], buffer=50000, fun=median)
+    
+    ext[ext<0]<-NA # remove negative values '-99'
+    
+    dimnames(ext)[[2]]<-1:(length(dimnames(ext)[[2]]))
+    
+    out<-data.frame(ptt=indat$ptt, country=indat$country, year=i, 
+                    SO_startDOY=indat$SO_startDOY, 
+                    SO_endDOY=indat$SO_endDOY, ext)
+    
+    out$ID<-with(out, paste(ptt, year, SO_startDOY, sep="_"))
+    
+    outm<-melt(out, id.vars=c('ID', 'ptt', 'country', 'year',
+                              'SO_startDOY', 'SO_endDOY'))
+    
+    outm$variable<-as.numeric(substr(outm$variable,
+                                     2,nchar(as.character(outm$variable))))
+   
+     outm<-outm[order(outm$ID),]
+     
+   
+    out2<-outm %>% group_by(ID) %>%
+      mutate(cuck_pres=ifelse(variable %in% 
[…]

### 2018-04-18 cuckoo_stopovers.R
Minor edits to cuckoo_stopovers. Re-analyses of migration strat with corrected data. Animation loop added to rainfall_compile

@@ -111,6 +111,10 @@ writeOGR(obj=stopovers_spdf, dsn="spatial",
 
 # To create master spredsheet with stopovers for Chris
 
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+
 dat<-read.csv("data/processed_movebank_cuckoos_hybrid_filter_bestofday_clean_stopovers_recalc.csv", h=T)
 
 # FORMAT!!
@@ -247,6 +251,17 @@ write.csv(dat, "data/stopover_table_bestofday_1daymin_recalc_biomes.csv", quote=
 ## Add columns with migration cohort to 
 ## sort the issue of crossing years on migration manually in excel
 
+## Manual fill of NA countries from GIS lookup
+
+dat[which(is.na(dat$country)),]
+dat[2,]$country<-'Belgium'
+dat[c(426,442),]$country<-'United Kingdom'
+dat[c(426,442),]$country<-'United Kingdom'
+dat[193,]$country<-'Netherlands'
+dat[207,]$country<-'United Kingdom'
+dat[833,]$country<-'United Kingdom'
+dat[568,]$country<-'Spain'
+dat[570,]$country<-'United Kingdom'
 
 # Exporting as KML files 
 

### 2018-04-18 rainfall_compile.R

@@ -144,6 +144,104 @@ for( i in 2012:2017) # or should I do per row
 
 write.csv(int_out, 'data/spring_rainfall_by_stopover.csv', row.names = F, quote=F)
 
+
+
+
+#### animation of rainfall #####
+
+data("wrld_simpl", package = 'maptools')
+
+library('sf')
+
+world <- st_as_sf(wrld_simpl)
+
+dat<-read.csv('data/stopover_bestofday_1daymin_recalc_spring_mig.csv', h=T)
+
+# remove negative values
+
+r2012[r2012<0]<-NA
+r2013[r2013<0]<-NA
+r2014[r2014<0]<-NA
+r2015[r2015<0]<-NA
+r2016[r2016<0]<-NA
+r2017[r2017<0]<-NA
+
+d2012<-dat[dat$year=='2012',]
+d2013<-dat[dat$year=='2013',]
+d2014<-dat[dat$year=='2014',]
+d2015<-dat[dat$year=='2015',]
+d2016<-dat[dat$year=='2016',]
+d2017<-dat[dat$year=='2017',]
+
+
+library(tmap)
+library(animation)
+
+tmaptools::palette_explorer() 
+
+
+tmap_mode('plot')
+
+saveGIF({
+for(i in 1:123)
+{
+ 
+  indat<-which(by(d2012, 1:nrow(d2012), FUN=function(x){i %in% x$SO_startDOY:x$SO_endDOY}))
+   
+  d_day<-d2012[indat,]
+  
+  d_day$name<-factor(d_day$name)
+  
+  if(nrow(d_day)==0){
+    d_day<-SpatialPointsDataFrame(SpatialPoints(cbind(8, 55), 
+            proj4string=CRS(projection(ras))), data=d_day[1,])
+            }else{
+            d_day<-SpatialPointsDataFrame(SpatialPoints(d_day[,7:8], 
+            proj4string=CRS(projection(ras))), data=d_day)
+            }
+  
+    ras1<-subset(r2012,i)
+    
+    if(i==1){cumra<-ras1}else{cumra<-cumra+ras1}
+    cumra[cumra<1]<-NA
+    
+    ras1[ras1>10]<-10
+    
+    p1<-rasterToPolygons(ras1, dissolve=T, fun=function(x){x==10})
+    if(is.null(p1)){p1<-rasterToPolygons(crop(ras1, extent(c(20,21, 55, 56))), dissolve=T)}
+    
+    
+    print(
[…]

### 2018-04-20 rainfall_compile.R
Made GIF loop for all years in tmap. Had to runon linux laptop cos ImageMagik not installed on bto laptop

@@ -177,6 +177,7 @@ d2017<-dat[dat$year=='2017',]
 library(tmap)
 library(animation)
 
+
 tmaptools::palette_explorer() 
 
 
@@ -190,38 +191,531 @@ for(i in 1:123)
    
   d_day<-d2012[indat,]
   
-  d_day$name<-factor(d_day$name)
-  
   if(nrow(d_day)==0){
     d_day<-SpatialPointsDataFrame(SpatialPoints(cbind(8, 55), 
-            proj4string=CRS(projection(ras))), data=d_day[1,])
+            proj4string=CRS(projection(ras))), data=d2012[1,])
             }else{
             d_day<-SpatialPointsDataFrame(SpatialPoints(d_day[,7:8], 
             proj4string=CRS(projection(ras))), data=d_day)
             }
   
+  if(i <= min(d2012$SO_startDOY)){
+    temp<-data.frame(time=c(1,2,3,4), long=c(8,9,10,11), lat=c(55,56,57,59))
+    sp.lines <- points_to_line(data = temp, 
+                               long = "long", 
+                               lat = "lat", 
+                               sort_field = "time")
+    }else{
+    dtemp<-d2012[d2012$SO_startDOY<=i,]
+    dtemp$SO_startDOY=as.integer(dtemp$SO_startDOY-1)
+    dtemp$SO_median_long=dtemp$SO_median_long-0.001
+    
+    if(length(unique(d2012[d2012$SO_startDOY<=i,]$ptt))>1)
+    {
+    sp.lines <- points_to_line(data = rbind(dtemp,
+                                      d2012[d2012$SO_startDOY<=i,]), 
+                               long = "SO_median_long", 
+                               lat = "SO_median_lat", 
+                               id_field='ptt',
+                               sort_field = "SO_startDOY")
+    }else{
+      sp.lines <- points_to_line(data = rbind(dtemp,
+                                              d2012[d2012$SO_startDOY<=i,]), 
+                                 long = "SO_median_long", 
+                                 lat = "SO_median_lat",
+                                 sort_field = "SO_startDOY")
+    }
+    }
+    
+  proj4string(sp.lines) <- CRS(projection(ras))
+    
+  
     ras1<-subset(r2012,i)
     
     if(i==1){cumra<-ras1}else{cumra<-cumra+ras1}
-    cumra[cumra<1]<-NA
     
     ras1[ras1>10]<-10
     
[…]

### 2018-04-23 stopover_rainfall_vis.Rmd
minor change to image size export

@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-```{r pressure, echo=FALSE, message=F, warning=F}
+```{r cuckoos, echo=FALSE, message=F, warning=F, fig.height=12, fig.width=12}
 
@@
                                                    "Sierra Leone"))
+outlist<-list()
 for( i in 2012:2017)
@@
 
-p<-ggplot(data=int_out[int_out$ptt==j & int_out$year==i,], 
+szr<-length(unique(paste(int_out[int_out$ptt==j & int_out$year==i,]$country,int_out[int_out$ptt==j & int_out$year==i,]$SO_startDOY )))
+
+temp=ggplot(data=int_out[int_out$ptt==j & int_out$year==i,], 
           aes(x=variable, y=value))+
-geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=100), fill='grey')+
+geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=50), fill='grey')+
   geom_bar(stat='identity', colour='black')+
@@
 
-  print(p)
+outlist[[paste(szr, i, j)]]<-temp
+print(temp)
 

### 2018-04-25 stopover_rainfall_vis.Rmd
Changes from laptop work 3rd week of Apr

@@
 library(ggplot2)
+library(dplyr)
 
@@
                                                    "Sierra Leone"))
-outlist<-list()
+int_out[is.na(int_out$value),]$value<-0
+int_out$ID2<-paste(int_out$SO_startDOY, int_out$country)
+
 for( i in 2012:2017)
@@
 
-szr<-length(unique(paste(int_out[int_out$ptt==j & int_out$year==i,]$country,int_out[int_out$ptt==j & int_out$year==i,]$SO_startDOY )))
+t1<-int_out[int_out$ptt==j & int_out$year==i,]
 
-temp=ggplot(data=int_out[int_out$ptt==j & int_out$year==i,], 
-          aes(x=variable, y=value))+
-geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=50), fill='grey')+
+t2<-t1 %>% group_by(ID2) %>% dplyr::mutate(cumrf=cumsum(value))
+
+temp=ggplot(data=t2,aes(x=variable, y=value))+
+geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=500), fill='grey')+
   geom_bar(stat='identity', colour='black')+
-  geom_smooth(span=0.5, se=FALSE, linetype='dashed', col='red')+
-  facet_wrap(~SO_startDOY+country)+theme_classic()+
-  ylab('daily rainfall (mm)')+ xlab('Day of year')+
+    geom_line(aes(y=cumrf), col='green', size=1)+
+  facet_wrap(~country+SO_startDOY, scales='free_y')+theme_classic()+
+  ylab('cumulative rainfall since 01 Jan (mm)')+ xlab('Day of year')+
           ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", dtemp$arrive_breeding, "   Mig speed:", round(dtemp$m4c_resid, 2)))
 
-outlist[[paste(szr, i, j)]]<-temp
 print(temp)
+
+#temp=ggplot(data=t2,aes(x=variable, y=value))+
+#geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=50), #fill='grey')+
+#  geom_bar(stat='identity', colour='black')+
+#  geom_smooth(span=0.5, se=FALSE, linetype='dashed', col='red')+
+#  facet_wrap(~country+SO_startDOY, scales='free_y')+theme_classic()+
+#  ylab('daily rainfall (mm)')+ xlab('Day of year')+
+#          ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", #dtemp$arrive_breeding, "   Mig speed:", round(dtemp$m4c_resid, 2)))
 

### 2018-04-25 cuckoo_stopovers.R

@@ -185,7 +185,7 @@ library(ggplot2)
 library(ggmap)
 
 # using best of day data
-dat<-read.csv("data/stopover_table_biomes_1daymin_recalc.csv", h=T)
+dat<-read.csv("data/stopover_table_bestofday_1daymin_recalc.csv", h=T)
 
 setwd("sourced_data/")
 countries<-readOGR(layer="TM_WORLD_BORDERS-0.3",

### 2018-04-25 rainfall_analyses.R

@@ -1,0 +1,78 @@
+#24/04/2018
+# Rainfall analyses 
+# Firstly pull in stopover data and analyse
+
+#### work around for different laptop/desktop directories ####
+#### !!!! RUN THIS !!!! ####
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+#### !!!!!!!!!!!!!!!!!! ####
+
+library(dplyr)
+library(ggplot2)
+
+int_out<-read.csv('data/spring_rainfall_by_stopover.csv', h=T)
+int_out[is.na(int_out$value),]$value<-0
+
+int_out$country<-factor(int_out$country, levels= c("Angola", "Democratic Republic of the Congo",
+                                                   "Congo", "Gabon", "Central African Republic", 
+                                                   "Cameroon", "Nigeria", "Togo", "Burkina Faso",
+                                                   "Ghana", "Cote d'Ivoire", "Liberia", "Guinea", 
+                                                   "Sierra Leone"))
+
+t2<-int_out %>% group_by(ID) %>% dplyr::mutate(cumrf=cumsum(value))
+
+t2$cuckbin<-0
+t2[t2$cuck_pres=='Y',]$cuckbin<-1
+
+library(lme4)
+
+m1<-glmer(cuckbin~poly(cumrf, 2)+country+(1|ID), family=binomial, data=t2)
+
+pred.df=expand.grid(country=unique(t2$country), 
+                    cumrf=0:1000)
+
+pred.df$p1<-predict(m1, newdata=pred.df, re.form=~0)
+
+
+qplot(data=t2, x=cumrf, y=cuckbin)+
+  geom_line(data=pred.df, (aes(x=cumrf, y=plogis(p1), colour=country)))
+
+# looks like the curve needs more flexability
+
+m2<-glmer(cuckbin~poly(cumrf, 2):country+(1|ID), family=binomial, data=t2)
+pred.df$p2<-predict(m2, newdata=pred.df, re.form=~0)
+
+qplot(data=t2, x=cumrf, y=cuckbin)+
+  geom_line(data=pred.df, (aes(x=cumrf, y=plogis(p2), colour=country)))
+
+anova(m1, m2)
+
+out=NULL
+for( i in unique(t2$country))
+{
+  tempdat<-t2[t2$country==i,]
[…]

### 2018-04-25 migration_strategies_and_arrival_time.R
couple of little analyses additions

@@ -620,3 +620,19 @@ Anova(SOc, test.statistic = 'F')
 SOne<-lmer(sumSOeurope~sumSOnorthAF+factor(year)+
             (1|ptt), data=dat)
 Anova(SOne, test.statistic = 'F')
+
+
+
+# extra analyses  25/04
+
+qplot(data=dat, y=arrive_breeding, x=breeding_loc)+geom_point(data=dat, aes(y=arrive_uk, x=breeding_loc), shape=2)+geom_point(data=dat, aes(y=DEPwestAF,x=breeding_loc), shape=3)+facet_wrap(~year)+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+geom_text(data=dat, aes(y=DEPwestAF+2,x=breeding_loc, label=arrive_breeding-DEPwestAF), nudge_x=0.5)+geom_text(data=dat, aes(y=arrive_breeding+2,x=breeding_loc, label=ptt), size=3)
+
+library(dplyr)
+
+
+d5<-d4 %>% group_by(ptt, year, region) %>% summarise(last_c=last(country))
+
+qplot(data=dat2, y=DEPwestAF, x=last_c, col=factor(year))
+
+
+

### 2018-04-25 migration_strategies_and_arrival_time.R
created dead birds table to rbind with live ones and then analyse

@@ -33,7 +33,7 @@ dat2$SO_year=as.numeric(substr(dat2$SO_start, 1,4))
 
 dat2<-rename(dat2, c("SO_year" = "year"))
 
-d2<-join_all(list(dat2, UK_travel), by=c("ptt", "year"))
+d2<-plyr::join_all(list(dat2, UK_travel), by=c("ptt", "year"))
 
 # this line subsets birds and years that I want to exclude from the 
 # the analyses i.e they do not complete a full cycle
@@ -100,9 +100,84 @@ d4$unreliable_exit<-NULL
 write.csv(d4, "data/stopover_bestofday_1daymin_recalc_spring_mig.csv", quote=F, row.names=F)
 
 
+#### to make d2 for birds that died ####
+
+dat2<-read.csv("data/stopover_table_bestofday_1daymin_recalc_biomes.csv", h=T)
+
+dat2$SO_start <- as.POSIXct(strptime(dat2$SO_start, "%Y-%m-%d %H:%M:%S"), "UTC")
+dat2$SO_end <- as.POSIXct(strptime(dat2$SO_end, "%Y-%m-%d %H:%M:%S"), "UTC")
+
+library(plyr)
+
+#re-define year so that it is the year that the stopover starts in
+dat2$SO_year=as.numeric(substr(dat2$SO_start, 1,4))
+
+dat2<-rename(dat2, c("SO_year" = "year"))
+
+dat2$UK_entry<-NA
+dat2$UK_exit<-NA
+dat2$breeding_entry<-NA
+dat2$breeding_exit<-NA
+dat2$UK_br_diff_days<-NA
+
+library(lubridate)
+
+dat2$SO_startDOY<-yday(
+  dat2$SO_start) 
+
+dat2$SO_endDOY<-yday(
+  dat2$SO_end) 
+
+
+
+dead_cucks=data.frame(ptt=c(62518,62520,62602,11597,11599,128296,128303,161323,134958,128300),
+                      year_died=c(2012,2012,2012,2013,2014,2014,2014,2017,2015,2015), 
+                      done_dead='yarp')
+
+d2<-plyr::join_all(list(dat2, dead_cucks), by=c("ptt"))
+
+d2<-d2[-which(is.na(d2$done_dead)),]
+
+# all birds except livingstone were tagged the summer before they died
+d2<-d2[-which(d2$ptt==128300 & d2$year==2013),]
+
[…]

### 2018-04-26 migration_strategies_and_arrival_time.R
Added code to analyse dead cuckoos in the mix

@@ -226,16 +226,16 @@ nrow(d4[d4$SO_days<3,]) # old appraoch lost ~ 25% of stopovers
 
 # OK now just want last stopover in west africa before bird flew
 
-out_tab<-ddply(d4, .(ptt, year, region), summarize,
+out_tab<-ddply(d4, .(ptt, year, dead, region), summarize,
                depart=max(SO_end), no_SO=length(SO_end),
                sum_SO=sum(SO_days), minlongWA=min(SO_median_long))
 
-out2<-ddply(d4, .(ptt, year), summarize, depart_winterSO=unique(winterSO_end),
+out2<-ddply(d4, .(ptt, year, dead), summarize, depart_winterSO=unique(winterSO_end),
             arrive_uk=unique(UK_entry), arrive_breeding=unique(breeding_entry))
             
 
 out2.5<-join_all(list(out2, out_tab[out_tab$region=="Central Africa",]),
-               by=c("ptt", "year"))
+               by=c("ptt", "year", 'dead'))
 out2.5$region<-NULL
 out2.5$minlongWA<-NULL
 out2.5<-rename(out2.5, c("depart"="DEPcentralAF",
@@ -243,7 +243,7 @@ out2.5<-rename(out2.5, c("depart"="DEPcentralAF",
                      "sum_SO"="sumSOcentralAF"))
 
 out3<-join_all(list(out2.5, out_tab[out_tab$region=="West Africa",]),
-                 by=c("ptt", "year"))
+                 by=c("ptt", "year", 'dead'))
 
 out3$region<-NULL
 out3<-rename(out3, c("depart"="DEPwestAF",
@@ -252,7 +252,7 @@ out3<-rename(out3, c("depart"="DEPwestAF",
                      'minlongWA'='minlongwestAF'))
 
 out3.5<-join_all(list(out3, out_tab[out_tab$region=="North Africa",]),
-               by=c("ptt", "year"))
+               by=c("ptt", "year", 'dead'))
 
 out3.5$region<-NULL
 out3.5$minlongWA<-NULL
@@ -261,7 +261,7 @@ out3.5<-rename(out3.5, c("depart"="DEPnorthAF",
                      "sum_SO"="sumSOnorthAF"))
 
 out4<-join_all(list(out3.5, out_tab[out_tab$region=="Europe",]),
-                 by=c("ptt", "year"))
+                 by=c("ptt", "year", 'dead'))
 
 out4$region<-NULL
 out4$minlongWA<-NULL
[…]

### 2018-05-29 ndvi_compile.R
First attempt using 'MODIS' package to download ndvi data

@@ -1,0 +1,17 @@
+## 25/05/18
+## Downloading, processing and extracting MODIS NDVI data
+
+library(MODIS)
+
+#check for GDAL
+MODISoptions()
+
+# add GDAL path to list
+Sys.setenv(PATH="C:\\Users\\mark.gr.miller\\Documents\\R\\R-3.4.4\\bin\\x64;C:\\ProgramData\\Oracle\\Java\\javapath;C:\\WINDOWS\\system32;C:\\WINDOWS;C:\\WINDOWS\\System32\\Wbem;C:\\WINDOWS\\System32\\WindowsPowerShell\\v1.0\\;C:\\Users\\mark.gr.miller\\AppData\\Local\\Microsoft\\WindowsApps;C:\\Users\\mark.gr.miller\\AppData\\Local\\Programs\\Git\\cmd;C://Program Files//QGIS 2.18//bin")
+
+# recheck
+MODISoptions()
+
+ng<-getTile(x='Nigeria')
+
+gd<-runGdal(product='MYD13A2', extent='Nigeria',  begin="2018001", end="2018005")

### 2018-05-29 ndvi_compile.R
Change to 'MODIStsp' package (very little code due to GUI use)

@@ -1,17 +1,16 @@
 ## 25/05/18
 ## Downloading, processing and extracting MODIS NDVI data
 
-library(MODIS)
+# Following tutorial:
+#https://cran.r-project.org/web/packages/MODIStsp/vignettes/MODIStsp.pdf
 
-#check for GDAL
-MODISoptions()
+#install.packages("gWidgetsRGtk2")
+library(gWidgetsRGtk2)
 
-# add GDAL path to list
-Sys.setenv(PATH="C:\\Users\\mark.gr.miller\\Documents\\R\\R-3.4.4\\bin\\x64;C:\\ProgramData\\Oracle\\Java\\javapath;C:\\WINDOWS\\system32;C:\\WINDOWS;C:\\WINDOWS\\System32\\Wbem;C:\\WINDOWS\\System32\\WindowsPowerShell\\v1.0\\;C:\\Users\\mark.gr.miller\\AppData\\Local\\Microsoft\\WindowsApps;C:\\Users\\mark.gr.miller\\AppData\\Local\\Programs\\Git\\cmd;C://Program Files//QGIS 2.18//bin")
+library(devtools)
+install_github("lbusett/MODIStsp")
 
-# recheck
-MODISoptions()
+library(MODIStsp)
+MODIStsp()
+# Thats all you need to use this tool! nice!
 
-ng<-getTile(x='Nigeria')
-
-gd<-runGdal(product='MYD13A2', extent='Nigeria',  begin="2018001", end="2018005")

### 2018-05-30 rainfall_ndvi_extract.R
Extract both rainfall and ndvi for years 2012-2018 as per two methods: 1) per stopover, showing conditions at stopover during whole spring/summer period; and 2) spatially, showing conditions of stopover in use against all possible stopover conditions for the same period

@@ -1,0 +1,153 @@
+## remote sensing data Rainfall and NDVI extract
+## 30/05/2018
+
+# Get cuckoo data
+
+dat<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig.csv', h=T)
+dat$dead<-0
+dat1<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig_dead.csv', h=T)
+dat1$dead<-1
+dat1$biome1<-NULL
+dat1$biome2<-NULL
+
+dat<-rbind(dat,dat1)
+#remove non- west african rows
+
+dat<-dat[which( ! dat$country%in% c('Algeria', 'France', 'Italy', 'Morocco' ,'Spain',
+                                    'United Kingdom', 'Portugal')),]
+
+
+# read in monthly data in netcdf format ok
+
+library(raster) # needs ncdf4 installed
+
+# stacks up all the layers
+#ras<-stack("sourced_data/PERSIANN_rainfall/monthly/PERSIANN_2018-03-26035525am.nc")
+ras<-stack("sourced_data/PERSIANN_rainfall/daily_inc2018/PERSIANN_2018-05-25054150am.nc")
+
+
+# warning in projection 
+
+print(raster(ras, layer=1))
+
+# basically it is specifcying WGS but CRS gets it a bit wrong
+
+ras@crs
+#fix 
+ras@crs<-CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
+
+
+plot(raster(ras, layer=5))
+
+library(maps)
+map('world', add=T)
+
+
+##### check data
+
+names(ras)
+
+# name is 'units: days since 2011-06-01'
+
+print(raster(ras, layer=1))
+
+
+# get annual rasters
+
+# 124 comes from 1;124 max DOY in west africa (4th May)
+
+r2012<-subset(ras, 213:(213+124))
+r2013<-subset(ras, 579:(579+124))
+r2014<-subset(ras, 944:(944+123))
+r2015<-subset(ras, 1309:(1309+124))
+r2016<-subset(ras, 1674:(1674+124))
+r2017<-subset(ras, 2040:(2040+124))
+r2018<-subset(ras, 2405:(2405+124))
+
+# With each annual raster, the position [x] of the layer is the DOY
+
+# I THINK that it is actually pulling the rainfall DOY before the one in
+# question, so 24hr time lag.. which is kinda what we want
+
+
+# Remove - values from rasters? should do !!! 
+
[…]

### 2018-05-30 ndvi_compile.R
Added gdal loop to reproject to WGS

@@ -14,3 +14,54 @@ library(MODIStsp)
 MODIStsp()
 # Thats all you need to use this tool! nice!
 
+# OK so I ran the MODIStsp tool for 1 km NDVI from Terra and 
+# Aqua see here for run parameters:
+#C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/MODIStsp_parameters.png
+
+## OK now explore and compile the data
+
+library(raster)
+
+tifz<-list.files("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI")
+
+#order them by date rather than satellite (Terra and Aqua)
+tifz<-tifz[order(substr(tifz, 14,21))]
+
+for(i in tifz)
+{
+if(which(i==tifz)==1){
+  rastack<-raster(paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI/",i))
+  }else{
+  rastack<-stack(rastack, paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI/",i))  
+  }
+print(i)  
+}
+
+
+plot(raster::subset(rastack, 1))
+
+# data needs scaling factor (0.0001) applied to adjust to wothin
+# NDVI bounds (-1 to 1)
+# Will do at later stage
+
+# # OK so thats all great but I need them in WGS format for visualising
+# raster's 'repojectRaster' seems to take ages so try gith GDAL
+
+library(gdalUtils)
+
+tifz<-list.files("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI")
+
+#order them by date rather than satellite (Terra and Aqua)
+tifz<-tifz[order(substr(tifz, 14,21))]
+
+for(i in tifz)
+{
+  gdalwarp(srcfile=paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI/",i),
+           dstfile=paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI_WGS/",i),
+           s_srs=projection(rastack),
+           t_srs='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs',
+           output_Raster=TRUE,
[…]

### 2018-05-30 rainfall_ndvi_extract.R
Per stopover extract now runs with rainfall and ndvi data. Note extract still using a 50km radius and median value, for ndvi could also take std dev?

@@ -1,6 +1,13 @@
 ## remote sensing data Rainfall and NDVI extract
 ## 30/05/2018
 
+#### work around for different laptop/desktop directories ####
+#### !!!! RUN THIS !!!! ####
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+#### !!!!!!!!!!!!!!!!!! ####
+
 # Get cuckoo data
 
 dat<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig.csv', h=T)
@@ -17,6 +24,8 @@ dat<-dat[which( ! dat$country%in% c('Algeria', 'France', 'Italy', 'Morocco' ,'Sp
                                     'United Kingdom', 'Portugal')),]
 
 
+##### %%%%%%%%% RAINFALL DATA %%%%%%%%%%%%###
+
 # read in monthly data in netcdf format ok
 
 library(raster) # needs ncdf4 installed
@@ -24,8 +33,6 @@ library(raster) # needs ncdf4 installed
 # stacks up all the layers
 #ras<-stack("sourced_data/PERSIANN_rainfall/monthly/PERSIANN_2018-03-26035525am.nc")
 ras<-stack("sourced_data/PERSIANN_rainfall/daily_inc2018/PERSIANN_2018-05-25054150am.nc")
-
-
 # warning in projection 
 
 print(raster(ras, layer=1))
@@ -42,18 +49,14 @@ plot(raster(ras, layer=5))
 library(maps)
 map('world', add=T)
 
-
 ##### check data
 
 names(ras)
-
 # name is 'units: days since 2011-06-01'
 
 print(raster(ras, layer=1))
 
-
 # get annual rasters
-
 # 124 comes from 1;124 max DOY in west africa (4th May)
 
 r2012<-subset(ras, 213:(213+124))
@@ -69,18 +72,37 @@ r2018<-subset(ras, 2405:(2405+124))
 # I THINK that it is actually pulling the rainfall DOY before the one in
 # question, so 24hr time lag.. which is kinda what we want
 
+##### %%%%%%%%% NDVI DATA %%%%%%%%%%%%###
 
-# Remove - values from rasters? should do !!! 
[…]

### 2018-05-30 stopover_rainfall_vis.Rmd
Now working with rainfall and ndvi

@@
 
-int_out<-read.csv('data/spring_rainfall_by_stopover.csv', h=T)
+int_out<-read.csv('data/spring_rainfall_NDVI_by_stopover_2018_dead.csv', h=T)
 
-dat<-read.csv("data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv", h=T)
+dat<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv", h=T)
 
@@
 geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=500), fill='grey')+
+    geom_bar(data=t2, aes(x=variable, y=value2*500, group=1), stat='identity', fill='red', linetype=0, alpha=1/3)+
   geom_bar(stat='identity', colour='black')+
@@
   ylab('cumulative rainfall since 01 Jan (mm)')+ xlab('Day of year')+
-          ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", dtemp$arrive_breeding, "   Mig speed:", round(dtemp$m4c_resid, 2)))
+          ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", dtemp$arrive_breeding))
+
+#"   Mig speed:", round(dtemp$m4c_resid, 2)
 

### 2018-05-30 rainfall_ndvi_extract.R

@@ -90,6 +90,9 @@ for(i in tifz)
 }
 
 projection(rastack)
+#MODIS naming convention means that the date on the product name is the start
+# of the 16 day period iver which data is averaged see here for more:
+#https://gis.stackexchange.com/questions/218564/how-do-modis-products-naming-conventions-work
 
 
 library(reshape2)
@@ -128,15 +131,14 @@ for( i in 2012:2018)
   
   # loop to rep ndvi values over each of the 8 days they represent
   extrep<-NULL
-  for(i in 1:(ncol(ext2)-1))
+  for(j in 1:15) # limits to day 121
   {
-   extrep<-cbind(extrep, matrix(rep(ext2[,i], 8), ncol=8)) 
+   extrep<-cbind(extrep, matrix(rep(ext2[,j], 8), ncol=8)) 
   }
   extrep<-cbind(extrep, matrix(rep(ext2[,16], 5), ncol=5))
   
 
-  
-  
+  out<-data.frame(out, extrep) # add the ndvi data to rainfall df
   
   out$ID<-with(out, paste(ptt, year, SO_startDOY, sep="_"))
   
@@ -164,7 +166,7 @@ for( i in 2012:2018)
 
 # write out
 
-write.csv(int_out, 'data/spring_rainfall_by_stopover_2018.csv', row.names = F, quote=F)
+write.csv(int_out, 'data/spring_rainfall_NDVI_by_stopover_2018_dead.csv', row.names = F, quote=F)
 
 
 ### Spatial extraction and analyses.
@@ -172,7 +174,7 @@ write.csv(int_out, 'data/spring_rainfall_by_stopover_2018.csv', row.names = F, q
 ### against rainfall of all other stopovers extract for n's date
 
 int_out2<-NULL
-for( i in 1:nrow(dat)) # or should I do per row
+for( i in 1:nrow(dat)) 
 {
   
   indat<-dat[i,]

### 2018-06-26 static_extract.R
Extraction of 20m landcover data and distance to steam layer

@@ -1,0 +1,74 @@
+## remote sensing data Rainfall and NDVI extract
+## 30/05/2018
+
+#### work around for different laptop/desktop directories ####
+#### !!!! RUN THIS !!!! ####
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+#### !!!!!!!!!!!!!!!!!! ####
+
+
+# Use detailed coord data now for extract
+
+dat<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords.csv', h=T)
+
+
+dat<-dat[which( ! dat$country%in% c('Algeria', 'France', 'Italy', 'Morocco' ,'Spain',
+                                    'United Kingdom', 'Portugal')),]
+
+
+# data accessed via the portal: http://maps.elie.ucl.ac.be/CCI/viewer/index.php
+
+
+
+library(raster) 
+
+ras<-raster("sourced_data/ESA_CCI_landcover/ESACCI-LC-L4-LC10-Map-20m-P1Y-2016-v1.0/ESACCI-LC-L4-LC10-Map-20m-P1Y-2016-v1.0.tif")
+
+#landcover<-extract(ras, dat[, 23:24], buffer=150, df=T) # add 1500 m buffer as landcover is c. 250 m resolution
+## only use 150m cos 1500 grabs too many!
+
+#landcover2<-extract(ras, dat[, 23:24], buffer=1500, fun=function(x){table(x)}) # add 1500 m buffer as landcover is c. 250 m resolution
+## only use 150m cos 1500 grabs too many!
+
+landcover3<-extract(ras, dat[, 23:24], buffer=1500) # add 1500 m buffer as landcover is c. 20 m!! resolution
+## on
+lc3<-lapply(landcover3, FUN=function(x)
+{
+  data.frame(matrix(table(x), ncol=length(table(x)),
+                    dimnames=list(NULL,unlist(dimnames(table(x))))))
+})
+
+library(dplyr)
+
+lc3a<-bind_rows(lc3) # binds up rows if columns don't match gives an NA - sweet!
+
+dat<-cbind(dat, lc3a)
+
+names(dat)[names(dat)=='X1']<-"Trees"
+names(dat)[names(dat)=='X2']<-"Shrubs"
+names(dat)[names(dat)=='X3']<-"Grasslands"
[…]

### 2018-06-26 stopover_rainfall_ndvi_detailcoords_vis.Rmd
Updates to add landcover info

---
title: "West Africa cuckoo stopover rainfall patterns"
author: "Mark Miller"
date: "17 April 2018"
output: html_document
---

```{r cuckoos, echo=FALSE, message=F, warning=F, fig.height=12, fig.width=12}

#### work around for different laptop/desktop directories ####
#### !!!! RUN THIS !!!! ####
 if(Sys.info()['nodename']=="D9L5812"){
   setwd("C:/cuckoo_tracking")}else{
    setwd("N:/cuckoo_tracking")}

int_out<-read.csv('data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', h=T)

dat<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv", h=T)

lc<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext.csv", h=T)

### visualisastion

library(ggplot2)
library(dplyr)

# order country factors by longitude

int_out$country<-factor(int_out$country, levels= c("Angola", "Democratic Republic of the Congo",
                                                   "Congo", "Gabon", "Central African Republic", 
                                                   "Cameroon", "Nigeria", "Togo", "Burkina Faso",
                                                   "Ghana", "Cote d'Ivoire", "Liberia", "Guinea", 
                                                   "Sierra Leone"))
int_out[is.na(int_out$value),]$value<-0

for( i in 2012:2018)
  {
  int1<-int_out[int_out$year==i,]

  for(j in unique(int1$ptt))
      {
dtemp<-dat[dat$year==i & dat$ptt==j,]

t1<-int_out[int_out$ptt==j & int_out$year==i,]

#landcover metrics
lc1<-lc[lc$ptt==j & lc$year==i,]

lc1<-lc1 %>% group_by(SO_startDOY, country) %>% dplyr::summarise(Tree=mean(Trees, na.rm=T), Grass=mean(Grasslands, na.rm=T), Crops=mean(Croplands, na.rm=T), Shrub=mean(Shrubs, na.rm=T),Bare=mean(Bare.areas, na.rm=T), Urban=mean(Built.up.areas, na.rm=T),
  Aquatic=mean(Aquatic.veg.reg.flooded, na.rm=T), Water=mean(Open.water, na.rm=T))

                                                             lc1[,3:10]<-round(lc1[,3:10]/rowSums(lc1[,3:10], na.rm=T) *100)

lc1$label<-""   
for(k in 1:nrow(lc1))
{
  lctemp<-lc1[k,3:10]

  lcouty<-NULL
  for(l in which(!is.na(lctemp)))
{
  tempy<-paste(names(lctemp)[l], ": ", lctemp[l], "% \n", sep="")
    lcouty<-paste(lcouty, tempy)
}

lc1[k,]$label<-lcouty

}

t1.1<-t1 %>% group_by(ptt, year, SO_startDOY, variable) %>% dplyr:: summarise_if(is.numeric, mean, na.rm = TRUE)

t1.2<-t1 %>% group_by(ptt, year, SO_startDOY, variable) %>% dplyr:: summarise_if(is.factor, first)

t1.3<-cbind(t1.1, t1.2)

t2<-t1.3 %>% group_by(SO_startDOY, country) %>% dplyr::mutate(cumrf=cumsum(value))

temp=ggplot(data=t2,aes(x=variable, y=value))+
geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=500), fill='grey')+
    geom_line(data=t2, aes(x=variable, y=value2*500), colour='red')+
  geom_line(data=t2, aes(x=variable, y=aquaNDVI*500), colour='cyan')+
  geom_line(data=t2, aes(x=variable, y=terrNDVI*500), colour='orange')+
  geom_bar(stat='identity', colour='black')+
    geom_line(aes(y=cumrf), col='green', size=1)+
  facet_wrap(~country+SO_startDOY, scales='free_y')+theme_classic()+
  ylab('cumulative rainfall since 01 Jan (mm)')+ xlab('Day of year')+
          ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", dtemp$arrive_breeding))+geom_label(data=lc1, aes(x=15, y=450, vjust=1,label=label), fill=NA)

#"   Mig speed:", round(dtemp$m4c_resid, 2)

print(temp)

#temp=ggplot(data=t2,aes(x=variable, y=value))+
#geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=50), #fill='grey')+
#  geom_bar(stat='identity', colour='black')+
#  geom_smooth(span=0.5, se=FALSE, linetype='dashed', col='red')+
#  facet_wrap(~country+SO_startDOY, scales='free_y')+theme_classic()+
#  ylab('daily rainfall (mm)')+ xlab('Day of year')+
#          ggtitle(paste("Year:",i,"   PTT:", j, "   Arr date:", #dtemp$arrive_breeding, "   Mig speed:", round(dtemp$m4c_resid, 2)))

  }
}

```

### 2018-07-04 bioclimatic_niche.R
First stab at bioclimatic unit stuff. Initially doing dynamic rainfall and NDVI

@@ -1,0 +1,50 @@
+########################
+# 03/07/2018
+# Defining the bioclimatic units cuckoos use on spring migration in Africa 
+# Using environemntal niche modelling but will need to be dynamic over time
+########################
+
+rm(list=ls())
+
+library(ggplot2)
+
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+
+# Use data extracted with the 'detailed coords' method
+
+dat<-read.csv('DATA/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+# summarise data for each stopover so that value per date is the mean
+# of the detail coords values
+
+library(dplyr)
+
+dat2<- dat %>% group_by(year, ptt, SO_startDOY, variable) %>%
+  summarise(country=first(country), SO_endDOY=first(SO_endDOY),
+            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
+            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),
+            terrANOM=mean(terrANOM, na.rm=T))
+
+# rearrange
+
+dat2<- dat2 %>% select(year, ptt, SO_startDOY, SO_endDOY, country, everything())
+
+precip<-dat2[,1:8]
+
+# add cumulative rainfall
+precip[which(is.na(precip$precip)),]$precip # fix na
+
+precip<-precip %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
+
+# drops duplicated daily values between 8 day modis images
+ndvi<-dat2[dat2$variable %in% seq(1, 125, 8),] 
+
+# rainfall plots
+
+p<-ggplot(data=precip[precip$variable %in% seq(1:31),],
+          aes(x=cumrf, colour=cuck_pres))
+
+p+geom_histogram(position='dodge')

### 2018-07-04 cuckoo_stopovers.R
additional code to tidy data for Durham uni

@@ -76,6 +76,8 @@ dat2$dead=1
 
 dat<-rbind(dat1, dat2)
 
+#write.csv(dat, "data/stopover_bestofday_2018_1daymin_recalc_spring_mig_BOTH.csv", quote=F, row.names=F)
+
 #do subset using join
 
 #drop uneeded columns from stopopvers
@@ -309,17 +311,67 @@ write.csv(dat, "data/stopover_table_bestofday_2018_1daymin_recalc_biomes.csv", q
 ## Add columns with migration cohort to 
 ## sort the issue of crossing years on migration manually in excel
 
+## cleaning up data for Durham uni provision
+
+dat<-read.csv("data/stopover_table_bestofday_2018_1daymin_recalc_biomes.csv", h=T)
+
+# remove unhelpful median rounded month and year
+
+dat$SO_month<-NULL
+dat$SO_year<-NULL
+
 ## Manual fill of NA countries from GIS lookup
 
 dat[which(is.na(dat$country)),]
 dat[2,]$country<-'Belgium'
 dat[c(426,442),]$country<-'United Kingdom'
-dat[c(426,442),]$country<-'United Kingdom'
 dat[193,]$country<-'Netherlands'
 dat[207,]$country<-'United Kingdom'
-dat[833,]$country<-'United Kingdom'
-dat[568,]$country<-'Spain'
+dat[846,]$country<-'United Kingdom'
 dat[570,]$country<-'United Kingdom'
+
+#add julian
+
+dat$SO_start <- as.POSIXct(strptime(dat$SO_start, "%Y-%m-%d %H:%M:%S"), "UTC")
+dat$SO_end <- as.POSIXct(strptime(dat$SO_end, "%Y-%m-%d %H:%M:%S"), "UTC")
+
+library(lubridate)
+
+dat$SO_startDOY<-yday(dat$SO_start) 
+
+dat$SO_endDOY<-yday(dat$SO_end) 
+
+dat$SO_length_hr<-dat$SO_days*24
+
+dat$SO_length_discrete_day<-(dat$SO_endDOY-dat$SO_startDOY)+1
+
+dat[dat$SO_length_discrete_day<0  & 
+      substr(dat$SO_start, 1,4)%in% c('2011', '2013','2014','2015','2017'),]$SO_length_discrete_day<-365-
+  (dat[dat$SO_length_discrete_day<0  & 
[…]

### 2018-07-04 ndvi_compile.R
Additions for filtering MODIS data - did not run due to processing limitations

@@ -49,22 +49,94 @@ plot(raster::subset(rastack, 1))
 
 library(gdalUtils)
 
-tifz<-list.files("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI")
+#tifz<-list.files("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI")
+
+# And for quality fields
+tifz<-list.files("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/QA_qual")
 
 #order them by date rather than satellite (Terra and Aqua)
-tifz<-tifz[order(substr(tifz, 14,21))]
+#tifz<-tifz[order(substr(tifz, 14,21))]
+#
+tifz<-tifz[order(substr(tifz, 17, 24))]
+
+p<-projection(raster(paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/QA_qual/",tifz[1])))
 
 for(i in tifz)
 {
-  gdalwarp(srcfile=paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI/",i),
-           dstfile=paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/NDVI_WGS/",i),
-           s_srs=projection(rastack),
+  gdalwarp(srcfile=paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/QA_qual/",i),
+           dstfile=paste0("C:/cuckoo_tracking/sourced_data/MODIStsp_NDVI/VI_16Days_1Km_v6/QA_qual_WGS/",i),
+           s_srs=p,
            t_srs='+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs',
            output_Raster=TRUE,
            overwrite=TRUE,verbose=TRUE)
   print(i)  
 }
 # much faster!
+
+# Ok now we apply a savitzky-golay-filter to smooth the NDVI timeseries
+# we will try two methods:
+#1) using raw ndvi data and
+#2) after masking out poor quality pixels
[…]

### 2018-07-04 paper_tables.R
addition for sahara migration length - need to refine

@@ -23,7 +23,7 @@ dat$mig_length<-dat$arrive_breeding-dat$DEPwestAF
 dat$WA_dur<-dat$DEPwestAF-dat$DEPcentralAF
 qplot(data=dat, x=year, y=WA_dur) # some large ones
 # hmm there are discrepencies.. use sumSOwestAF as conservative estimate
-dat$DEPwestAF-dat$DEPcentralAF - dat$sumSOwestAF
+dat$WA_dur<-dat$DEPcentralAF - dat$sumSOwestAF
 
 # correct em
 dat[dat$ptt==115586 & dat$year==2014,]$sumSOwestAF<-22.19065+4
@@ -38,6 +38,15 @@ dat$year<-relevel(dat$year, ref='2015')
 # relevel to central well sampled year
 dat$breeding_loc<-relevel(dat$breeding_loc, ref='Thetford Forest')
 
+## variable for sahara crossing length - check!
+
+dat$sahara_arr<-dat$DEPnorthAF-dat$sumSOnorthAF
+
+dat$sahara_arr<-ifelse(is.na(dat$sahara_arr),
+                       dat$DEPeurope-dat$sumSOeurope,
+                       dat$DEPnorthAF-dat$sumSOnorthAF)
+
+dat$sahara_mig<-dat$sahara_arr-dat$DEPwestAF
 
 library(lme4)
 

### 2018-07-04 bioclimatic_niche.R
progress on rainfall and ndvi

@@ -35,16 +35,95 @@ dat2<- dat2 %>% select(year, ptt, SO_startDOY, SO_endDOY, country, everything())
 precip<-dat2[,1:8]
 
 # add cumulative rainfall
-precip[which(is.na(precip$precip)),]$precip # fix na
+precip[which(precip$precip>500),]$precip<-0 # remove erroneous vals
+precip[which(is.na(precip$precip)),]$precip<-0 # fix na
 
 precip<-precip %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
 
+# Not doing now as drops needed info
 # drops duplicated daily values between 8 day modis images
-ndvi<-dat2[dat2$variable %in% seq(1, 125, 8),] 
+#ndvi<-dat2[dat2$variable %in% seq(1, 125, 8),]
+
+ndvi<-dat2
+ndvi$precip<-NULL
+
+# add month and region
+ndvi$month=1
+ndvi[ndvi$variable %in% (32:59),]$month=2
+ndvi[ndvi$variable %in% (60:90),]$month=3
+ndvi[ndvi$variable %in% (91:125),]$month=4
+
+ndvi$region<-'WestAF'
+
+ndvi[ndvi$country %in% c("Congo","Cameroon",
+                     "Central African Republic",
+                     "Democratic Republic of the Congo",
+                     "Angola", "Gabon"),]$region<-"CentralAF"
+
+
+# ndvi plots
+
+#across all sites
+p<-ggplot(data=ndvi, aes(x=NDVI, fill=cuck_pres))
+
+p+geom_histogram(position='dodge')+facet_wrap(~month)
+
+p+geom_histogram(position='dodge')+facet_grid(region~month, scales='free_y')
+
+# order countries by longitude
+levels(ndvi$country)
+ndvi$country = factor(ndvi$country,
+                      levels(ndvi$country)[c(1,2,11,13,10,6,9,14,12,3,4,8,5,7)])
+
+p<-ggplot(data=ndvi[-which(ndvi$country %in% c('Angola', 'Burkina Faso', 'Liberia')),],
+          aes(x=aquaNDVI, fill=cuck_pres))
+
+jpeg("outputs/aquaNDVI_month_country.jpg",
+     width =   11.69, height =8.27 , units ="in", res =300)
+
+p+geom_histogram(position='dodge', binwidth=0.05)+
+  scale_fill_manual(values=c("N" = "grey", "Y"="black"))+
[…]

### 2018-07-04 rainfall_ndvi_extract.R
distance to river extract

@@ -268,6 +268,10 @@ int2$terrANOM<-na.locf(int2$terrANOM)
 write.csv(int2, 'data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', row.names = F, quote=F)
 
 
+
+
+## OLD
+
 ### Spatial extraction and analyses.
 ### comparison between rainfall at start of stopover n
 ### against rainfall of all other stopovers extract for n's date

### 2018-07-04 static_extract.R

@@ -99,11 +99,18 @@ spdatproj<-spTransform(spdat, CRS=CRS("+proj=aea +lat_1=20 +lat_2=-23 +lat_0=0 +
 
 riversproj<-spTransform(riversCROP, CRS=CRS("+proj=aea +lat_1=20 +lat_2=-23 +lat_0=0 +lon_0=25 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"))
 
-for(i in 1:nrow(dat))
-{
-dat[i,]$d_river<-gDistance(spdatproj[i,], riversproj)
-print(i)
-}
+# byid creaes matrix of min distances from every river to point
+rivers2<-gDistance(spdatproj, riversproj, byid=T)
+
+# apply function to find the nearest river from the matrix to each point
+dat$d_river<-apply(rivers2, 2, min)
+
+# do sma efor Strahler order => 2 indicating larger rivers
+
+rivers2<-gDistance(spdatproj, riversproj[riversproj$Strahler>1,], byid=T)
+
+# apply function to find the nearest river from the matrix to each point
+dat$d_river_str2<-apply(rivers2, 2, min)
 
 write.csv(dat, "data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", quote=F, row.names=F)
 

### 2018-07-04 stopover_rainfall_ndvi_detailcoords_vis.Rmd
adding d rivers to stopover knitr visualisation

@@
 
-lc<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext.csv", h=T)
+lc<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", h=T)
 
@@
 lc1<-lc1 %>% group_by(SO_startDOY, country) %>% dplyr::summarise(Tree=mean(Trees, na.rm=T), Grass=mean(Grasslands, na.rm=T), Crops=mean(Croplands, na.rm=T), Shrub=mean(Shrubs, na.rm=T),Bare=mean(Bare.areas, na.rm=T), Urban=mean(Built.up.areas, na.rm=T),
-  Aquatic=mean(Aquatic.veg.reg.flooded, na.rm=T), Water=mean(Open.water, na.rm=T))
+  Aquatic=mean(Aquatic.veg.reg.flooded, na.rm=T), Water=mean(Open.water, na.rm=T), D_riv=
+    mean(d_river), D_riv2=mean(d_river_str2))
 
@@
 }
+lcouty<-paste(lcouty, "D riv:", round(lc1[k,11]/1000,1), "km \n", "D riv2:", round(lc1[k,12]/1000,1), "km \n")
 

### 2018-07-09 rainfall_ndvi_extract.R
Fix to fill NA Modis values in GIMMS extract

@@ -233,6 +233,25 @@ grimms$START.DATE <-NULL
 grimms$END.DATE <-NULL
 grimms$ORDINAL.DATE <-NULL
 
+## extra code to fill the relicates of 'variable' ahead of the join 
+grimms2<-do.call('rbind', list(grimms,grimms,grimms,grimms,
+                              grimms,grimms,grimms,grimms))
+
+grimms2<-grimms2[order(grimms2$ID, grimms2$variable),]
+
+grimms2<-grimms2[grimms2$variable!=145,] # remove dataset 145 (d145-153)
+# as not needed, actually above 125 but anyways....
+
+# repeat sequence 1:144 for each ID (times 2 cos both satellites)
+grimms2$v2<-rep((sort(rep(1:144, 2))), length(unique(grimms2$ID)))
+
+#set grimms2 as grimms and v2 as variable
+grimms<-grimms2
+grimms$variable<-grimms$v2
+grimms$v2<-NULL
+
+
+# split into seperate satellites
 gaqua<-grimms[grimms$sat=='AQUA',]
 
 gaqua$sat<-NULL
@@ -256,14 +275,8 @@ int_out$ID<-paste(int_out$ptt, int_out$year,
 int2<-left_join(int_out, gaqua, by=c('ID', 'variable'))
 int2<-left_join(int2, gterr, by=c('ID', 'variable'))
 
-# the 'na.locf' function fills NA with previous value as desired
-library(zoo)
-int2$aquaNDVI<-na.locf(int2$aquaNDVI)
-int2$aquaANOM<-na.locf(int2$aquaANOM)
-
-int2$terrNDVI<-na.locf(int2$terrNDVI)
-int2$terrANOM<-na.locf(int2$terrANOM)
-
+# legit NAs from GIMMS dataset
+which(!is.na(int2$aquaNDVI))
 
 write.csv(int2, 'data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', row.names = F, quote=F)
 

### 2018-07-10 paper_tables.R
Getting carried away with paper analyses. will tidy later

@@ -2,6 +2,13 @@
 # Making output tables for cuckoo paper. Initially descriptive tables
 # of migration metrics from previously calculated stopover summary table
 
+#### work around for different laptop/desktop directories ####
+#### !!!! RUN THIS !!!! ####
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+#### !!!!!!!!!!!!!!!!!! ####
+
 # read in most complete dataset
 
 dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)
@@ -18,6 +25,111 @@ dat[dat$ptt==128296 & dat$year==2014,]$DEPwestAF<-NA
 
 #set up migration length variable
 dat$mig_length<-dat$arrive_breeding-dat$DEPwestAF
+
+qplot(data=dat, x=DEPwestAF, y=mig_length)
+
+
+qplot(data=dat, x=DEPwestAF, y=WA_mig_len, colour=breeding_loc)+facet_wrap(~year)
+
+library(lme4)
+
+m4<-lmer(arrive_breeding~DEPwestAF+
+           (1|ptt)+(1|year), data=dat)
+
+summary(m4)
+
+m4c<-lm(arrive_breeding~
+          DEPwestAF, data=dat)
+
+
+# plot
+
+newdat<-data.frame(DEPwestAF=70:130, arrive_breeding=1)
+newdat$p1<-predict(m4, newdata=newdat, re.form=~0)
+
+predmat<-model.matrix(arrive_breeding~DEPwestAF, data=newdat)
+vcv<-vcov(m4c)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+
+qplot(data=dat, x=DEPwestAF, y=arrive_breeding)+
+  geom_line(data=newdat, aes(x=DEPwestAF, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=DEPwestAF, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=DEPwestAF, y=uc), colour='red', linetype='dashed')+
[…]

### 2018-07-10 paper_wireframe_knit.Rmd
Initial setup of wireframe for first paper

---
title: "Paper Wireframe"
author: "Mark Miller"
date: "10 July 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=F}
if(Sys.info()['nodename']=="D9L5812"){
  setwd("C:/cuckoo_tracking")}else{
    setwd("N:/cuckoo_tracking")}

dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)

# set NA or manually correct cuckoo depature dates that
# are incorrect, from manual examination

dat[dat$ptt==115586 & dat$year==2014,]$DEPwestAF<-106
dat[dat$ptt==134955 & dat$year==2015,]$DEPwestAF<-NA
dat[dat$ptt==134956 & dat$year==2015,]$DEPwestAF<-NA

# and dead ones
dat[dat$ptt==128296 & dat$year==2014,]$DEPwestAF<-NA

```

## Aims

Focus on cuckoo spring departure from West Africa. Address several questions:

Q1) Is cuckoo arrival in the UK constrained by their departure from West Africa ?

Q2) To what extent is cuckoo departure from West Africa controlled by endogenous or environmental factors ?

Q3) What are the mechanisms that underpin potential controlling factors ?

## Methods

## Results

#Q1)

```{r pressure, echo=FALSE}
plot(pressure)
```

### 2018-07-11 paper_wireframe_knit.Rmd
Progress with Q1 mainly

@@
 ```{r, echo=F}
+library('lme4')
+library('car')
+library('ggplot2')
+
 if(Sys.info()['nodename']=="D9L5812"){
@@
 
-#Q1)
+### Q1)
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+Arrival at UK breeding grounds is less correlated with earlier stages of spring migration, but appears to be still be influenced by movements in West Africa.
+
+```{r, echo=F, fig.cap='Correlation between arrival at UK breeding grounds and stages of spring migration: departure from furthest southerly wintering latitude (pink); Central Africa (cyan); West Africa (blue); North Africa (green); Europe (red); and arrival on UK mainland (black).', warning=F, message=F}
+
+ggplot(data=dat, aes(y=arrive_breeding))+
+  geom_point(aes(x=arrive_uk), col=1, shape=1)+
+  geom_point(aes(x=DEPeurope), col=2, shape=1)+
+  geom_point(aes(x=DEPnorthAF), col=3, shape=1)+
+  geom_point(aes(x=DEPwestAF), col=4, shape=1)+
+  geom_point(aes(x=DEPcentralAF), col=5, shape=1)+
+  geom_point(aes(x=depart_winterSO), col=6, shape=1)+
+  xlim(c(-71, 142))+ylab('Arrival at breeding grounds DOY')+
+  xlab('DOY')+theme_classic()
+
 ```
+
+Stats confirm that date of departure from West Africa significantly impacts subsequent date of arrival at Uk breeding grounds, but prior stages of spring migration do not.
+
+```{r, echo=F}
+d1<-dat
+  names(d1)[10]<-'arrive.breeding.uk'
+  names(d1)[4]<-'depart.wintering'
+  names(d1)[5]<-'depart.central.africa'
+  names(d1)[6]<-'depart.west.africa'
+
+m1<-lmer(arrive.breeding.uk~depart.wintering+depart.central.africa+depart.west.africa+(1|year)+(1|ptt), data=d1)
+
+Anova(m1, test.statistic='F')
+
+```
+
+For every extra day spent in West Africa cuckoos arrive 0.78 days 
+later back at UK breeding grounds.
+
+```{r, warning=F, message=F, echo=F}
+m4<-lmer(arrive_breeding~DEPwestAF+
+           (1|ptt)+(1|year), data=dat)
+
+newdat<-data.frame(DEPwestAF=70:130, arrive_breeding=1)
+newdat$p1<-predict(m4, newdata=newdat, re.form=~0)
+
+predmat<-model.matrix(arrive_breeding~DEPwestAF, data=newdat)
+vcv<-vcov(m4)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+
+qplot(data=dat, x=DEPwestAF, y=arrive_breeding)+
+  geom_line(data=newdat, aes(x=DEPwestAF, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=DEPwestAF, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=DEPwestAF, y=uc), colour='red', linetype='dashed')+
+  xlab("Departure date West Africa (DOY)")+
+  ylab("Arrival at breeding grounds (DOY)")+
+  theme_bw()
+```
+
+There was only very slight support for cuckoos being able to compensate for later departure from West Africa through faster migration.
+*might omit, likely influenced by outliers*
+
+```{r, warning=F, message=F, echo=F}
+d1$migration.length<-d1$arrive.breeding.uk-d1$depart.west.africa
+
+m5<-lmer(migration.length~depart.west.africa+
+           (1|ptt)+(1|year), data=d1)
+
+qplot(data=d1, x=depart.west.africa, y=migration.length)+
+  xlab("Departure date West Africa (DOY)")+
+  ylab("Migration length (days)")+
+  theme_bw()
+
+Anova(m5, test.statistic = 'F')
+```
+
+### Q2)
+
+```{r, warning=F, message=F, echo=F}
+
+dat$year<-as.character(dat$year)
+
+#reorder breeding location in latitude order
+
+# relevel to central well sampled year
+dat$breeding_loc<-factor(dat$breeding, levels=c('Scotland', 'Lake District', 'North York Moors', 'Sherwood Forest',  'Wales', 'Thetford Forest', 'Norfolk Broads','Ashdown Forest', 'New Forest', 'Dartmoor'))
+
+m1<-lmer(DEPwestAF~year+breeding_loc+year+(1|ptt), data=dat)
+
+library(stargazer)
+
+stargazer(m1, type='text',
+covariate.labels=c("Constant: Scotland, 2012","2013", "2014","2015", "2016", "2017", "2018", 'Lake District', 'North York Moors', 'Sherwood Forest',  'Wales', 'Thetford Forest', 'Norfolk Broads','Ashdown Forest', 'New Forest', 'Dartmoor'),
+dep.var.labels ="West Africa depature date",
+ci=TRUE, intercept.bottom = FALSE, omit.table.layout = "sn", 
+digits=1, model.numbers=FALSE,  report='vcs')
+
+```
+
+```{r, warning=F, message=F, echo=F}
+
+env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+# summarise data for each stopover so that value per date is the mean
+# of the detail coords values
+
+library(dplyr)
+
+dat2<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
+  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
+            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
+            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),
+            terrANOM=mean(terrANOM, na.rm=T))
+
+dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
+dat2[which(is.na(dat2$precip)),]$precip<-0 # fix na
+
+dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
+
+# rearrange
+
+dat2<- dat2 %>% select(year, ptt, SO_startDOY, SO_endDOY, country, everything())
+
+dat2.1<-subset(dat2, cuck_pres=='Y')
+
+dat2.2<-dat2.1 %>% group_by(year, ptt, SO_startDOY) %>%
+  summarise(cumrf=mean(cumrf, na.rm=T),
+            NDVI= mean(NDVI, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),
+            terrANOM=mean(terrANOM, na.rm=T),
+            SOlen=first(SO_endDOY)-first(SO_startDOY))
+
+dat2.3<-dat2.2 %>% group_by(year, ptt) %>%
+  summarise(cumrf=last(cumrf),
+            NDVI= last(NDVI), 
+            aquaNDVI=last(aquaNDVI),
+            aquaANOM=last(aquaANOM),
+            terrNDVI=last(terrNDVI),
+            terrANOM=last(terrANOM),
+            last_SOstart=last(SO_startDOY))
+
+dat$year<-as.integer(dat$year)
+
+dat3<-full_join(dat, dat2.3, by=c('ptt', 'year'))
+
+m6<-lmer(DEPwestAF~aquaANOM+
+           (1|ptt)+(1|year), data=dat3)
+
+summary(m6)
+
+# plot
+
+newdat<-data.frame(aquaANOM=seq(-0.15,0.15, 0.005), DEPwestAF=1)
+newdat$p1<-predict(m6, newdata=newdat, re.form=~0)
+
+predmat<-model.matrix(DEPwestAF~aquaANOM, data=newdat)
+vcv<-vcov(m6)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+
+qplot(data=dat3, x=aquaANOM, y=DEPwestAF)+
+  geom_line(data=newdat, aes(x=aquaANOM, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=aquaANOM, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=aquaANOM, y=uc), colour='red', linetype='dashed')+
+  xlab("NDVI anomoly")+
+  ylab("Departure date West Africa (DOY)")+
+  theme_bw()
+
+Anova(m6, test.statistic = 'F')
+
+```

### 2018-07-11 stopover_rainfall_ndvi_detailcoords_vis.Rmd
Added NDVI anomaly values to plot

@@
   geom_line(data=t2, aes(x=variable, y=terrNDVI*500), colour='orange')+
+    geom_line(data=t2, aes(x=variable, y=aquaANOM*1000+200), colour='cyan', linetype=2)+
+    geom_line(data=t2, aes(x=variable, y=terrANOM*1000+200), colour='orange', linetype=2)+
+  geom_hline(yintercept=200)+
   geom_bar(stat='identity', colour='black')+

### 2018-07-13 stopover_rainfall_ndvi_detailcoords_vis.Rmd
Added code to graph TAMSAT rainfall

--- 2018-07-04 stopover_rainfall_ndvi_detailcoords_vis.Rmd
@@
 
-int_out<-read.csv('data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', h=T)
+int_out<-read.csv('data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', h=T)
 
@@
 
+t2.tmcum<- t1.3 %>% group_by(ptt, year, SO_startDOY,country, key) %>% dplyr::summarise(tamRAIN=first(tamRAIN), variable=first(variable)) %>%
+  dplyr::mutate(cumrfTAM=cumsum(tamRAIN))
+
 temp=ggplot(data=t2,aes(x=variable, y=value))+
@@
   geom_line(data=t2, aes(x=variable, y=terrNDVI*500), colour='orange')+
+    geom_line(data=t2, aes(x=variable, y=aquaANOM*1000+200), colour='cyan', linetype=2)+
+    geom_line(data=t2, aes(x=variable, y=terrANOM*1000+200), colour='orange', linetype=2)+
+  geom_line(data=t2, aes(x=variable, y=tamRAINanom+200), colour='dark green', linetype=2)+
+  geom_hline(yintercept=200)+
   geom_bar(stat='identity', colour='black')+
     geom_line(aes(y=cumrf), col='green', size=1)+
+  geom_line(data=t2.tmcum, aes(x=variable, y=cumrfTAM), col='dark green', size=1)+
   facet_wrap(~country+SO_startDOY, scales='free_y')+theme_classic()+

### 2018-07-13 rainfall_ndvi_extract.R
Code added to extract TAMSAT rainfall and anomoly data

@@ -281,6 +281,89 @@ which(!is.na(int2$aquaNDVI))
 write.csv(int2, 'data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', row.names = F, quote=F)
 
 
+#################
+# extra code to extract TAMSAT rainfall and anomoly data also and
+# add to existing file
+
+library(raster)
+library(dplyr)
+
+int2<-read.csv('data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+# get coord data
+dat<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords.csv', h=T)
+
+# join lat long
+
+dat$ID=paste(dat$ptt, dat$year, dat$SO_startDOY, dat$timestamp, sep='_')
+
+int2$ID <-as.character(int2$ID)
+
+int2<-left_join(int2, select(dat, ID, long, lat), by='ID')
+
+# rearrange
+int2<- int2 %>% select(ID, ptt, year, SO_startDOY, SO_endDOY, country,
+                       timestamp, long, lat, everything())
+
+int2$tamRAIN<-NA
+
+int2$tamRAINanom<-NA
+
+# set up key to get data
+
+int2$key<-'01-pt1'
+int2[int2$variable %in% 6:10,]$key <-'01-pt2'
+int2[int2$variable %in% 11:15,]$key <-'01-pt3'
+int2[int2$variable %in% 16:20,]$key <-'01-pt4'
+int2[int2$variable %in% 21:25,]$key <-'01-pt5'
+int2[int2$variable %in% 26:30,]$key <-'01-pt6'
+
+int2[int2$variable %in% 31:35,]$key <-'02-pt1'
+int2[int2$variable %in% 36:40,]$key <-'02-pt2'
+int2[int2$variable %in% 41:45,]$key <-'02-pt3'
+int2[int2$variable %in% 46:50,]$key <-'02-pt4'
+int2[int2$variable %in% 51:55,]$key <-'02-pt5'
+int2[int2$variable %in% 56:60,]$key <-'02-pt6'
+
+int2[int2$variable %in% 61:65,]$key <-'03-pt1'
+int2[int2$variable %in% 66:70,]$key <-'03-pt2'
+int2[int2$variable %in% 71:75,]$key <-'03-pt3'
[…]

### 2018-07-16 paper_wireframe_knit.Rmd
Getting tests in logical order and finishing endogenous section

@@
 
-```{r, echo=F}
+```{r, echo=F, warning=F, message=F}
 library('lme4')
@@
 
+#### Data cleaning
+
 # set NA or manually correct cuckoo depature dates that
@@
 dat[dat$ptt==128296 & dat$year==2014,]$DEPwestAF<-NA
+
+## fill NA in mig route
+
+extraz<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv', h=T)
+
+# fills some
+dat<-left_join(dat, select(extraz, ptt, year, autumn_mig), by=c('ptt', 'year'))
+
+# fill 2018 birds
+dat[c(47, 44, 42),]$autumn_mig.y='SW'
+dat[39,]$autumn_mig.y='SE'
+# and dead birds
+dat[c(2,20,45),]$autumn_mig.y='SW'
+dat[c(1,3,23,25,31),]$autumn_mig.y='SE'
+
+names(dat)[names(dat)=='autumn_mig.y']<-'autumn_mig'
+dat$autumn_mig.x<-NULL
 
@@
   xlim(c(-71, 142))+ylab('Arrival at breeding grounds DOY')+
-  xlab('DOY')+theme_classic()
+  xlab('DOY')+theme_bw()
 
@@
 
+## Endogenous factors that influence dpearture date from West Africa.
+
+No convincing evidence that different breeding populations depart West Africa at different times. *slight significance comes from very early New Forest population*
+
 ```{r, warning=F, message=F, echo=F}
 
-dat$year<-as.character(dat$year)
-
 #reorder breeding location in latitude order
-
 # relevel to central well sampled year
@@
 
-m1<-lmer(DEPwestAF~year+breeding_loc+year+(1|ptt), data=dat)
+m1<-lmer(DEPwestAF~breeding_loc+(1|year)+(1|ptt), data=dat)
+
+Anova(m1, test.statistic = 'F')
+
+qplot(data=dat, x=breeding_loc, y=DEPwestAF)+theme_bw()+theme(axis.text.x=element_text(angle=90))+xlab("Breeding location")+ylab("Departure date West Africa (DOY)")+geom_label(data=data.frame(), aes(label = 'Anova F=4.55, p=0.023', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+No evidence that different autumn migration routes impact timing of spring departure from West Africa.
+
+```{r, warning=F, message=F, echo=F}
+# and for migratory route
+m1<-lmer(DEPwestAF~autumn_mig+(1|year)+(1|ptt), data=dat)
+
+Anova(m1, test.statistic = 'F')
+
+qplot(data=dat, x=autumn_mig, y=DEPwestAF, geom='boxplot')+xlab("Autumn migratory route")+ylab("Departure date West Africa (DOY)")+geom_label(data=data.frame(), aes(label = 'Anova F=0.81, p=0.373', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+For birds that were tracked through multiple spring migrations there are inter-bird differences in West African departure date and varying degrees of
+inter-annual variation within birds.
+
+```{r, warning=F, message=F, echo=F}
+multi_yr<-names(which(table(dat$ptt)>1))
+
+multi_yr<-multi_yr[multi_yr!=128300]
+
+m1<-lmer(DEPwestAF~factor(ptt)+(1|year), data=dat[dat$ptt %in% multi_yr ,])
+
+Anova(m1, test.statistic = 'F')
+
+ggplot(data=dat[dat$ptt %in% multi_yr ,], aes(x=factor(ptt), y=DEPwestAF ))+xlab("Bird ID")+ylab("Departure date West Africa (DOY)")+geom_jitter(width=0.08,shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=4.13, p=0.017', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+
+```
+
+## Environmental factors that influence departure date from West Africa.
+
+First, there appears no overall difference between years in departure date from West Africa (median: 102 DOY).
+*This might support internal regulation for timing of migration out of West Africa by birds e.g.
+photoperiod (endogenous) but could be an artefact of sample size - more tracking would reveal statistically significant differences between years?*
+
+```{r, warning=F, message=F, echo=F}
+dat$year<-as.character(dat$year)
+
+m1<-lmer(DEPwestAF~year+(1|ptt), data=dat)
+
+Anova(m1, test.statistic = 'F')
+
+ggplot(data=dat, aes(x=year, y=DEPwestAF ))+xlab("Year")+ylab("Departure date West Africa (DOY)")+geom_jitter(width=0.08,shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=1.27, p=0.315', x = -Inf, y = Inf), hjust = 0, vjust = 1)
 
@@
 stargazer(m1, type='text',
-covariate.labels=c("Constant: Scotland, 2012","2013", "2014","2015", "2016", "2017", "2018", 'Lake District', 'North York Moors', 'Sherwood Forest',  'Wales', 'Thetford Forest', 'Norfolk Broads','Ashdown Forest', 'New Forest', 'Dartmoor'),
+covariate.labels=c("Constant: 2012","2013", "2014","2015", "2016", "2017", "2018"),
 dep.var.labels ="West Africa depature date",
@@
 
-env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_by_stopover_detailcoords_2018_dead.csv', h=T)
+env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', h=T)
 
@@
             NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),
-            terrANOM=mean(terrANOM, na.rm=T))
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
 
@@
             NDVI= mean(NDVI, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),
-            terrANOM=mean(terrANOM, na.rm=T),
-            SOlen=first(SO_endDOY)-first(SO_startDOY))
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),terrANOM=mean(terrANOM, na.rm=T),tamRAIN=mean(tamRAIN, na.rm=T),tamRAINanom=mean(tamRAINanom, na.rm=T), SOlen=first(SO_endDOY)-first(SO_startDOY))
 
@@
             terrANOM=last(terrANOM),
+             tamRAIN=last(tamRAIN),
+            tamRAINanom=last(tamRAINanom),
             last_SOstart=last(SO_startDOY))

### 2018-07-17 paper_wireframe_knit.Rmd
Tidying and addition of intra-bird final WA stopover plot

@@
 library('ggplot2')
+library('dplyr')
+library('maps')
+library('stargazer')
 
@@
 
+#summary data
 dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)
+
+#stopover coord data (median points)
+sovers<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_BOTH.csv', h=T)
 
@@
 
-For birds that were tracked through multiple spring migrations there are inter-bird differences in West African departure date and varying degrees of
-inter-annual variation within birds.
+For birds that were tracked through multiple spring migrations there are inter-bird differences in West African departure date and differing levels of inter-annual variation within birds.
 
@@
 
+```
+
+Differing levels of intra-bird variation were also evident in final stopover location before departure from West Africa
+
+```{r, warning=F, message=F, echo=F}
+
+multi_yr_SO<-sovers[sovers$ptt %in% multi_yr,]
+
+notWA<-c('Italy', 'France', 'Spain', 'Algeria',
+        'United Kingdom', 'Morocco')
+
+multi_yr_SO<-multi_yr_SO[-which(multi_yr_SO$country %in% notWA),]
+
+lastWASO<-multi_yr_SO %>% group_by(ptt, year) %>% summarise(long=last(SO_median_long), lat=last(SO_median_lat), country=last(country))
+
+map = map_data("world")
+map = subset(map, region=='Africa')
+
+ggplot() + geom_polygon(data = map,
+               aes(x=long, y = lat, group = group),
+               fill = NA, colour="darkgray", size=0.5)+
+  geom_point(data=lastWASO, aes(x=long, y=lat, colour=factor(ptt)), shape=1, size=2)+
+  geom_line(data=lastWASO, aes(x=long, y=lat, colour=factor(ptt)), alpha=0.2)+coord_cartesian()+
+   coord_map(xlim = c(-16, 5),ylim = c(14, 3))+xlab('Longitude')+ylab('Latitude')+theme_bw()+ labs(colour = "Bird ID")
 ```
@@
 
-library(stargazer)
-
 stargazer(m1, type='text',
@@
 # of the detail coords values
-
-library(dplyr)
 

### 2018-07-17 paper_wireframe_knit.Rmd
Addition of habitat and leaflet viewer

@@
 
+## landcover and river data
+lc<-read.csv("C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", h=T)
+
+#extra data for filling migratory route
+
+extraz<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv', h=T)
+
 #### Data cleaning
@@
 ## fill NA in mig route
-
-extraz<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv', h=T)
-
 # fills some
@@
 dat$autumn_mig.x<-NULL
+
+### Add departure info from WA e.g. country, lat and long, length of final stopover and 2nd before last
+
+notWA<-c('Italy', 'France', 'Spain', 'Algeria',
+        'United Kingdom', 'Morocco', 'Portugal')
+
+soversWA<-sovers[-which(sovers$country %in% notWA),]
+
+dat<-left_join(dat, soversWA %>% group_by(ptt, year) %>% summarise( dep_country=last(country), dep_long=last(SO_median_long), dep_lat=last(SO_median_lat), st_finalSO=last(SO_startDOY), dur_finalSO_n2=(nth(SO_endDOY, -2)-nth(SO_startDOY, -2))), by=c('ptt', 'year'))
+
+#calc final stopover length (cos we corrected some values in the summary table)
+
+dat$dur_finalSO=dat$DEPwestAF-dat$st_finalSO
+
+#landcover metrics
+
+lc1<-lc %>% group_by(ptt, year, SO_startDOY) %>% dplyr::summarise(Tree=mean(Trees, na.rm=T), Grass=mean(Grasslands, na.rm=T), Crops=mean(Croplands, na.rm=T), Shrub=mean(Shrubs, na.rm=T),Bare=mean(Bare.areas, na.rm=T), Urban=mean(Built.up.areas, na.rm=T),
+  Aquatic=mean(Aquatic.veg.reg.flooded, na.rm=T), Water=mean(Open.water, na.rm=T), D_riv=
+    mean(d_river), D_riv2=mean(d_river_str2))
+
+# landcover types expressed as a precentage                                                            
+lc1[,4:10]<-round(lc1[,4:10]/rowSums(lc1[,4:10], na.rm=T) *100)                             
+# join stopover duration to test associations
+
+lc1<-left_join(lc1, select(soversWA, ptt, year, SO_startDOY, SO_days), by=c('ptt', 'year', 'SO_startDOY'))
 
@@
 
-Anova(m5, test.statistic = 'F')
+#Anova(m5, test.statistic = 'F')
 ```
@@
 
-## Endogenous factors that influence dpearture date from West Africa.
+## Endogenous factors that influence departure date from West Africa.
 
@@
 
-Anova(m1, test.statistic = 'F')
+#Anova(m1, test.statistic = 'F')
 
@@
 
-```{r, warning=F, message=F, echo=F}
+```{r, warning=T, message=T, echo=F}
 
@@
 map = map_data("world")
-map = subset(map, region=='Africa')
 
 ggplot() + geom_polygon(data = map,
-               aes(x=long, y = lat, group = group),
-               fill = NA, colour="darkgray", size=0.5)+
+               aes(x=long, y = lat, group = group),fill = NA, colour="darkgray", size=0.5)+
   geom_point(data=lastWASO, aes(x=long, y=lat, colour=factor(ptt)), shape=1, size=2)+
@@
    coord_map(xlim = c(-16, 5),ylim = c(14, 3))+xlab('Longitude')+ylab('Latitude')+theme_bw()+ labs(colour = "Bird ID")
+
 ```
@@
 digits=1, model.numbers=FALSE,  report='vcs')
+
+```
+
+Country of departure seems to have little overall affect on date of departure but note that birds who leave early (< day 100) all leave from Ivory Coast and (to a lesser extent) Ghana.
+
+```{r, warning=F, message=F, echo=F}
+
+#Anova(lmer(DEPwestAF~dep_country+ (1|ptt)+(1|year), data=dat), test.statistic = 'F')
+
+dtemp<-na.omit(dat[,c(6,22:24)])
+
+dtemp$dep_country<-factor(dtemp$dep_country, levels=c('Sierra Leone', 'Guinea', "Cote d'Ivoire", 'Ghana', 'Burkina Faso', 'Nigeria'))
+m1<-lmer(DEPwestAF~breeding_loc+(1|year)+(1|ptt), data=dat)
+
+#Anova(m1, test.statistic = 'F')
+
+ggplot(data=dtemp, aes(x=dep_country, y=DEPwestAF ))+xlab("Departure country")+ylab("Departure date West Africa (DOY)")+geom_jitter(width=0.1,shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.552, p=0.735', x = -Inf, y = Inf), hjust = 0, vjust = 1)+theme(axis.text.x=element_text(angle=90))
+```
+
+ Habitat?? - test each category against dep from west africa
+
+```{r, warning=F, message=F, echo=F} 
+ggplot(data=dat, aes(x=dur_finalSO, y=DEPwestAF ))+xlab("Departure country")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()
+
+```
+
+Explore dataset
+
+```{r, warning=F, message=F, echo=F} 
+library(sf)
+library(leaflet)
+
+dat_sp<-st_as_sf(na.omit(dat[,c(1,2,6,23,24)]), coords = c("dep_long","dep_lat"), crs="+proj=longlat +datum=WGS84")
+
+m <- leaflet() %>%
+  addTiles() %>% 
+  addMarkers(data = dat_sp, popup= paste("ptt:", dat_sp$ptt, "<br>", "year:", dat_sp$year,"<br>", "Departure date:", dat_sp$DEPwestAF))
+
+m
 
@@
 
-summary(m6)
+#summary(m6)
 
@@
   ylab("Departure date West Africa (DOY)")+
-  theme_bw()
+  theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=4.65, p=0.038', x = -Inf, y = Inf), hjust = 0, vjust = 1)+theme(axis.text.x=element_text(angle=90))
 
-Anova(m6, test.statistic = 'F')
+#Anova(m6, test.statistic = 'F')
 

### 2018-07-18 paper_wireframe_knit.Rmd
More progress, added section on rainfall

@@
 library('stargazer')
+library('splines')
+library('sf')
+library('leaflet')
 
@@
 # join stopover duration to test associations
+# lc1 becomes dataset to test stopover length in relation to habitat
 
 lc1<-left_join(lc1, select(soversWA, ptt, year, SO_startDOY, SO_days), by=c('ptt', 'year', 'SO_startDOY'))
+
+## join landcover of final stopover to dat
+dat<-left_join(dat, (lc1 %>% group_by(ptt, year) %>% dplyr::summarise_all(last)), by=c('ptt', 'year'))
 
@@
 
-There was only very slight support for cuckoos being able to compensate for later departure from West Africa through faster migration.
-*might omit, likely influenced by outliers*
+Cuckoos that departed West Africa early had longer migrations back to the breeding grounds but those that departed after median departure date (day 102) could not compensate for delayed departure with faster migration.
 
@@
 
+# try segmented model using median departure date as break point
+# doent work with mixed model
+
+m5a<-lm(migration.length~depart.west.africa*(depart.west.africa<102)+depart.west.africa*(depart.west.africa>102), data=d1)
+
+m5b<-lmer(migration.length~bs(depart.west.africa, deg=1, knots=102)+
+           (1|ptt)+(1|year), data=d1)
+
+newdat<-data.frame(depart.west.africa=70:130, migration.length=1)
+newdat$p1<-predict(m5b, newdata=newdat, re.form=~0)
+
 qplot(data=d1, x=depart.west.africa, y=migration.length)+
@@
   ylab("Migration length (days)")+
-  theme_bw()
+  theme_bw()+geom_line(data=newdat, aes(x=depart.west.africa, y=p1), colour='red')+geom_label(data=data.frame(), aes(label = 'Anova F=3.54, p=0.046', x = -Inf, y = Inf), hjust = 0, vjust = 1)
 
 #Anova(m5, test.statistic = 'F')
+```
+
+## Focus on West Africa
+
+Explore the dataset. Each point gives the final stopover in West Africa before migration. Colours denote different years.
+
+```{r, warning=F, message=F, echo=F}
+
+dat_sp<-st_as_sf(na.omit(dat[,c(1,2,6,23,24)]), coords = c("dep_long","dep_lat"), crs="+proj=longlat +datum=WGS84")
+
+dat_sp$year<-factor(dat_sp$year)
+
+pal <- colorFactor(c("navy", "blue", 'cyan', 'green', 'yellow', 'orange', 'red'), domain = c("2012", "2013", '2014', '2015', '2016', '2017', '2018'))
+
+m <- leaflet(dat_sp) %>%
+  addTiles() %>% 
+  addCircleMarkers( popup= paste("ptt:", dat_sp$ptt, "<br>", "year:", dat_sp$year,"<br>", "Departure date:", dat_sp$DEPwestAF), color=~pal(year), stroke=F, radius=6, fillOpacity = 0.5)
+
+m
+
 ```
@@
 
-qplot(data=dat, x=breeding_loc, y=DEPwestAF)+theme_bw()+theme(axis.text.x=element_text(angle=90))+xlab("Breeding location")+ylab("Departure date West Africa (DOY)")+geom_label(data=data.frame(), aes(label = 'Anova F=4.55, p=0.023', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+ggplot(data=dat, aes(x=breeding_loc, y=DEPwestAF))+geom_jitter(width=0.08, shape=1)+theme_bw()+theme(axis.text.x=element_text(angle=90))+xlab("Breeding location")+ylab("Departure date West Africa (DOY)")+geom_label(data=data.frame(), aes(label = 'Anova F=4.55, p=0.023', x = -Inf, y = Inf), hjust = 0, vjust = 1)
 ```
@@
 
- Habitat?? - test each category against dep from west africa
+## Habitat Variables
+
+Habitat of the final stopover before departure from West Africa does not seem to have a huge imapact.
+
+The percentage of tree cover does not affect departure date.
 
 ```{r, warning=F, message=F, echo=F}
-ggplot(data=dat, aes(x=dur_finalSO, y=DEPwestAF ))+xlab("Departure country")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()
+
+m1<-lmer(DEPwestAF~Tree+(1|year)+(1|ptt), data=dat)
+
+#Anova(m1, test.statistic = 'F')
+
+ggplot(data=dat, aes(x=Tree, y=DEPwestAF ))+xlab("% tree cover of final stopover")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.27, p=0.609', x = -Inf, y = Inf), hjust = 0, vjust = 1)
 
@@
 
-Explore dataset
+The percentage of grass cover does appear to affect departure date. * Not sure how believeable this is*
 
 ```{r, warning=F, message=F, echo=F} 
-library(sf)
-library(leaflet)
+m1<-lmer(DEPwestAF~Grass+(1|year)+(1|ptt), data=dat)
 
-dat_sp<-st_as_sf(na.omit(dat[,c(1,2,6,23,24)]), coords = c("dep_long","dep_lat"), crs="+proj=longlat +datum=WGS84")
+#Anova(m1, test.statistic = 'F')
 
-m <- leaflet() %>%
-  addTiles() %>% 
-  addMarkers(data = dat_sp, popup= paste("ptt:", dat_sp$ptt, "<br>", "year:", dat_sp$year,"<br>", "Departure date:", dat_sp$DEPwestAF))
+newdat<-data.frame(DEPwestAF=1, Grass=1:75)
+newdat$p1<-predict(m1, newdata=newdat, re.form=~0)
 
-m
+predmat<-model.matrix(DEPwestAF~Grass, data=newdat)
+vcv<-vcov(m1)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+
+qplot(data=dat, y=DEPwestAF, x=Grass)+
+  geom_line(data=newdat, aes(x=Grass, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=Grass, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=Grass, y=uc), colour='red', linetype='dashed')+
+  ylab("Departure date West Africa (DOY)")+
+  xlab("% grass cover of habitat")+
+  theme_bw()+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=7.54, p=0.011', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+Distance to rivers does not affect departure date.
+
+```{r, warning=F, message=F, echo=F}
+
+m1<-lmer(DEPwestAF~D_riv+(1|year)+(1|ptt), data=dat)
+
+#Anova(m1, test.statistic = 'F')
+
+ggplot(data=dat, aes(x=D_riv, y=DEPwestAF ))+xlab("Distance to nearest river (m)")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.39, p=0.531', x = -Inf, y = Inf), hjust = 0, vjust = 1)
 
 ```
+
+## Dynamic environmental variables
+
+NDVI anomoly appears important. Greener final stopover sites allow earlier departure from West. *Now to test rainfall and greening in more detail*
 
@@
 
-newdat<-data.frame(aquaANOM=seq(-0.15,0.15, 0.005), DEPwestAF=1)
+newdat<-data.frame(aquaANOM=seq(-0.1,0.15, 0.005), DEPwestAF=1)
 newdat$p1<-predict(m6, newdata=newdat, re.form=~0)
@@
 ```
+
+Drought-breaking rains do not limit departure date specifically, however they do limit arrival into West Africa from Central Africa.
+Birds do not move into West Africa until rains have arrived.Blue blur is rainfall for each stopover Red line is arrival of cuckoo.
+
+```{r, warning=F, message=F, echo=F}
+
+dat2<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
+  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
+            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
+            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
+
+dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
+dat2[which(is.na(dat2$precip)),]$precip<-0 # fix na
+
+dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
+
+dat2<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
+
+# relevel to central well sampled year
+dat2$country<-factor(dat2$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
+
+ggplot()+geom_line(data=dat2[dat2$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(dat2 %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
+  ylab("Cumulative rainfall (mm)")
+
+```
+
+The first stopover in West Africa is timed to catch rains
+
+```{r, warning=F, message=F, echo=F}
+
+temp<-dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)
+
+qplot(data=temp, x=SO_startDOY, y=cumrf)+facet_wrap(~country)+theme_bw()+xlim(c(0,120))+xlab("Day of year")+ylab("Cumulative rainfall (mm)")
+```

### 2018-07-26 paper_wireframe_knit.Rmd
plotting rainfall and ndvi at time of arrival in WA and departure from CA

@@
 library('leaflet')
+library('gridExtra')
 
@@
 Drought-breaking rains do not limit departure date specifically, however they do limit arrival into West Africa from Central Africa.
-Birds do not move into West Africa until rains have arrived.Blue blur is rainfall for each stopover Red line is arrival of cuckoo.
+Birds do not move into West Africa until rains have arrived.
+Blue blur is rainfall for each stopover, black line is first arrival of cuckoo and red lines are subsequent stopovers.
 
@@
 
-ggplot()+geom_line(data=dat2[dat2$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(dat2 %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
+ggplot()+geom_line(data=dat2[dat2$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(dat2 %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+
+  geom_vline(data=(dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
   ylab("Cumulative rainfall (mm)")
+
+```
+
+Shown with mean rainfall per country (blue line) and mean arrival date for first stopovers in West Africa (red line), we can see rains arrive at generally the same time across  in Cote d'Ivoire than in Ghana and so are cuckoos?
+
+```{r, warning=F, message=F, echo=F}
+
+temp<-dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first) %>% group_by(country) %>% summarise(MSO_startDOY=mean(SO_startDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI))
+
+temp2<-dat2 %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
+
+p1<-ggplot()+
+geom_line(data=temp2,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
+geom_line(data=temp2,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
+geom_line(data=temp2,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
+geom_vline(data=(dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp, aes(xintercept = MSO_startDOY), colour='red', size=1.5)+
+   facet_grid(country~.)+xlab("Day of year")+
+  ylab("Cumulative rainfall (mm)")+theme_bw()
+
+p2<-ggplot()+geom_line(data=temp2,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
+geom_line(data=temp2,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
+geom_line(data=temp2,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
+geom_vline(data=(dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp, aes(xintercept = MSO_startDOY), colour='red', size=1.5)+
+   facet_grid(country~.)+xlab("Day of year")+
+  ylab("NDVI")+theme_bw()
+
+grid.arrange(p1, p2, ncol=2)
+
+```
+
+Compare this to when cuckoos leave their last stopover in central Africa
+
+```{r, warning=F, message=F, echo=F}
+
+dat3<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
+  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
+            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
+            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
+
+dat3[which(dat3$precip>500),]$precip<-0 # remove erroneous vals
+dat3[which(is.na(dat3$precip)),]$precip<-0 # fix na
+
+dat3<-dat3 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
+
+dat3<- dat3[which(dat3$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon')),]
+
+temp<-dat3[dat3$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(last) %>% group_by(country) %>% summarise(MSO_endDOY=mean(SO_endDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI, na.rm=T))
+
+temp2<-dat3 %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
+
+p1<-ggplot()+
+geom_line(data=temp2,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
+geom_line(data=temp2,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
+geom_line(data=temp2,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
+geom_vline(data=(dat3[dat3$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(last)), aes(xintercept = SO_endDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp, aes(xintercept = MSO_endDOY), colour='red', size=1.5)+
+   facet_grid(country~.)+xlab("Day of year")+
+  ylab("Cumulative rainfall (mm)")+theme_bw()
+
+p2<-ggplot()+geom_line(data=temp2,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
+geom_line(data=temp2,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
+geom_line(data=temp2,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
+geom_vline(data=(dat3[dat3$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(last)), aes(xintercept = SO_endDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp, aes(xintercept = MSO_endDOY), colour='red', size=1.5)+
+   facet_grid(country~.)+xlab("Day of year")+
+  ylab("NDVI")+theme_bw()
+
+grid.arrange(p1, p2, ncol=2)
 

### 2018-07-26 paper_wireframe_knit.Rmd
Shuffle env data read in to top and add new leaflet map

@@
 dat<-left_join(dat, (lc1 %>% group_by(ptt, year) %>% dplyr::summarise_all(last)), by=c('ptt', 'year'))
+
+#### AND THE ENV DATA #####
+
+env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+# summarise data for each stopover so that value per date is the mean
+# of the detail coords values
+
+dat2<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
+  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
+            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
+            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
+
+dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
+dat2[which(is.na(dat2$precip)),]$precip<-0 # fix na
+
+dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
+
+# rearrange
+
+dat2<- dat2 %>% select(year, ptt, SO_startDOY, SO_endDOY, country, everything())
+
+dat2.1<-subset(dat2, cuck_pres=='Y')
+
+dat2.2<-dat2.1 %>% group_by(year, ptt, SO_startDOY) %>%
+  summarise(cumrf=mean(cumrf, na.rm=T),
+            NDVI= mean(NDVI, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
+            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),terrANOM=mean(terrANOM, na.rm=T),tamRAIN=mean(tamRAIN, na.rm=T),tamRAINanom=mean(tamRAINanom, na.rm=T), SOlen=first(SO_endDOY)-first(SO_startDOY))
 
@@
 
+So what do these early leaving birds do in their runup to departing Cote d'Ivoire that is different to other birds?
+
+```{r, warning=F, message=F, echo=F}
+
+soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
+
+soversWA2<-left_join(soversWA1, dat2.2[,c(1,2, 3,7)], by=c('ptt', 'year', 'SO_startDOY'))
+
+forsp<-na.omit(soversWA2[,c(1,2,7,8,9,17,18,22,23)])
+
+forsp<-forsp %>% group_by(ptt, year) %>% mutate(nzz=n()) %>% filter(nzz>1) # drops Peckham in 2016 that cocks up linestring
+
+dat_sp<-st_as_sf(forsp, coords = c("SO_median_long","SO_median_lat"), crs="+proj=longlat +datum=WGS84")
+
+dat_ls<-dat_sp %>% group_by(ptt, year, DEPwestAF) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
+
+dat_ls
+
+d2<-dat_sp[dat_sp$DEPwestAF<101,]
+d3<-dat_sp[dat_sp$DEPwestAF>100,]
+
+m <- leaflet() %>%
+  addTiles() %>%
+
+  addPolylines(data=dat_ls[dat_ls$DEPwestAF<101,], weight=1, color='blue', group='Early departure') %>%
+
+    addPolylines(data=dat_ls[dat_ls$DEPwestAF>100,], weight=1, color='orange', group='Late departure') %>%
+
+  addCircleMarkers(data=d2, popup= paste("ptt:", d2$ptt, "<br>", "year:", d2$year,"<br>", "NDVIanom:", d2$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, group='Early departure') %>%
+
+    addCircleMarkers(data=d3, popup= paste("ptt:", d3$ptt, "<br>", "year:", d3$year,"<br>", "NDVIanom:", d3$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, group='Late departure') %>%
+
+  addLayersControl(overlayGroups = c('Early departure', 'Late departure'), options=layersControlOptions(collapsed=F))
+
+m
+
+```
+
 ## Habitat Variables
@@
 ```{r, warning=F, message=F, echo=F}
-
-env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', h=T)
-
-# summarise data for each stopover so that value per date is the mean
-# of the detail coords values
-
-dat2<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
-  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
-            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
-            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
-
-dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
-dat2[which(is.na(dat2$precip)),]$precip<-0 # fix na
-
-dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
-
-# rearrange
-
-dat2<- dat2 %>% select(year, ptt, SO_startDOY, SO_endDOY, country, everything())
-
-dat2.1<-subset(dat2, cuck_pres=='Y')
-
-dat2.2<-dat2.1 %>% group_by(year, ptt, SO_startDOY) %>%
-  summarise(cumrf=mean(cumrf, na.rm=T),
-            NDVI= mean(NDVI, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),terrANOM=mean(terrANOM, na.rm=T),tamRAIN=mean(tamRAIN, na.rm=T),tamRAINanom=mean(tamRAINanom, na.rm=T), SOlen=first(SO_endDOY)-first(SO_startDOY))
 
@@
 
-Compare this to when cuckoos leave their last stopover in central Africa
-
-```{r, warning=F, message=F, echo=F}
-
-dat3<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
-  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
-            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
-            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
-
-dat3[which(dat3$precip>500),]$precip<-0 # remove erroneous vals
-dat3[which(is.na(dat3$precip)),]$precip<-0 # fix na
-
-dat3<-dat3 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
-
-dat3<- dat3[which(dat3$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon')),]
-
-temp<-dat3[dat3$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(last) %>% group_by(country) %>% summarise(MSO_endDOY=mean(SO_endDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI, na.rm=T))
-
-temp2<-dat3 %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
-
-p1<-ggplot()+
-geom_line(data=temp2,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
-geom_line(data=temp2,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
-geom_line(data=temp2,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
-geom_vline(data=(dat3[dat3$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(last)), aes(xintercept = SO_endDOY), colour='black', alpha=0.5)+
-geom_vline(data=temp, aes(xintercept = MSO_endDOY), colour='red', size=1.5)+
-   facet_grid(country~.)+xlab("Day of year")+
-  ylab("Cumulative rainfall (mm)")+theme_bw()
-
-p2<-ggplot()+geom_line(data=temp2,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
-geom_line(data=temp2,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
-geom_line(data=temp2,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
-geom_vline(data=(dat3[dat3$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(last)), aes(xintercept = SO_endDOY), colour='black', alpha=0.5)+
-geom_vline(data=temp, aes(xintercept = MSO_endDOY), colour='red', size=1.5)+
-   facet_grid(country~.)+xlab("Day of year")+
-  ylab("NDVI")+theme_bw()
-
-grid.arrange(p1, p2, ncol=2)
-
-```
-
 The first stopover in West Africa is timed to catch rains

### 2018-07-27 rainfall_ndvi_extract.R
Code added to extract emodis NDVI and anomaly data from 250m rasters

@@ -364,7 +364,68 @@ for(i in unique(int2$year))
 
 write.csv(int2, 'data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', row.names = F, quote=F)
 
+#### Code to extract NDVI and NDVI anomoly from 250m res data
+# https://earlywarning.usgs.gov/fews/search/Africa
 
+int2<-read.csv('data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+# join lat long
+
+int2$emodisNDVI<-NA
+
+int2$emodisANOM<-NA
+
+# set up key to get data
+
+int2$key<-'01'
+int2[int2$variable %in% 11:20,]$key <-'02'
+int2[int2$variable %in% 21:30,]$key <-'03'
+int2[int2$variable %in% 31:40,]$key <-'04'
+int2[int2$variable %in% 41:50,]$key <-'05'
+int2[int2$variable %in% 51:60,]$key <-'06'
+int2[int2$variable %in% 61:70,]$key <-'07'
+int2[int2$variable %in% 71:80,]$key <-'08'
+int2[int2$variable %in% 81:90,]$key <-'09'
+int2[int2$variable %in% 91:100,]$key <-'10'
+int2[int2$variable %in% 101:110,]$key <-'11'
+int2[int2$variable %in% 111:120,]$key <-'12'
+int2[int2$variable %in% 121:130,]$key <-'13'
+
+
+for(i in unique(int2$year))
+{
+  for(j in unique(int2[int2$year==i,]$key))
+  {
+    ras_ndvi<-raster(paste('F:/NDVIsmooth/wa', 
+                         substr(i, 3,4), j,'.tif',sep=""))  
+    
+    ras_anom<-raster(paste('F:/NDVIanom/wa', 
+                           substr(i, 3,4), j,'stmdn.tif',sep=""))
+    
+    int2[int2$year==i & int2$key==j,]$emodisNDVI<-
+      extract(ras_ndvi, int2[int2$year==i & int2$key==j,8:9], na.rm=T,
+              buffer=1000, fun=function(x){mean(x[x<201], na.rm=T)}) # 1 km buffer cos rater is 250m= v fine res  
+    
+    int2[int2$year==i & int2$key==j,]$emodisanom<-
[…]

### 2018-07-27 ndvi_compile.R
Code to grab emodis and REF rainfall data from https://edcintl.cr.usgs.gov server

@@ -283,3 +283,33 @@ write.csv(out, 'C:/cuckoo_tracking/data/GRIMMS_0.25deg_NDVI_climatology.csv', qu
 # reduce size
 
 
+## code to grab modis ndvi anomaly data
+
+for(i in c(1201:1214, 1301:1314, 1401:1414, 1501:1514,
+           1601:1614, 1701:1714, 1801:1813))
+{
+  download.file(url=paste0('https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/fews/web/africa/west/dekadal/emodis/ndvi_c6/mediananomaly/downloads/dekadal/wa',
+                           i,'stmdn.zip'),
+                destfile = paste0('F:/NDVIanom/ndvianom', i, '.zip'))
+  
+  download.file(url=paste0('https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/fews/web/africa/west/dekadal/emodis/ndvi_c6/temporallysmoothedndvi/downloads/dekadal/wa',
+                           i,'.zip'),
+                destfile = paste0('F:/NDVIsmooth/ndvismooth', i, '.zip'))
+}
+
+
+# to grab rfe rainfall and long term mean data
+
+expg<-expand.grid(1801:1805, 1:3)
+idz<-paste0(expg[,1], expg[,2])
+for(j in idz)
+{
+  download.file(url=paste0('https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/fews/web/africa/west/dekadal/rfe/dekadrainfall/downloads/dekadal/w',
+                           j,'rb.zip'),
+                destfile = paste0('F:/RFErainfall/RFE', j, '.zip'))
+  
+  download.file(url=paste0('https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/fews/web/africa/west/dekadal/rfe/longtermanom/downloads/dekadal/wlm',
+                           substr(j, 3,5),'rb.zip'),
+                destfile = paste0('F:/RFEltm/RFEltm', substr(j, 3,5), '.zip'))
+  
+}

### 2018-07-27 static_extract.R
Added section to attribute detailed coords with env (precip and ndvi) data

@@ -114,3 +114,50 @@ dat$d_river_str2<-apply(rivers2, 2, min)
 
 write.csv(dat, "data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", quote=F, row.names=F)
 
+### code to attribute stopover bestofday locations with env data and depature form 
+### West Africa details for vis in GIS
+
+library(dplyr)
+
+soversWA<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", h=T)
+
+dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)
+
+# set NA or manually correct cuckoo depature dates that
+# are incorrect, from manual examination
+
+dat[dat$ptt==115586 & dat$year==2014,]$DEPwestAF<-106
+dat[dat$ptt==134955 & dat$year==2015,]$DEPwestAF<-NA
+dat[dat$ptt==134956 & dat$year==2015,]$DEPwestAF<-NA
+
+# and dead ones
+dat[dat$ptt==128296 & dat$year==2014,]$DEPwestAF<-NA
+
+env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_emodis_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+names(env)[names(env)=='value']<-'precip'
+names(env)[names(env)=='value2']<-'ndvi'
+
+soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
+
+soversWA1[,25:31]<-round(soversWA1[,25:31]/rowSums(soversWA1[,25:31], na.rm=T) *100)                      
+
+env[which(env$precip>500),]$precip<-0 # remove erroneous vals
+env[which(is.na(env$precip)),]$precip<-0 # fix na
+
+dat2<-env %>% group_by(year, ptt, timestamp) %>% dplyr::mutate(cumrf=cumsum(precip))
+
+dat2<-dplyr::arrange(dat2, ptt, year, timestamp, variable)
+dat2$tamRAIN[which(!(1:118625 %in% seq(5,118625, 5)))]<-0
[…]

### 2018-07-27 paper_wireframe_knit.Rmd
Changes to env data munipulation

@@
 
-env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_by_stopover_detailcoords_2018_dead.csv', h=T)
+env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_emodis_by_stopover_detailcoords_2018_dead.csv', h=T)
+
+names(env)[names(env)=='value']<-'precip'
+names(env)[names(env)=='value2']<-'ndvi'
+env$cuck_pres<-as.numeric(env$cuck_pres) # abs=1, pres=2
 
@@
 
-dat2<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
-  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
-            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
-            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
+dat2<- env[,c(2,3,4,10:22)] %>% group_by(year, ptt, SO_startDOY, variable) %>%
+  summarise_all(mean,na.rm=T)
 
@@
 
-# rearrange
+# for tamsat cumrf
+dat2<-dplyr::arrange(dat2, ptt, year, SO_startDOY, variable)
+dat2$tamRAIN[which(!(1:21375 %in% seq(5,21375, 5)))]<-0
+dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(tamcumrf=cumsum(tamRAIN))
 
-dat2<- dat2 %>% select(year, ptt, SO_startDOY, SO_endDOY, country, everything())
-
-dat2.1<-subset(dat2, cuck_pres=='Y')
+# select cond during SO..and befoe??
+dat2.1<-subset(dat2, cuck_pres==2)
 
 dat2.2<-dat2.1 %>% group_by(year, ptt, SO_startDOY) %>%
-  summarise(cumrf=mean(cumrf, na.rm=T),
-            NDVI= mean(NDVI, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T),terrANOM=mean(terrANOM, na.rm=T),tamRAIN=mean(tamRAIN, na.rm=T),tamRAINanom=mean(tamRAINanom, na.rm=T), SOlen=first(SO_endDOY)-first(SO_startDOY))
+  summarise_all(mean, na.rm=T)
 
@@
 
+#setwd('C:/cuckoo_tracking/data/spatial');st_write(dat_ls,'spring_linesWA.shp')
+
 d2<-dat_sp[dat_sp$DEPwestAF<101,]
@@
 dat2.3<-dat2.2 %>% group_by(year, ptt) %>%
-  summarise(cumrf=last(cumrf),
-            NDVI= last(NDVI), 
-            aquaNDVI=last(aquaNDVI),
-            aquaANOM=last(aquaANOM),
-            terrNDVI=last(terrNDVI),
-            terrANOM=last(terrANOM),
-             tamRAIN=last(tamRAIN),
-            tamRAINanom=last(tamRAINanom),
-            last_SOstart=last(SO_startDOY))
+  summarise_all(last)
 

### 2018-07-31 paper_wireframe_knit.Rmd
Fixes so code works with expanded analyses

@@
 
+temp<- env %>% group_by(year, ptt, SO_startDOY, variable) %>% summarise(country=first(country))
+
+dat2$country<-temp$country
+
 dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
@@
 dat2.2<-dat2.1 %>% group_by(year, ptt, SO_startDOY) %>%
-  summarise_all(mean, na.rm=T)
+  summarise_if(is.numeric, mean, na.rm=T)
 
@@
 So what do these early leaving birds do in their runup to departing Cote d'Ivoire that is different to other birds?
+The tracks of early departing birds are grouped with blue lines and late departing with orange lines.
+Each stopover is coloured either red (indicating below average NDVI) or green (above average NDVI) to see if early birds chain together 'greener stopovers.
 
@@
 
+dat$year<-as.integer(dat$year)
+
 soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
 
-soversWA2<-left_join(soversWA1, dat2.2[,c(1,2, 3,7)], by=c('ptt', 'year', 'SO_startDOY'))
+soversWA2<-left_join(soversWA1, dat2.2[,c(1,2, 3,9)], by=c('ptt', 'year', 'SO_startDOY'))
 
@@
 
-dat_ls
+#dat_ls
 
@@
 
-  addCircleMarkers(data=d2, popup= paste("ptt:", d2$ptt, "<br>", "year:", d2$year,"<br>", "NDVIanom:", d2$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, group='Early departure') %>%
+  addCircleMarkers(data=d2, popup= paste("ptt:", d2$ptt, "<br>", "year:", d2$year,"<br>", "NDVIanom:", d2$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, opacity=0.2, group='Early departure') %>%
 
-    addCircleMarkers(data=d3, popup= paste("ptt:", d3$ptt, "<br>", "year:", d3$year,"<br>", "NDVIanom:", d3$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, group='Late departure') %>%
+    addCircleMarkers(data=d3, popup= paste("ptt:", d3$ptt, "<br>", "year:", d3$year,"<br>", "NDVIanom:", d3$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5,opacity=0.2, group='Late departure') %>%
 
@@
 
-dat2<- env %>% group_by(year, ptt, SO_startDOY, variable) %>%
-  summarise(country=last(country), SO_endDOY=first(SO_endDOY),
-            cuck_pres=first(cuck_pres), precip=mean(value, na.rm=T),
-            NDVI= mean(value2, na.rm=T), aquaNDVI=mean(aquaNDVI, na.rm=T),
-            aquaANOM=mean(aquaANOM, na.rm=T),terrNDVI=mean(terrNDVI, na.rm=T), terrANOM=mean(terrANOM, na.rm=T), tamRAIN=mean(tamRAIN, na.rm=T), tamRAINanom=mean(tamRAINanom, na.rm=T) )
+temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
 
-dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
-dat2[which(is.na(dat2$precip)),]$precip<-0 # fix na
+temp$country<-factor(temp$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
 
-dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))
-
-dat2<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
-
-# relevel to central well sampled year
-dat2$country<-factor(dat2$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
-
-ggplot()+geom_line(data=dat2[dat2$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(dat2 %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+
-  geom_vline(data=(dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
+ggplot()+geom_line(data=temp[temp$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(temp %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+
+  geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
   ylab("Cumulative rainfall (mm)")
@@
 
-Shown with mean rainfall per country (blue line) and mean arrival date for first stopovers in West Africa (red line), we can see rains arrive at generally the same time across in Cote d'Ivoire than in Ghana and so are cuckoos?
+Shown with mean rainfall (blue line) and NDVI (green line) per country and mean arrival date for first stopovers in West Africa (red line), we can see rains arrive at generally the same time in Cote d'Ivoire and Ghana and so do cuckoos
 
-```{r, warning=F, message=F, echo=F}
+```{r, warning=F, message=F, echo=F, fig.height=12, fig.width=6}
 
-temp<-dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first) %>% group_by(country) %>% summarise(MSO_startDOY=mean(SO_startDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI))
+temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
 
-temp2<-dat2 %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
+temp$country<-factor(temp$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
+
+temp2<-temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first) %>% group_by(country) %>% summarise(MSO_startDOY=mean(SO_startDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI))
+
+temp3<-temp %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
 
 p1<-ggplot()+
-geom_line(data=temp2,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
-geom_line(data=temp2,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
-geom_line(data=temp2,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
-geom_vline(data=(dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
-geom_vline(data=temp, aes(xintercept = MSO_startDOY), colour='red', size=1.5)+
+geom_line(data=temp3,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
+geom_line(data=temp3,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
+geom_line(data=temp3,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
+geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp2, aes(xintercept = MSO_startDOY), colour='red', size=1)+
    facet_grid(country~.)+xlab("Day of year")+
@@
 
-p2<-ggplot()+geom_line(data=temp2,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
-geom_line(data=temp2,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
-geom_line(data=temp2,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
-geom_vline(data=(dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
-geom_vline(data=temp, aes(xintercept = MSO_startDOY), colour='red', size=1.5)+
+p2<-ggplot()+geom_line(data=temp3,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
+geom_line(data=temp3,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
+geom_line(data=temp3,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
+geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp2, aes(xintercept = MSO_startDOY), colour='red', size=1)+
    facet_grid(country~.)+xlab("Day of year")+
@@
 
-temp<-dat2[dat2$cuck_pres=='Y',] %>% group_by(ptt, year) %>% summarise_all(first)
+temp<-dat2[dat2$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)
 

### 2018-08-03 ndvi_rainfall_playground.R
Working through ideas for final model to determine departure date from West Africa

@@ -1,0 +1,215 @@
+### 31/07/18
+
+### Area to explore and develop env (NDVI and rainfall) metrics
+### to test against various responses relating to movement withing/ from
+### West Africa
+
+library(dplyr)
+library(ggplot2)
+
+if(Sys.info()['nodename']=="D9L5812"){
+  setwd("C:/cuckoo_tracking")}else{
+    setwd("N:/cuckoo_tracking")}
+
+#summary data
+dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)
+
+#stopover coord data (median points)
+sovers<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_BOTH.csv', h=T)
+
+#extra data for filling migratory route
+
+extraz<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv', h=T)
+
+#### Data cleaning
+
+# set NA or manually correct cuckoo depature dates that
+# are incorrect, from manual examination
+
+dat[dat$ptt==115586 & dat$year==2014,]$DEPwestAF<-106
+dat[dat$ptt==134955 & dat$year==2015,]$DEPwestAF<-NA
+dat[dat$ptt==134956 & dat$year==2015,]$DEPwestAF<-NA
+
+# and dead ones
+dat[dat$ptt==128296 & dat$year==2014,]$DEPwestAF<-NA
+
+## fill NA in mig route
+# fills some
+dat<-left_join(dat, select(extraz, ptt, year, autumn_mig), by=c('ptt', 'year'))
+
+# fill 2018 birds
+dat[c(47, 44, 42),]$autumn_mig.y='SW'
+dat[39,]$autumn_mig.y='SE'
+# and dead birds
+dat[c(2,20,45),]$autumn_mig.y='SW'
+dat[c(1,3,23,25,31),]$autumn_mig.y='SE'
+
+names(dat)[names(dat)=='autumn_mig.y']<-'autumn_mig'
+dat$autumn_mig.x<-NULL
+
+### Add departure info from WA e.g. country, lat and long, length of final stopover and 2nd before last
+
+notWA<-c('Italy', 'France', 'Spain', 'Algeria',
+         'United Kingdom', 'Morocco', 'Portugal')
+
+soversWA<-sovers[-which(sovers$country %in% notWA),]
+
[…]

### 2018-08-24 paper_wireframe_knit.Rmd
Added section with rainfall/ndvi curves and dep~waarr plots. Still has some model stuff below that will be removed/corrected

@@
 library('gridExtra')
+library('MuMIn')
 
@@
   summarise_if(is.numeric, mean, na.rm=T)
+
+# read in model-ready dataset
+
+dmod<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_attrib_modelready.csv', h=T)
+
+dmod$tWA=dmod$DEPwestAF-dmod$departureGarr
+
+dmod<-dmod[,c(1,2,6,22,27,28,29,37,39,52,57,63,74,85,87,92, 115)]
+
+dmod<-na.omit(dmod)
 
@@
 
+Taking a step back, it is important to if there is any environmental
+control over how birds arrive into West Africa.
+
+Drought-breaking rains do not limit departure date specifically, however they do limit arrival into West Africa from Central Africa.
+Birds do not move into West Africa until rains have arrived.
+Blue blur is rainfall for each stopover, black line is first arrival of cuckoo and red lines are subsequent stopovers.
+
+```{r, warning=F, message=F, echo=F}
+
+temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
+
+temp$country<-factor(temp$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
+
+ggplot()+geom_line(data=temp[temp$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(temp %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+
+  geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
+  ylab("Cumulative rainfall (mm)")
+
+```
+
+Shown with mean rainfall (blue line) and NDVI (green line) per country and mean arrival date for first stopovers in West Africa (red line), we can see rains arrive at generally the same time in Cote d'Ivoire and Ghana and so do cuckoos
+
+```{r, warning=F, message=F, echo=F, fig.height=12, fig.width=6}
+
+temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
+
+temp$country<-factor(temp$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
+
+temp2<-temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first) %>% group_by(country) %>% summarise(MSO_startDOY=mean(SO_startDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI))
+
+temp3<-temp %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
+
+p1<-ggplot()+
+geom_line(data=temp3,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
+geom_line(data=temp3,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
+geom_line(data=temp3,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
+geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp2, aes(xintercept = MSO_startDOY), colour='red', size=1)+
+   facet_grid(country~.)+xlab("Day of year")+
+  ylab("Cumulative rainfall (mm)")+theme_bw()
+
+p2<-ggplot()+geom_line(data=temp3,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
+geom_line(data=temp3,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
+geom_line(data=temp3,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
+geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
+geom_vline(data=temp2, aes(xintercept = MSO_startDOY), colour='red', size=1)+
+   facet_grid(country~.)+xlab("Day of year")+
+  ylab("NDVI")+theme_bw()
+
+grid.arrange(p1, p2, ncol=2)
+
+```
+
+Despite rain generally arriving in West Africa at the same time, there
+is variance of ~ 60 days in the times in which cuckoos arrive into
+West Africa (with the bulk of birds arriving between day 55 and day 95)
+
+There is however, little relationship between arrival date into West Africa
+and departure date from West Africa
+
+```{r, warning=F, message=F, echo=F}
+
+m1<-lmer(DEPwestAF~departureGarr+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF%in%85:120,])
+
+ggplot(data=dmod, aes(x=departureGarr, y=DEPwestAF ))+xlab("Arrival into West Africa (DOY)")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=1.83, p=0.185 *3 outliers removed*', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+
+#Anova(m1, test.statistic = 'F')
+
+```
+
+Note: if the above model includes the 3 outliers then a there is a significant relationship... * need to decide which is more biologically plausible*
+
+```{r, warning=F, message=F, echo=F}
+m1<-lmer(DEPwestAF~departureGarr+(1|ptt)+(1|year), data=dmod)
+#Anova(m1, test.statistic = 'F')
+
+newdat<-data.frame(DEPwestAF=1, departureGarr=25:100)
+newdat$p1<-predict(m1, newdata=newdat, re.form=~0)
+
+predmat<-model.matrix(DEPwestAF~departureGarr, data=newdat)
+vcv<-vcov(m1)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+
+qplot(data=dmod, y=DEPwestAF, x=departureGarr)+
+  geom_line(data=newdat, aes(x=departureGarr, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=departureGarr, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=departureGarr, y=uc), colour='red', linetype='dashed')+
+  ylab("Days spent in West Africa")+
+  xlab("Arrival into West Africa (DOY)")+
+  theme_bw()+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=10.47, p=0.0027', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+Because almost all birds depart West Africa between day 90 and day 110,
+later arriving birds spend significantly shorter time in West Africa
+
+```{r, warning=F, message=F, echo=F}
+m1<-lmer(tWA~departureGarr+(1|year)+(1|ptt), data=dmod)
+
+#Anova(m1, test.statistic = 'F')
+
+newdat<-data.frame(tWA=1, departureGarr=25:100)
+newdat$p1<-predict(m1, newdata=newdat, re.form=~0)
+
+predmat<-model.matrix(tWA~departureGarr, data=newdat)
+vcv<-vcov(m1)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+
+qplot(data=dmod, y=tWA, x=departureGarr)+
+  geom_line(data=newdat, aes(x=departureGarr, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=departureGarr, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=departureGarr, y=uc), colour='red', linetype='dashed')+
+  ylab("Days spent in West Africa")+
+  xlab("Arrival into West Africa (DOY)")+
+  theme_bw()+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=63.35, p=3.19e-09', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+When graphed with average rainfall and ndvi curved (across all WA stopovers) we see that early arriving birds into West Africa spend a long time there when greening is yet to/just starting whereas late arriving birds spend a short time at near-peak greenness.
+Interestingly,
+the mean arrival date into west africa (day 71; black line) corresponds with earlier departures, which could be facilitated by concurrent peak greening and rainfall rates.
+
+```{r, warning=F, message=F, echo=F}
+temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
+
+temp3<-temp %>% group_by(variable) %>%
+  summarise(Mcumrf=mean(tamcumrf), rfmin=mean(tamcumrf)-sd(tamcumrf), rfmax=mean(tamcumrf)+sd(tamcumrf),
+   MaquaNDVI=mean(emodisNDVI, na.rm=T), ndmin=(mean(emodisNDVI, na.rm=T)-sd(emodisNDVI, na.rm=T)),
+   ndmax=(mean(emodisNDVI, na.rm=T)+sd(emodisNDVI, na.rm=T)))
+
+#lets make variable!
+
+ggplot(data=dmod, aes(x=departureGarr, y=DEPwestAF))+
+  geom_line(data=temp3,aes(x=variable, y=Mcumrf/2), size=1, colour='blue')+
+  geom_line(data=temp3,aes(x=variable, y=rfmin/2), colour='grey', linetype=2)+
+  geom_line(data=temp3,aes(x=variable, y=rfmax/2), colour='grey', linetype=2)+ 
+  geom_line(data=temp3,aes(x=variable, y=MaquaNDVI*100), size=1, colour='dark green')+
+  geom_line(data=temp3,aes(x=variable, y=ndmin*100), colour='orange', linetype=2)+
+  geom_line(data=temp3,aes(x=variable, y=ndmax*100), colour='orange', linetype=2)+
+  geom_point(aes(colour=tWA))+scale_y_continuous(breaks=seq(0, 130, 10), limits=c(0,130))+
+  scale_x_continuous(breaks=seq(0, 130, 10))+
+  scale_colour_gradientn(colours=topo.colors(8))+
+  theme_bw()+labs(y="Departure from West Africa (DOY)", x="Arrival into West Africa (DOY)", colour='Days in WA')+geom_vline(aes(xintercept=71), colour='black')
+```
+
+Visualised another way with time coloured gradient.
+
+```{r, warning=F, message=F, echo=F} 
+df <- expand.grid(departureGarr=0:130,
+                  DEPwestAF=75:130)
+
+df$filly=df$DEPwestAF-df$departureGarr
+
+df[df$filly<0 | df$filly>80,]$filly<-NA
+
+ggplot() + 
+  geom_tile(data=df, aes(x=departureGarr, y=DEPwestAF, fill=filly),alpha = 0.75) +      
+  scale_fill_gradient(low='#ffeda0', high='#f03b20', na.value=NA)+
+  geom_line(data=temp3,aes(x=variable, y=Mcumrf/2), size=1, colour='blue')+
+  geom_line(data=temp3,aes(x=variable, y=rfmin/2), colour='grey', linetype=2)+
+  geom_line(data=temp3,aes(x=variable, y=rfmax/2), colour='grey', linetype=2)+ 
+  geom_line(data=temp3,aes(x=variable, y=MaquaNDVI*100), size=1, colour='dark green')+
+  geom_line(data=temp3,aes(x=variable, y=ndmin*100), colour='orange', linetype=2)+
+  geom_line(data=temp3,aes(x=variable, y=ndmax*100), colour='orange', linetype=2)+
+  geom_point(data=dmod, aes(x=departureGarr, y=DEPwestAF), shape=1)+
+  theme_bw()+scale_y_continuous(breaks=seq(0, 130, 10), limits=c(0,130))+
+  scale_x_continuous(breaks=seq(0, 130, 10))+
+  labs(y="Departure from West Africa (DOY)", x="Arrival into West Africa (DOY)", fill='Days in WA')+geom_vline(aes(xintercept=71), colour='black')
+
+```
+
+OK so are there any explanatory variables that can improve a model
+
+```{r, warning=F, message=F, echo=F}
+
+# make two gradient metrics
+
+dmod$ndvi_last_ang<-dmod$ndviTdiff_last*dmod$ndvidiff_last
+dmod$rain_first_ang<-(dmod$monsTdiff_first+25)*dmod$raindiff2_first
+
+m0<-lmer(DEPwestAF~(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m1<-lmer(DEPwestAF~dep_country+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m2<-lmer(DEPwestAF~firstCdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m3<-lmer(DEPwestAF~departureGarr+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m4<-lmer(DEPwestAF~nSOdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m5<-lmer(DEPwestAF~monsTdiff_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m6<-lmer(DEPwestAF~raindiff2_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m7<-lmer(DEPwestAF~emodisNDVI_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m8<-lmer(DEPwestAF~ndviTdiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m9<-lmer(DEPwestAF~ndvidiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m10<-lmer(DEPwestAF~emodisANOM_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m11<-lmer(DEPwestAF~Trees_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m12<-lmer(DEPwestAF~Croplands_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m13<-lmer(DEPwestAF~d_river_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+
+m14<-lmer(DEPwestAF~rain_first_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+
+m15<-lmer(DEPwestAF~ndvi_last_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+
+anova(m0,m1,m2,m3,m4,m5,m6,m7,m8,m9,m10, m11, m12,m13, m14, m15)
+
+moddat<-dmod[dmod$DEPwestAF %in% 85:120,]
+
+mfin<-lmer(DEPwestAF~departureGarr+nSOdepartureG+monsTdiff_first*raindiff2_first+emodisNDVI_last+ndviTdiff_last*ndvidiff_last+emodisANOM_mean+Trees_mean+Croplands_mean+d_river_mean+(1|ptt)+(1|year), data=moddat,  na.action = na.fail)
+
+mdr<-dredge(mfin)
+
+get.models(mdr, subset = delta < 4, method = "REML")
+
+#qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=nSOdepartureG)
+
+#qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=ndviTdiff_last)
+
+#qplot(data=dmod[dmod$DEPwestAF %in% 85:120 & dmod$emodisANOM_mean>-0.05,], y=DEPwestAF, x=emodisANOM_mean)
+```
+
 So what do these early leaving birds do in their runup to departing Cote d'Ivoire that is different to other birds?
@@
 
-The percentage of grass cover does appear to affect departure date. * Not sure how believeable this is*
+The percentage of grass cover does appear to affect departure date. *Not sure how believeable this is*
 
@@
 ```
-
-Drought-breaking rains do not limit departure date specifically, however they do limit arrival into West Africa from Central Africa.
-Birds do not move into West Africa until rains have arrived.
-Blue blur is rainfall for each stopover, black line is first arrival of cuckoo and red lines are subsequent stopovers.
-
-```{r, warning=F, message=F, echo=F}
-
-temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
-
-temp$country<-factor(temp$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
-
-ggplot()+geom_line(data=temp[temp$cumrf>0,],aes(x=variable, y=cumrf, group=paste(ptt, year,SO_startDOY)), size=2, alpha=0.08, colour='blue')+theme_bw()+geom_vline(data=(temp %>% group_by(ptt, year, SO_startDOY) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='red', alpha=0.5)+
-  geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+facet_wrap(~country)+xlab("Day of year")+
-  ylab("Cumulative rainfall (mm)")
-
-```
-
-Shown with mean rainfall (blue line) and NDVI (green line) per country and mean arrival date for first stopovers in West Africa (red line), we can see rains arrive at generally the same time in Cote d'Ivoire and Ghana and so do cuckoos
-
-```{r, warning=F, message=F, echo=F, fig.height=12, fig.width=6}
-
-temp<- dat2[- which(dat2$country %in% c('Angola', 'Central African Republic','Congo', 'Democratic Republic of the Congo', 'Gabon',  'Cameroon', 'Burkina Faso', 'Liberia')),]
-
-temp$country<-factor(temp$country, levels=c('Nigeria', 'Togo', 'Ghana', "Cote d'Ivoire", 'Guinea', 'Sierra Leone'))
-
-temp2<-temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first) %>% group_by(country) %>% summarise(MSO_startDOY=mean(SO_startDOY), Mrf=mean(cumrf), Mnd=mean(aquaNDVI))
-
-temp3<-temp %>% group_by(country, variable) %>% summarise(Mcumrf=mean(cumrf), rfmin=mean(cumrf)-sd(cumrf), rfmax=mean(cumrf)+sd(cumrf),                                                          MaquaNDVI=mean(aquaNDVI, na.rm=T), ndmin=(mean(aquaNDVI, na.rm=T)-sd(aquaNDVI, na.rm=T)), ndmax=(mean(aquaNDVI, na.rm=T)+sd(aquaNDVI, na.rm=T)))
-
-p1<-ggplot()+
-geom_line(data=temp3,aes(x=variable, y=Mcumrf), size=1, colour='blue')+
-geom_line(data=temp3,aes(x=variable, y=rfmin), colour='grey', linetype=2)+
-geom_line(data=temp3,aes(x=variable, y=rfmax), colour='grey', linetype=2)+  
-geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
-geom_vline(data=temp2, aes(xintercept = MSO_startDOY), colour='red', size=1)+
-   facet_grid(country~.)+xlab("Day of year")+
-  ylab("Cumulative rainfall (mm)")+theme_bw()
-
-p2<-ggplot()+geom_line(data=temp3,aes(x=variable, y=MaquaNDVI), size=1, colour='dark green')+
-geom_line(data=temp3,aes(x=variable, y=ndmin), colour='orange', linetype=2)+
-geom_line(data=temp3,aes(x=variable, y=ndmax), colour='orange', linetype=2)+
-geom_vline(data=(temp[temp$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)), aes(xintercept = SO_startDOY), colour='black', alpha=0.5)+
-geom_vline(data=temp2, aes(xintercept = MSO_startDOY), colour='red', size=1)+
-   facet_grid(country~.)+xlab("Day of year")+
-  ylab("NDVI")+theme_bw()
-
-grid.arrange(p1, p2, ncol=2)
-
-```
-
-The first stopover in West Africa is timed to catch rains
-
-```{r, warning=F, message=F, echo=F}
-
-temp<-dat2[dat2$cuck_pres==2,] %>% group_by(ptt, year) %>% summarise_all(first)
-
-qplot(data=temp, x=SO_startDOY, y=cumrf)+facet_wrap(~country)+theme_bw()+xlim(c(0,120))+xlab("Day of year")+ylab("Cumulative rainfall (mm)")
-```

### 2018-08-30 cuckoo_demography.R
First extract of cuckoo dmeography data from Bird Atlas for tagging sites

@@ -1,0 +1,79 @@
+## 29/08/2018
+
+## script to buffer capture locations per population and then
+## extract cuckoo population chnage data from Bird Atlas 
+
+library(dplyr)
+library(sf)
+library(sp)
+library(tmap)
+library(rgdal)
+
+## get best of day locations
+
+bod<-read.csv('C:/cuckoo_tracking/data/movebank_cuckoos_hybridfilter_bestofday_2018_clean.csv', h=T)
+
+bod<-bod %>% group_by(ptt) %>%summarise_all(first)
+
+#split scotland into 2 sites
+
+bod[bod$capture.location=='Scotland',]
+
+bod$capture.location<-as.character(bod$capture.location)
+
+bod[bod$location.lat>57,]$capture.location<-'Kinloch_Skye'
+
+bodsp<-SpatialPoints(cbind(bod$location.long,
+                    bod$location.lat), 
+              CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
+
+bodspdf<-SpatialPointsDataFrame(bodsp, data=bod)
+
+tmap_mode("view")
+qtm(bodspdf)
+
+# to british national grid proj
+bodproj<-spTransform(bodspdf, 
+                     CRS('+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +towgs84=446.448,-125.157,542.06,0.15,0.247,0.842,-20.489 +units=m +no_defs'))
+bodsf<-st_as_sf(bodproj)
+
+# read in Bird Atlas data
+
+batl<-read.csv('C:/cuckoo_tracking/sourced_data/cuckoo_demography/cuckoo Atlas abundance change.csv', h=T)
+
+batlsp<-SpatialPoints(cbind(batl$easting,
+                            batl$northing), 
+                     CRS("+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +towgs84=446.448,-125.157,542.06,0.15,0.247,0.842,-20.489 +units=m +no_defs"))
+
+batlspdf<-SpatialPointsDataFrame(batlsp, data=batl)
+
+batlsf<-st_as_sf(batlspdf)
+
+#https://stackoverflow.com/questions/49200458/find-nearest-features-using-sf-in-r
[…]

### 2018-08-30 west_africa_departure_model.R
Messy thoughts on best way to proceed with analyses. Will need tidying later.

@@ -194,26 +194,38 @@ temp3<-temp %>% group_by(variable) %>%
    MaquaNDVI=mean(emodisNDVI, na.rm=T), ndmin=(mean(emodisNDVI, na.rm=T)-sd(emodisNDVI, na.rm=T)),
    ndmax=(mean(emodisNDVI, na.rm=T)+sd(emodisNDVI, na.rm=T)))
 
-ggplot(data=dmod, aes(x=departureGarr, y=DEPwestAF, colour=DEPwestAF-departureGarr))+
+#lets make variable!
+
+dmod$tWA=dmod$DEPwestAF-dmod$departureGarr
+
+ggplot(data=dmod, aes(x=departureGarr, y=DEPwestAF, colour=tWA))+
   geom_line(data=temp3,aes(x=variable, y=Mcumrf/2), size=1, colour='blue')+
   geom_line(data=temp3,aes(x=variable, y=rfmin/2), colour='grey', linetype=2)+
   geom_line(data=temp3,aes(x=variable, y=rfmax/2), colour='grey', linetype=2)+ 
   geom_line(data=temp3,aes(x=variable, y=MaquaNDVI*100), size=1, colour='dark green')+
   geom_line(data=temp3,aes(x=variable, y=ndmin*100), colour='orange', linetype=2)+
   geom_line(data=temp3,aes(x=variable, y=ndmax*100), colour='orange', linetype=2)+
-  geom_point()+
+  geom_point()+scale_y_continuous(breaks=seq(0, 150, 10))+
+  scale_x_continuous(breaks=seq(0, 120, 10))+
   scale_colour_gradientn(colours=terrain.colors(10))+
   theme_bw()
 
 #colour graded background:
 #https://stackoverflow.com/questions/39466505/diagonal-grading-of-background-color-of-ggplot-graph-in-r
 
-df <- expand.grid(departureGarr=25:100,
+df <- expand.grid(departureGarr=0:130,
                   DEPwestAF=75:130)
 
 df$filly=df$DEPwestAF-df$departureGarr 
 
 df[df$filly<0 | df$filly>80,]$filly<-NA
+
+# add in m.arr line from code below
+m.arr<-lmer(tWA~departureGarr+(1|ptt)+(1|year), data=dmod)
+
+tempdf<-data.frame(departureGarr=0:130)
+tempdf$predDEPWA<-tempdf$departureGarr+
+  predict(m.arr, newdata=tempdf, re.form=~0)
 
 ggplot() + 
[…]

### 2018-08-30 stopover_rainfall_ndvi_detailcoords_vis_DEPWAbirds.Rmd
Minor change to specify just birds that we are using in DEPwestAF analyses

---
title: "West Africa cuckoo stopover rainfall patterns"
author: "Mark Miller"
date: "31 July 2018"
output: html_document
---

```{r cuckoos, echo=FALSE, message=F, warning=F, fig.height=9}

#### work around for different laptop/desktop directories ####
#### !!!! RUN THIS !!!! ####
 if(Sys.info()['nodename']=="D9L5812"){
   setwd("C:/cuckoo_tracking")}else{
    setwd("N:/cuckoo_tracking")}

library(ggplot2)
library(dplyr)
#summary data
dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)

# set NA or manually correct cuckoo depature dates that
# are incorrect, from manual examination

dat[dat$ptt==115586 & dat$year==2014,]$DEPwestAF<-106
dat[dat$ptt==134955 & dat$year==2015,]$DEPwestAF<-NA
dat[dat$ptt==134956 & dat$year==2015,]$DEPwestAF<-NA

# and dead ones
dat[dat$ptt==128296 & dat$year==2014,]$DEPwestAF<-NA

env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_emodis_by_stopover_detailcoords_2018_dead.csv', h=T)

names(env)[names(env)=='value']<-'precip'
names(env)[names(env)=='value2']<-'ndvi'
env$cuck_pres<-as.numeric(env$cuck_pres) # abs=1, pres=2

dat<-left_join(dat, env %>% group_by(ptt, year) %>% arrange(SO_startDOY, variable, .by_group=T) %>%
                 summarise(dep_country=last(country)),
               by=c('ptt', 'year'))

# can test individual countries
#dat<-dat[dat$dep_country=="Cote d'Ivoire",]
#dat<-dat[dat$dep_country=="Ghana",]

# summarise data for each stopover so that value per date is the mean
# of the detail coords values

dat2<- env[,c(2,3,4,5,10:22)] %>% group_by(year, ptt, SO_startDOY, variable) %>%
  summarise_all(mean,na.rm=T)

temp<- env %>% group_by(year, ptt, SO_startDOY, variable) %>% summarise(country=first(country))

dat2$country<-temp$country

dat2[which(dat2$precip>500),]$precip<-0 # remove erroneous vals
dat2[which(is.na(dat2$precip)),]$precip<-0 # fix na

dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(cumrf=cumsum(precip))

# for tamsat cumrf
dat2<-dplyr::arrange(dat2, ptt, year, SO_startDOY, variable)
dat2$tamRAIN[which(!(1:21375 %in% seq(5,21375, 5)))]<-0
dat2<-dat2 %>% group_by(year, ptt, SO_startDOY) %>% dplyr::mutate(tamcumrf=cumsum(tamRAIN))

### visualisastion

for(i in 1:nrow(dat))
{
  dtemp<-dat[i,]

  if(is.na(dtemp$DEPwestAF)){next}

t2<-dat2[dat2$ptt==dtemp$ptt & dat2$year==dtemp$year,]

temp=ggplot(data=t2)+
geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=500), fill='grey')+

  geom_line( aes(x=variable, y=emodisNDVI*500), colour='dark green')+

    geom_line( aes(x=variable, y=emodisANOM*1000+200), colour='purple', linetype=2)+
  geom_line( aes(x=variable, y=tamRAINanom+200), colour='dark blue', linetype=2)+
  geom_hline(yintercept=200)+
    geom_line(aes(x=variable, y=cumrf), col='blue', size=1)+
  geom_line(aes(x=variable, y=tamcumrf), col='cyan', size=1)+
  facet_wrap(~SO_startDOY+country)+theme_classic()+
  ylab('cumulative rainfall since 01 Jan (mm)')+ xlab('Day of year')+
          ggtitle(paste("PTT:", dtemp$ptt, "Year:",dtemp$year, "Dep WA:", dtemp$DEPwestAF))+ylim(c(0,500))

# for lite version without:
# geom_line( aes(x=variable, y=aquaANOM*1000+200), colour='pink', linetype=2)+
#  geom_line( aes(x=variable, y=aquaNDVI*500), colour='green')+
#
#

print(temp)

}

```

### 2018-08-30 paper_wireframe_knit.Rmd
Removal of final section testing env variables as need to re-think approach.

@@
 ```
-
-OK so are there any explanatory variables that can improve a model
-
-```{r, warning=F, message=F, echo=F}
-
-# make two gradient metrics
-
-dmod$ndvi_last_ang<-dmod$ndviTdiff_last*dmod$ndvidiff_last
-dmod$rain_first_ang<-(dmod$monsTdiff_first+25)*dmod$raindiff2_first
-
-m0<-lmer(DEPwestAF~(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m1<-lmer(DEPwestAF~dep_country+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m2<-lmer(DEPwestAF~firstCdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m3<-lmer(DEPwestAF~departureGarr+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m4<-lmer(DEPwestAF~nSOdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m5<-lmer(DEPwestAF~monsTdiff_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m6<-lmer(DEPwestAF~raindiff2_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m7<-lmer(DEPwestAF~emodisNDVI_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m8<-lmer(DEPwestAF~ndviTdiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m9<-lmer(DEPwestAF~ndvidiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m10<-lmer(DEPwestAF~emodisANOM_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m11<-lmer(DEPwestAF~Trees_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m12<-lmer(DEPwestAF~Croplands_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m13<-lmer(DEPwestAF~d_river_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-
-m14<-lmer(DEPwestAF~rain_first_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-
-m15<-lmer(DEPwestAF~ndvi_last_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-
-anova(m0,m1,m2,m3,m4,m5,m6,m7,m8,m9,m10, m11, m12,m13, m14, m15)
-
-moddat<-dmod[dmod$DEPwestAF %in% 85:120,]
-
-mfin<-lmer(DEPwestAF~departureGarr+nSOdepartureG+monsTdiff_first*raindiff2_first+emodisNDVI_last+ndviTdiff_last*ndvidiff_last+emodisANOM_mean+Trees_mean+Croplands_mean+d_river_mean+(1|ptt)+(1|year), data=moddat,  na.action = na.fail)
-
-mdr<-dredge(mfin)
-
-get.models(mdr, subset = delta < 4, method = "REML")
-
-#qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=nSOdepartureG)
-
-#qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=ndviTdiff_last)
-
-#qplot(data=dmod[dmod$DEPwestAF %in% 85:120 & dmod$emodisANOM_mean>-0.05,], y=DEPwestAF, x=emodisANOM_mean)
-```
-
-So what do these early leaving birds do in their runup to departing Cote d'Ivoire that is different to other birds?
-The tracks of early departing birds are grouped with blue lines and late departing with orange lines.
-Each stopover is coloured either red (indicating below average NDVI) or green (above average NDVI) to see if early birds chain together 'greener stopovers.
-
-```{r, warning=F, message=F, echo=F}
-
-dat$year<-as.integer(dat$year)
-
-soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
-
-soversWA2<-left_join(soversWA1, dat2.2[,c(1,2, 3,9)], by=c('ptt', 'year', 'SO_startDOY'))
-
-forsp<-na.omit(soversWA2[,c(1,2,7,8,9,17,18,22,23)])
-
-forsp<-forsp %>% group_by(ptt, year) %>% mutate(nzz=n()) %>% filter(nzz>1) # drops Peckham in 2016 that cocks up linestring
-
-dat_sp<-st_as_sf(forsp, coords = c("SO_median_long","SO_median_lat"), crs="+proj=longlat +datum=WGS84")
-
-dat_ls<-dat_sp %>% group_by(ptt, year, DEPwestAF) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
-
-#dat_ls
-
-#setwd('C:/cuckoo_tracking/data/spatial');st_write(dat_ls,'spring_linesWA.shp')
-
-d2<-dat_sp[dat_sp$DEPwestAF<101,]
-d3<-dat_sp[dat_sp$DEPwestAF>100,]
-
-m <- leaflet() %>%
-  addTiles() %>%
-
-  addPolylines(data=dat_ls[dat_ls$DEPwestAF<101,], weight=1, color='blue', group='Early departure') %>%
-
-    addPolylines(data=dat_ls[dat_ls$DEPwestAF>100,], weight=1, color='orange', group='Late departure') %>%
-
-  addCircleMarkers(data=d2, popup= paste("ptt:", d2$ptt, "<br>", "year:", d2$year,"<br>", "NDVIanom:", d2$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, opacity=0.2, group='Early departure') %>%
-
-    addCircleMarkers(data=d3, popup= paste("ptt:", d3$ptt, "<br>", "year:", d3$year,"<br>", "NDVIanom:", d3$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5,opacity=0.2, group='Late departure') %>%
-
-  addLayersControl(overlayGroups = c('Early departure', 'Late departure'), options=layersControlOptions(collapsed=F))
-
-m
-
-```
-
-## Habitat Variables
-
-Habitat of the final stopover before departure from West Africa does not seem to have a huge imapact.
-
-The percentage of tree cover does not affect departure date.
-
-```{r, warning=F, message=F, echo=F}
-
-m1<-lmer(DEPwestAF~Tree+(1|year)+(1|ptt), data=dat)
-
-#Anova(m1, test.statistic = 'F')
-
-ggplot(data=dat, aes(x=Tree, y=DEPwestAF ))+xlab("% tree cover of final stopover")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.27, p=0.609', x = -Inf, y = Inf), hjust = 0, vjust = 1)
-
-```
-
-The percentage of grass cover does appear to affect departure date. *Not sure how believeable this is*
-
-```{r, warning=F, message=F, echo=F} 
-m1<-lmer(DEPwestAF~Grass+(1|year)+(1|ptt), data=dat)
-
-#Anova(m1, test.statistic = 'F')
-
-newdat<-data.frame(DEPwestAF=1, Grass=1:75)
-newdat$p1<-predict(m1, newdata=newdat, re.form=~0)
-
-predmat<-model.matrix(DEPwestAF~Grass, data=newdat)
-vcv<-vcov(m1)
-semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
-newdat$lc<-newdat$p1-semod*1.96
-newdat$uc<-newdat$p1+semod*1.96
-
-qplot(data=dat, y=DEPwestAF, x=Grass)+
-  geom_line(data=newdat, aes(x=Grass, y=p1), colour='red')+
-  geom_line(data=newdat, aes(x=Grass, y=lc), colour='red', linetype='dashed')+
-  geom_line(data=newdat, aes(x=Grass, y=uc), colour='red', linetype='dashed')+
-  ylab("Departure date West Africa (DOY)")+
-  xlab("% grass cover of habitat")+
-  theme_bw()+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=7.54, p=0.011', x = -Inf, y = Inf), hjust = 0, vjust = 1)
-```
-
-Distance to rivers does not affect departure date.
-
-```{r, warning=F, message=F, echo=F}
-
-m1<-lmer(DEPwestAF~D_riv+(1|year)+(1|ptt), data=dat)
-
-#Anova(m1, test.statistic = 'F')
-
-ggplot(data=dat, aes(x=D_riv, y=DEPwestAF ))+xlab("Distance to nearest river (m)")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.39, p=0.531', x = -Inf, y = Inf), hjust = 0, vjust = 1)
-
-```
-
-## Dynamic environmental variables
-
-NDVI anomoly appears important. Greener final stopover sites allow earlier departure from West. *Now to test rainfall and greening in more detail*
-
-```{r, warning=F, message=F, echo=F}
-
-dat2.3<-dat2.2 %>% group_by(year, ptt) %>%
-  summarise_all(last)
-
-dat$year<-as.integer(dat$year)
-
-dat3<-full_join(dat, dat2.3, by=c('ptt', 'year'))
-
-m6<-lmer(DEPwestAF~aquaANOM+
-           (1|ptt)+(1|year), data=dat3)
-
-#summary(m6)
-
-# plot
-
-newdat<-data.frame(aquaANOM=seq(-0.1,0.15, 0.005), DEPwestAF=1)
-newdat$p1<-predict(m6, newdata=newdat, re.form=~0)
-
-predmat<-model.matrix(DEPwestAF~aquaANOM, data=newdat)
-vcv<-vcov(m6)
-semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
-newdat$lc<-newdat$p1-semod*1.96
-newdat$uc<-newdat$p1+semod*1.96
-
-qplot(data=dat3, x=aquaANOM, y=DEPwestAF)+
-  geom_line(data=newdat, aes(x=aquaANOM, y=p1), colour='red')+
-  geom_line(data=newdat, aes(x=aquaANOM, y=lc), colour='red', linetype='dashed')+
-  geom_line(data=newdat, aes(x=aquaANOM, y=uc), colour='red', linetype='dashed')+
-  xlab("NDVI anomoly")+
-  ylab("Departure date West Africa (DOY)")+
-  theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=4.65, p=0.038', x = -Inf, y = Inf), hjust = 0, vjust = 1)+theme(axis.text.x=element_text(angle=90))
-
-#Anova(m6, test.statistic = 'F')
-
-```

### 2018-08-30 ndvi_rainfall_playground.R
Fix to include Burkina Faso

@@ -60,7 +60,7 @@ dat<-left_join(dat, soversWA %>% group_by(ptt, year) %>%
 dat<-left_join(dat, soversWA %>%
                  filter(country %in% c("Ghana","Cote d'Ivoire",
                                        "Guinea", "Sierra Leone",
-                                       ' Burkina Faso', 'Togo','Liberia',
+                                       'Burkina Faso', 'Togo','Liberia',
                                        'Nigeria')) %>%
                  group_by(ptt, year) %>%
                  summarise(firstCdepartureG=first(country),

### 2018-08-30 cuckoo_demography.R
Edits to re-run with Chris grid cell suggestions and to re-run with spring birds subset

@@ -15,6 +15,19 @@ bod<-read.csv('C:/cuckoo_tracking/data/movebank_cuckoos_hybridfilter_bestofday_2
 
 bod<-bod %>% group_by(ptt) %>%summarise_all(first)
 
+## extra section to filter to just birds used in spring analyses
+dmod<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_attrib_modelready.csv', h=T)
+
+# just uk arrivals
+dmod<-dmod[-which(is.na(dmod$arrive_uk)),]
+
+bod<-bod %>% filter(ptt %in% unique(dmod$ptt))
+
+bod %>% mutate(lat1=ifelse(is.na(lat.tagged), location.lat, lat.tagged),
+               long1=ifelse(is.na(long.tagged), location.long, long.tagged)) -> bod
+
+write.csv(bod, 'C:/cuckoo_tracking/sourced_data/cuckoo_demography/first_point_ptt_bestofday_springbirds.csv', quote=F, row.names=F)
+
 #split scotland into 2 sites
 
 bod[bod$capture.location=='Scotland',]
@@ -60,13 +73,13 @@ for(i in seq_len(nrow(bodsf))){
 
 centredat<- do.call('rbind', closest)
 
-write.csv(centredat, 'C:/cuckoo_tracking/sourced_data/cuckoo_demography/Altas_centroids_nearest_tag_loc.csv', quote=F, row.names=F)
+write.csv(centredat, 'C:/cuckoo_tracking/sourced_data/cuckoo_demography/Altas_centroids_nearest_tag_loc_spring_birds.csv', quote=F, row.names=F)
 
 ## Read in shapefile of edited Bird Atlas sites
 
 setwd('C:/cuckoo_tracking/sourced_data')
 
-bat_site<-readOGR(layer='Atlas_tagging_areas_added', dsn='cuckoo_demography')
+bat_site<-readOGR(layer='Atlas_tagging_areas_added_spring_birds', dsn='cuckoo_demography')
 
 qplot(data=bat_site@data, x=bat_site@coords[,1], y=bat_site@coords[,2], colour=tagloc)
 
@@ -76,4 +89,16 @@ site_dem<-site_dem[site_dem$tagloc!='na',]
 
 site_dem %>% group_by(tagloc) %>% 
[…]

### 2018-08-30 cuckoo_demography.R
Added calc and plot of arrival dates against decline

@@ -102,3 +102,36 @@ b2<-batl[batl$tenkm%in%c_cells$TQ42,]
 
 write.csv(b2, 'C:/cuckoo_tracking/sourced_data/cuckoo_demography/Atlas_centroids_hewson_nature.csv', quote=F, row.names=F)
 
+
+# quick bit of analyses
+
+## extra section to filter to just birds used in spring analyses
+dmod<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_attrib_modelready.csv', h=T)
+
+dmod$breeding_loc<-as.character(dmod$breeding_loc)
+dmod[dmod$ptt==128300,]$breeding_loc<-'Kinloch Skye'
+dmod[dmod$breeding_loc=='Scotland',]$breeding_loc<-'Trossachs'
+
+pop_trend<-site_dem %>% group_by(tagloc) %>% 
+  summarise(mean_diff=mean(diff), sd_diff=sd(diff))
+
+pop_trend$tagloc<-recode(pop_trend$tagloc,
+                         'Ashdown For'='Ashdown Forest',
+                         'Lak District'='Lake District',
+                         'N York Moors' = 'North York Moors',
+                         'Norfolk Br'= 'Norfolk Broads',
+                         'Sherwood For'='Sherwood Forest',
+                         'Thetford For'='Thetford Forest')
+
+dmod<-left_join(dmod, pop_trend, by=c('breeding_loc'='tagloc'))
+# warning ok
+
+qplot(data=dmod, y=arrive_breeding, x=mean_diff, colour=breeding_loc)
+
+d2<-dmod %>% group_by(breeding_loc) %>% summarise_all(mean, na.rm=T)
+
+
+ggplot(data=d2, aes(y=arrive_uk, x=mean_diff))+
+  geom_point(aes(colour=breeding_loc), shape=2, size=2)+
+  geom_point(data=dmod ,aes(colour=breeding_loc), alpha=0.2)
+

### 2018-09-07 cuckoo_compile_clean_movebank_30Jan.R
Add code to make spring migration lines

@@ -537,3 +537,22 @@ writeOGR(obj=out_line, dsn="spatial",
          overwrite_layer = T)
 
 
+## Additional code to make spring migration only lines
+
+library(dplyr)
+library(sf)
+
+dat<-read.csv("C:/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_2018_clean_stopovers_recalc.csv", h=T)
+
+dmod<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_attrib_modelready.csv', h=T)
+
+dat$pttyear<-paste(dat$ptt, dat$year.x)
+dmod$pttyear<-paste(dmod$ptt, dmod$year)
+
+dat<-dat %>% filter(pttyear %in% unique(dmod$pttyear) & julian<145)
+
+dat_sp<-st_as_sf(dat, coords = c("long","lat"), crs="+proj=longlat +datum=WGS84")
+dat_ls<-dat_sp %>% group_by(pttyear) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
+
+st_write(dat_ls, "C:/cuckoo_tracking/data/spatial/spring_lines.shp")
+

### 2018-09-07 paper_wireframe_knit.Rmd
Updated to include population changes and graphic with monsoon arrival

@@
 library('ggplot2')
-library('dplyr')
 library('maps')
@@
 library('MuMIn')
+library('raster')
+library('rgdal')
+library('dplyr')
 
@@
 
-dmod<-dmod[,c(1,2,6,22,27,28,29,37,39,52,57,63,74,85,87,92, 115)]
+dmodWA<-dmod
 
-dmod<-na.omit(dmod)
+dmodWA$tWA=dmodWA$DEPwestAF-dmodWA$departureGarr
+
+dmodWA<-dmodWA[,c(1,2,6,20,22,27,28,29,37,39,52,57,63,74,85,87,92, 115)]
+
+dmodWA<-na.omit(dmodWA)
 
@@
 
-For every extra day spent in West Africa cuckoos arrive 0.78 days 
-later back at UK breeding grounds.
+For every extra day departure from West Africa is delayed cuckoos arrive 0.78 days later back at UK breeding grounds.
 
@@
 
-m5<-lmer(migration.length~depart.west.africa+
-           (1|ptt)+(1|year), data=d1)
+#m5<-lmer(migration.length~depart.west.africa+(1|ptt)+(1|year), data=d1)
 
@@
 
-m5a<-lm(migration.length~depart.west.africa*(depart.west.africa<102)+depart.west.africa*(depart.west.africa>102), data=d1)
+#m5a<-lm(migration.length~depart.west.africa*(depart.west.africa<102)+depart.west.africa*(depart.west.africa>102), data=d1)
 
-m5b<-lmer(migration.length~bs(depart.west.africa, deg=1, knots=102)+
-           (1|ptt)+(1|year), data=d1)
+#m5b<-lmer(migration.length~bs(depart.west.africa, deg=1, knots=102)+
+#           (1|ptt)+(1|year), data=d1)
 
-newdat<-data.frame(depart.west.africa=70:130, migration.length=1)
-newdat$p1<-predict(m5b, newdata=newdat, re.form=~0)
+#newdat<-data.frame(depart.west.africa=70:130, migration.length=1)
+#newdat$p1<-predict(m5b, newdata=newdat, re.form=~0)
 
@@
   ylab("Migration length (days)")+
-  theme_bw()+geom_line(data=newdat, aes(x=depart.west.africa, y=p1), colour='red')+geom_label(data=data.frame(), aes(label = 'Anova F=3.54, p=0.046', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+  theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=3.54, p=0.046', x = -Inf, y = Inf), hjust = 0, vjust = 1)
 
@@
 ggplot(data=dat, aes(x=breeding_loc, y=DEPwestAF))+geom_jitter(width=0.08, shape=1)+theme_bw()+theme(axis.text.x=element_text(angle=90))+xlab("Breeding location")+ylab("Departure date West Africa (DOY)")+geom_label(data=data.frame(), aes(label = 'Anova F=4.55, p=0.023', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+However when we group these populations into Upland/Lowland as per Hewson nature paper we see that upland birds appear to be running on a later migration schedule than lowland birds.
+Upland birds have later arrival to and departure from West Africa, and later arrival to the UK mainland than lowland birds.
+*This is potentially indicative of migration adjusted to later spring phenology in upland areas than lowlands*
+
+```{r, warning=F, message=F, echo=F, fig.height=15, fig.width=6}
+#library(lmerTest)
+library(gridExtra)
+
+dmod$breeding_hab<-'Upland'
+dmod[dmod$breeding_loc %in% c('Ashdown Forest', 'New Forest', 'Norfolk Broads',
+                              'Sherwood Forest', 'Thetford Forest'),]$breeding_hab<-'Lowland'
+
+m1<-lmer(depart_winterSO~breeding_hab+(1|ptt)+(1|year), data=dmod)
+m2<-lmer(DEPcentralAF~breeding_hab+(1|ptt)+(1|year), data=dmod)
+m2a<-lmer(departureGarr~breeding_hab+(1|ptt)+(1|year), data=dmod)
+m3<-lmer(DEPwestAF~breeding_hab+(1|ptt)+(1|year), data=dmod)
+m4<-lmer(arrive_uk~breeding_hab+(1|ptt)+(1|year), data=dmod)
+m5<-lmer(arrive_breeding~breeding_hab+(1|ptt)+(1|year), data=dmod)
+
+#anova(m1)
+#anova(m2)
+#anova(m2a)
+#anova(m3)
+#anova(m4)
+#anova(m5)
+
+p1<-ggplot(data=dmod, aes(x=breeding_hab, y=depart_winterSO))+geom_boxplot()+
+  geom_jitter(width=0.08)+
+  geom_label(data=data.frame(), 
+             aes(label = 'Anova F=4.24, p=0.052', x = -Inf, y = Inf), hjust = -1.5, vjust = 1)+
+  labs(x='', y='DOY', title='Depart winter (most southern) stopover')
+
+p2<-ggplot(data=dmod, aes(x=breeding_hab, y=DEPcentralAF))+geom_boxplot()+
+  geom_jitter(width=0.08)+
+  geom_label(data=data.frame(), 
+             aes(label = 'Anova F=2.01, p=0.171', x = -Inf, y = Inf), hjust = -1.5, vjust = 1)+
+  labs(x='', y='DOY', title='Depart Central Africa')
+
+p3<-ggplot(data=dmod, aes(x=breeding_hab, y=departureGarr))+geom_boxplot()+
+  geom_jitter(width=0.08)+
+  geom_label(data=data.frame(), 
+             aes(label = 'Anova F=5.78, p=0.025', x = -Inf, y = Inf), hjust = -1.5, vjust = 1)+
+  labs(x='', y='DOY', title='Arrive West Africa')
+
+p4<-ggplot(data=dmod, aes(x=breeding_hab, y=DEPwestAF))+geom_boxplot()+
+  geom_jitter(width=0.08)+
+  geom_label(data=data.frame(), 
+             aes(label = 'Anova F=4.53, p=0.045', x = -Inf, y = Inf), hjust = -1.5, vjust = 1)+
+  labs(x='', y='DOY', title='Depart West Africa')
+
+p5<-ggplot(data=dmod, aes(x=breeding_hab, y=arrive_uk))+geom_boxplot()+
+  geom_jitter(width=0.08)+
+  geom_label(data=data.frame(), 
+             aes(label = 'Anova F=6.21, p=0.017', x = -Inf, y = Inf), hjust = -1.5, vjust = 1)+
+  labs(x='',y='DOY', title='Arrival at UK mainland')
+
+p6<-ggplot(data=dmod, aes(x=breeding_hab, y=arrive_breeding))+geom_boxplot()+
+  geom_jitter(width=0.08)+
+  geom_label(data=data.frame(), 
+             aes(label = 'Anova F=7.29, p=0.021', x = -Inf, y = Inf), hjust = -1.5, vjust = 1)+
+  labs(x='Breeding habitat', y='DOY', title='Arrival at breeding grounds')
+
+grid.arrange(p1, p2, p3, p4, p5, p6, ncol=1, nrow=6)
+
 ```
@@
 
+A final test if to see if dates of arrival into the UK or departure from West Africa are related to population declines.
+The number of tagging sites gives good spatial coverage over the UK.
+Number of Bird Atlas cells attributed to each population are defined based on tagging locations as per Hewson nature paper.
+
+```{r, warning=F, message=F, echo=F}
+setwd('C:/cuckoo_tracking/sourced_data')
+
+bat_site<-readOGR(layer='Atlas_tagging_areas_added_spring_birds', dsn='cuckoo_demography', verbose = FALSE)
+
+qplot(data=bat_site@data, x=bat_site@coords[,1], y=bat_site@coords[,2], colour=tagloc)+labs(x='Long', y='Lat', colour='Population area')+xlim(c(0, 700000))
+```
+
+Levels of decline are similar to Hewson paper but differ slightly as only represent tagging locations of birds that complete spring migration.
+
+```{r, warning=F, message=F, echo=F}
+site_dem<-bat_site@data
+
+site_dem<-site_dem[site_dem$tagloc!='na',]
+
+dmod$breeding_loc<-as.character(dmod$breeding_loc)
+dmod[dmod$ptt==128300,]$breeding_loc<-'Kinloch Skye'
+dmod[dmod$breeding_loc=='Scotland',]$breeding_loc<-'Trossachs'
+
+pop_trend<-site_dem %>% group_by(tagloc) %>% 
+  summarise(mean_diff=mean(diff), sd_diff=sd(diff))
+
+pop_trend$tagloc<-recode(pop_trend$tagloc,
+                         'Ashdown For'='Ashdown Forest',
+                         'Lak District'='Lake District',
+                         'N York Moors' = 'North York Moors',
+                         'Norfolk Br'= 'Norfolk Broads',
+                         'Sherwood For'='Sherwood Forest',
+                         'Thetford For'='Thetford Forest')
+
+pop_trend
+```
+
+Plotted against arrival date into the UK, faded circles are individual bird arrivals and triangles the mean per population. Not sure if there is much of a trend here, up to you!
+
+```{r, warning=F, message=F, echo=F}
+dmod2<-left_join(dmod, pop_trend, by=c('breeding_loc'='tagloc'))
+# warning ok
+
+qplot(data=dmod2, y=arrive_breeding, x=mean_diff, colour=breeding_loc)
+
+dmod3<-dmod2 %>% group_by(breeding_loc) %>% summarise_all(mean, na.rm=T)
+
+dmod3$breeding_hab<-'Upland'
+dmod3[dmod3$breeding_loc %in% c('Ashdown Forest', 'New Forest', 'Norfolk Broads', 'Sherwood Forest', 'Thetford Forest'),]$breeding_hab<-'Lowland'
+
+ggplot(data=dmod3, aes(y=arrive_uk, x=mean_diff))+
+  geom_point(aes(colour=breeding_loc), shape=2, size=2)+
+  geom_point(data=dmod2 ,aes(colour=breeding_loc), alpha=0.3)+theme_bw()+labs(x='Mean population difference (1998-91 to 2008-11)', y='UK mainland arrival date (DOY)', colour='Population')
+
+```
+
+Nothing in West Africa departure date.
+
+```{r, warning=F, message=F, echo=F}
+ggplot(data=dmod3, aes(y=DEPwestAF, x=mean_diff))+
+  geom_point(aes(colour=breeding_loc), shape=2, size=2)+
+  geom_point(data=dmod2 ,aes(colour=breeding_loc), alpha=0.3)+theme_bw()+labs(x='Mean population difference (1998-91 to 2008-11)', y='Departure from West Africa (DOY)', colour='Population')
+```
+
+Just for confimation that declines are greater in Lowland populations - Yes.
+
+```{r, warning=F, message=F, echo=F}
+
+ggplot(data=dmod3, aes(y=mean_diff, x=breeding_hab))+geom_point(shape=1)+theme_bw()+labs(x='Mean population difference (1998-91 to 2008-11)', y='Breeding habitat')
+```
+
 ## Environmental factors that influence departure date from West Africa.
@@
 ggplot(data=dat, aes(x=year, y=DEPwestAF ))+xlab("Year")+ylab("Departure date West Africa (DOY)")+geom_jitter(width=0.08,shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=1.27, p=0.315', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
 
+Coefficient values for each year, 95% CI given in brackets.
+
+```{r, warning=F, message=F, echo=F}
 stargazer(m1, type='text',
@@
 
-Taking a step back, it is important to if there is any environmental
-control over how birds arrive into West Africa.
+On closer inspection we can see that the earlier arriving (into West Africa) half of birds (majority lowland birds) all end up departing via Ivory Coast (and a few via Ghana).
+Birds that arrive later into West Africa (majority upland birds) depart from 6 countries spread across the region.
+*This is suggestive of Ivory Coast having preferred habitat and/or receiving rains earlier than the rest of the region.
+Also this suggests that after ~day 80 rain has reached most countries in West Africa and late arriving birds can find suitable resources to fuel up pre-sahara migration across the region*
+
+```{r, warning=F, message=F, echo=F}
+
+qplot(data=dmodWA, x=departureGarr, y=DEPwestAF, colour=dep_country)+labs(x='West Africa arrival (DOY)', y='West Africa departure (DOY)', shape='Departure Country', colour='Breeding habitat')+theme_bw()
+
+```
+
+Same plot looked at with upland/lowland classification to show differences
+
+```{r, warning=F, message=F, echo=F}
+
+dmodWA$breeding_hab<-'Upland'
+dmodWA[dmodWA$breeding_loc %in% c('Ashdown Forest', 'New Forest', 'Norfolk Broads',
+                              'Sherwood Forest', 'Thetford Forest'),]$breeding_hab<-'Lowland'
+
+qplot(data=dmodWA, x=departureGarr, y=DEPwestAF, colour=breeding_hab, shape=dep_country)+theme_bw()+
+    labs(y='Departure from West Africa (DOY)', x='Arrival into West Africa (DOY)', colour='Departure country')
+
+```
+
+Taking a step back, it is important to if there is any environmental control over how birds arrive into West Africa.
 
@@
   geom_line(data=newdat, aes(x=departureGarr, y=uc), colour='red', linetype='dashed')+
-  ylab("Days spent in West Africa")+
+  ylab("Departure date West Africa (DOY)")+
   xlab("Arrival into West Africa (DOY)")+
@@
 ```
+
+Rather than looking at arrival of first rains we can concentrate on arrival of the monsoon, in particular its temporal and spatial variation.
+In the visualisation below we define monsoon arrival as the day of year when there has been cumulatively > 100ml of rain(we can modify this to any value) for each pixel.
+Turn of all the yearly tick boxes on the right then explore each year in turn to see temporal variation.
+Pan and zoom to explore spatial variation and how birds within each year time their stopovers to co-incide with arrival of monsoon.
+*Note how in 2016 rains are late across the entire region, but timing of departures from West Africa do not seem to be affected*
+
+```{r, warning=F, message=F, echo=F}
+
+r2012_mon<-raster('C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2012.tif')
+r2013_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2013.tif')
+r2014_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2014.tif')
+r2015_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2015.tif')
+r2016_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2016.tif')
+r2017_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2017.tif')
+r2018_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2018.tif')
+
+setwd('C:/cuckoo_tracking/sourced_data')
+
+wrld<-readOGR(layer='TM_WORLD_BORDERS-0.3', dsn='country_borders', verbose = FALSE)
+
+dat$year<-as.integer(dat$year)
+ soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
+
+ forsp<-na.omit(soversWA1[,c(1,2,7,8,9,17,18,22)])
+ forsp<-forsp %>% group_by(ptt, year) %>% mutate(nzz=n()) %>% filter(nzz>1) # drops Peckham in 2016 that cocks up linestring
+ dat_sp<-st_as_sf(forsp, coords = c("SO_median_long","SO_median_lat"), crs="+proj=longlat +datum=WGS84")
+ dat_ls<-dat_sp %>% group_by(ptt, year, DEPwestAF) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
+ #dat_ls
+ #setwd('C:/cuckoo_tracking/data/spatial');st_write(dat_ls,'spring_linesWA.shp')
+
+pal <- colorBin(c('#9e0142','#d53e4f','#f46d43','#fdae61','#fee08b','#e6f598','#abdda4','#66c2a5','#3288bd','#5e4fa2'), c(NA, 0:125), na.color = "transparent")
+
+ m <- leaflet() %>%
+  addTiles() %>%
+
+  addRasterImage(r2012_mon, colors = pal, opacity = 0.8, group='2012') %>%
+     addRasterImage(r2013_mon, colors = pal, opacity = 0.8, group='2013') %>%
+     addRasterImage(r2014_mon, colors = pal, opacity = 0.8, group='2014') %>%
+     addRasterImage(r2015_mon, colors = pal, opacity = 0.8, group='2015') %>%
+     addRasterImage(r2016_mon, colors = pal, opacity = 0.8, group='2016') %>%
+     addRasterImage(r2017_mon, colors = pal, opacity = 0.8, group='2017') %>%
+     addRasterImage(r2018_mon, colors = pal, opacity = 0.8, group='2018') %>%
+   addLegend(pal = pal, values = c(NA, 0:125),
+    title = "Monsoon arrival day") %>%
+
+   addPolylines(data=wrld[wrld$REGION==2,], color=1, weight=2, opacity=1) %>%
+
+  addPolylines(data=dat_ls[dat_ls$year==2012,], weight=1, color=1, group='2012', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2012,], popup= paste("ptt:", dat_sp[dat_sp$year==2012,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2012,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2012,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2012,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2012') %>%
+
+     addPolylines(data=dat_ls[dat_ls$year==2013,], weight=1, color=1, group='2013', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2013,], popup= paste("ptt:", dat_sp[dat_sp$year==2013,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2013,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2013,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2013,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2013') %>%
+
+     addPolylines(data=dat_ls[dat_ls$year==2014,], weight=1, color=1, group='2014', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2014,], popup= paste("ptt:", dat_sp[dat_sp$year==2014,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2014,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2014,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2014,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2014') %>%
+
+     addPolylines(data=dat_ls[dat_ls$year==2015,], weight=1, color=1, group='2015', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2015,], popup= paste("ptt:", dat_sp[dat_sp$year==2015,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2015,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2015,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2015,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2015') %>%
+
+     addPolylines(data=dat_ls[dat_ls$year==2016,], weight=1, color=1, group='2016', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2016,], popup= paste("ptt:", dat_sp[dat_sp$year==2016,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2016,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2016,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2016,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2016') %>%
+
+     addPolylines(data=dat_ls[dat_ls$year==2017,], weight=1, color=1, group='2017', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2017,], popup= paste("ptt:", dat_sp[dat_sp$year==2017,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2017,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2017,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2017,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2017') %>%
+
+     addPolylines(data=dat_ls[dat_ls$year==2018,], weight=1, color=1, group='2018', dashArray='4,2') %>%
+
+  addCircleMarkers(data=dat_sp[dat_sp$year==2018,], popup= paste("ptt:", dat_sp[dat_sp$year==2018,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2018,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2018,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2018,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2018') %>%
+
+     addLayersControl(overlayGroups = c('2012', '2013', '2014', '2015', '2016', '2017', '2018'), options=layersControlOptions(collapsed=F))
+
+m
+
+```
+
+OK so are there any explanatory environmental variables that can explain the variation in cuckoo departure from West Africa
+
+```{r, warning=F, message=F, echo=F} 
+ # make two gradient metrics
+ dmod$ndvi_last_ang<-dmod$ndviTdiff_last*dmod$ndvidiff_last
+dmod$rain_first_ang<-(dmod$monsTdiff_first+25)*dmod$raindiff2_first
+ m0<-lmer(DEPwestAF~(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m1<-lmer(DEPwestAF~dep_country+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m2<-lmer(DEPwestAF~firstCdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m3<-lmer(DEPwestAF~departureGarr+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m4<-lmer(DEPwestAF~nSOdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m5<-lmer(DEPwestAF~monsTdiff_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m6<-lmer(DEPwestAF~raindiff2_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m7<-lmer(DEPwestAF~emodisNDVI_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m8<-lmer(DEPwestAF~ndviTdiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m9<-lmer(DEPwestAF~ndvidiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m10<-lmer(DEPwestAF~emodisANOM_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m11<-lmer(DEPwestAF~Trees_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m12<-lmer(DEPwestAF~Croplands_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+m13<-lmer(DEPwestAF~d_river_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+ m14<-lmer(DEPwestAF~rain_first_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+ m15<-lmer(DEPwestAF~ndvi_last_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
+ anova(m0,m1,m2,m3,m4,m5,m6,m7,m8,m9,m10, m11, m12,m13, m14, m15)
+ moddat<-dmod[dmod$DEPwestAF %in% 85:120,]
+ mfin<-lmer(DEPwestAF~departureGarr+nSOdepartureG+monsTdiff_first*raindiff2_first+emodisNDVI_last+ndviTdiff_last*ndvidiff_last+emodisANOM_mean+Trees_mean+Croplands_mean+d_river_mean+(1|ptt)+(1|year), data=moddat,  na.action = na.fail)
+ mdr<-dredge(mfin)
+ get.models(mdr, subset = delta < 4, method = "REML")
+ #qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=nSOdepartureG)
+ #qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=ndviTdiff_last)
+ #qplot(data=dmod[dmod$DEPwestAF %in% 85:120 & dmod$emodisANOM_mean>-0.05,], y=DEPwestAF, x=emodisANOM_mean)
+```
+
+ So what do these early leaving birds do in their runup to departing Cote d'Ivoire that is different to other birds?
+The tracks of early departing birds are grouped with blue lines and late departing with orange lines.
+Each stopover is coloured either red (indicating below average NDVI) or green (above average NDVI) to see if early birds chain together 'greener stopovers.
+
+```{r, warning=F, message=F, echo=F} 
+ dat$year<-as.integer(dat$year)
+ soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
+ soversWA2<-left_join(soversWA1, dat2.2[,c(1,2, 3,9)], by=c('ptt', 'year', 'SO_startDOY'))
+ forsp<-na.omit(soversWA2[,c(1,2,7,8,9,17,18,22,23)])
+ forsp<-forsp %>% group_by(ptt, year) %>% mutate(nzz=n()) %>% filter(nzz>1) # drops Peckham in 2016 that cocks up linestring
+ dat_sp<-st_as_sf(forsp, coords = c("SO_median_long","SO_median_lat"), crs="+proj=longlat +datum=WGS84")
+ dat_ls<-dat_sp %>% group_by(ptt, year, DEPwestAF) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
+ #dat_ls
+ #setwd('C:/cuckoo_tracking/data/spatial');st_write(dat_ls,'spring_linesWA.shp')
+ d2<-dat_sp[dat_sp$DEPwestAF<101,]
+d3<-dat_sp[dat_sp$DEPwestAF>100,]
+ m <- leaflet() %>%
+  addTiles() %>%
+
+  addPolylines(data=dat_ls[dat_ls$DEPwestAF<101,], weight=1, color='blue', group='Early departure') %>%
+
+    addPolylines(data=dat_ls[dat_ls$DEPwestAF>100,], weight=1, color='orange', group='Late departure') %>%
+
+  addCircleMarkers(data=d2, popup= paste("ptt:", d2$ptt, "<br>", "year:", d2$year,"<br>", "NDVIanom:", d2$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, opacity=0.2, group='Early departure') %>%
+
+    addCircleMarkers(data=d3, popup= paste("ptt:", d3$ptt, "<br>", "year:", d3$year,"<br>", "NDVIanom:", d3$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5,opacity=0.2, group='Late departure') %>%
+
+  addLayersControl(overlayGroups = c('Early departure', 'Late departure'), options=layersControlOptions(collapsed=F))
+
+m           
+```
+
+Habitat Variables
+Habitat of the final stopover before departure from West Africa does not seem to have a huge imapact.
+The percentage of tree cover does not affect departure date.
+
+```{r, warning=F, message=F, echo=F} 
+ m1<-lmer(DEPwestAF~Tree+(1|year)+(1|ptt), data=dat)
+ #Anova(m1, test.statistic = 'F')
+ ggplot(data=dat, aes(x=Tree, y=DEPwestAF ))+xlab("% tree cover of final stopover")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.27, p=0.609', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+ ```
+ The percentage of grass cover does appear to affect departure date. *Not sure how believeable this is*
+ ```{r, warning=F, message=F, echo=F} 
+m1<-lmer(DEPwestAF~Grass+(1|year)+(1|ptt), data=dat)
+ #Anova(m1, test.statistic = 'F')
+ newdat<-data.frame(DEPwestAF=1, Grass=1:75)
+newdat$p1<-predict(m1, newdata=newdat, re.form=~0)
+ predmat<-model.matrix(DEPwestAF~Grass, data=newdat)
+vcv<-vcov(m1)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+ qplot(data=dat, y=DEPwestAF, x=Grass)+
+  geom_line(data=newdat, aes(x=Grass, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=Grass, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=Grass, y=uc), colour='red', linetype='dashed')+
+  ylab("Departure date West Africa (DOY)")+
+  xlab("% grass cover of habitat")+
+  theme_bw()+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=7.54, p=0.011', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+Distance to rivers does not affect departure date.
+
+```{r, warning=F, message=F, echo=F} 
+ m1<-lmer(DEPwestAF~D_riv+(1|year)+(1|ptt), data=dat)
+ #Anova(m1, test.statistic = 'F')
+ ggplot(data=dat, aes(x=D_riv, y=DEPwestAF ))+xlab("Distance to nearest river (m)")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.39, p=0.531', x = -Inf, y = Inf), hjust = 0, vjust = 1)
+```
+
+Dynamic environmental variables
+NDVI anomoly appears important. Greener final stopover sites allow earlier departure from West. *Now to test rainfall and greening in more detail*
+
+```{r, warning=F, message=F, echo=F}
+ dat2.3<-dat2.2 %>% group_by(year, ptt) %>%
+  summarise_all(last)
+ dat$year<-as.integer(dat$year)
+ dat3<-full_join(dat, dat2.3, by=c('ptt', 'year'))
+ m6<-lmer(DEPwestAF~aquaANOM+
+           (1|ptt)+(1|year), data=dat3)
+ #summary(m6)
+ # plot
+ newdat<-data.frame(aquaANOM=seq(-0.1,0.15, 0.005), DEPwestAF=1)
+newdat$p1<-predict(m6, newdata=newdat, re.form=~0)
+ predmat<-model.matrix(DEPwestAF~aquaANOM, data=newdat)
+vcv<-vcov(m6)
+semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
+newdat$lc<-newdat$p1-semod*1.96
+newdat$uc<-newdat$p1+semod*1.96
+ qplot(data=dat3, x=aquaANOM, y=DEPwestAF)+
+  geom_line(data=newdat, aes(x=aquaANOM, y=p1), colour='red')+
+  geom_line(data=newdat, aes(x=aquaANOM, y=lc), colour='red', linetype='dashed')+
+  geom_line(data=newdat, aes(x=aquaANOM, y=uc), colour='red', linetype='dashed')+
+  xlab("NDVI anomoly")+
+  ylab("Departure date West Africa (DOY)")+
+  theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=4.65, p=0.038', x = -Inf, y = Inf), hjust = 0, vjust = 1)+theme(axis.text.x=element_text(angle=90))
+ #Anova(m6, test.statistic = 'F')
+```

### 2018-09-07 stopover_rainfall_ndvi_detailcoords_vis_DEPWAbirds.Rmd
small addition to make best plot

@@
 
+#make epic plot
+#ggplot(data=t2)+
+#    geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, #ymax=500), fill='grey')+
+#    geom_line( aes(x=variable, y=emodisNDVI*500), colour='light #green', size=1)+
+#    geom_line(aes(x=variable, y=cumrf), col='blue', size=1)+
+#    facet_wrap(~country, ncol=1)+theme_classic()+
+#    ylab('Cumulative rainfall since 01 Jan (mm)')+ xlab('Day of #year')+
+#  scale_y_continuous(limits=c(0,500),sec.axis = #sec_axis(~./500, name='NDVI'))
+
 print(temp)

### 2018-09-07 west_africa_departure_model.R
minor edit

@@ -577,8 +577,3 @@ grid.arrange(p1, p2, p3, p4, p5, p6, ncol=1, nrow=6)
 
 dev.off()
 
-
-## bit of code for monsoon arrival
-
-ggplot(data=dat2.1, aes(x=SO_median_long, y=SO_median_lat))+scale_colour_gradientn(colours=topo.colors(10))+facet_wrap(~year)+geom_point(aes(colour=monsarrDOY))+theme_bw()
-

### 2018-09-17 rainfall_compile.R
tiny (old) edit from lubuntu

@@ -265,9 +265,9 @@ for(i in 1:123)
 
 #### big one
 
-ani.options(ani.width=1000, ani.height=1000)
+ani.options(ani.width=2000, ani.height=2000)
 
-saveGIF({
+saveHTML({
   for(i in 1:110)
   {
     
@@ -394,6 +394,9 @@ saveGIF({
      
      
      ras1<-subset(r2013,i)
+     
+     if(i==87){ras1<-subset(r2013,86)}
+     # hack to exclude erreroneous image
      
      if(i==1){cumra13<-ras1}else{cumra13<-cumra13+ras1}
      
@@ -715,7 +718,7 @@ saveGIF({
   }
   print(i)
   
-}, interval=1, "outputs/rainfall_stopoversALL.gif")
+}, interval=1, img.name="rplot", htmlfile="rainfall_stopoversALL.html")
 
 
 

### 2018-09-17 gd_cuck_plots.Rmd

@@
 
-dat<-read.csv("~/BTO/cuckoo_tracking/data/stopover_bestofday_spring_mig.csv", h=T)
+dat<-read.csv("C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig.csv", h=T)
 
-dat2<-read.csv("~/BTO/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_clean_stopovers.csv", h=T)
+dat2<-read.csv("C:/cuckoo_tracking/data/processed_movebank_cuckoos_hybrid_filter_2018_clean_stopovers_recalc.csv", h=T)
 
-UK_travel<-read.csv("~/BTO/cuckoo_tracking/data/complete_cycles_UK_all_timing.csv", h=T)
+UK_travel<-read.csv("C:/cuckoo_tracking/data/complete_cycles_UK_all_timing.csv", h=T)
 
@@
 
+names(dat3)[names(dat3)=='year.x']<-'year'
+
 m1 <- get_map(location = c(5, 30), zoom=4)
@@
 {
-  cols <- c("2012" = "red", "2013" = "blue", "2014" = "darkgreen", "2015" = "orange", "2016"= "cyan", "2017"= "black")
+  cols <- c("2012" = "red", "2013" = "blue", "2014" = "darkgreen", "2015" = "orange", "2016"= "cyan", "2017"= "black",
+            "2018"= "purple")
 
   g<-ggmap(m1)+
-    geom_path(data=dat3[dat3$ptt==i,], aes(x=long, y=lat,
-                                             colour=factor(year.x)))+
+    geom_path(data=dat3[dat3$ptt==i,], aes(x=long, y=lat,                                      colour=factor(year)))+
     geom_jitter(data=dat3[dat3$ptt==i,], aes(x=long, y=lat,
-                colour=factor(year.x)), shape=1, width = 0.5, height= 0.5)+ggtitle(i)+
+                colour=factor(year)), shape=1, width = 0.5, height= 0.5)+ggtitle(i)+
 
@@
                 colour=factor(year), label=substr(SO_start, 6,10)))+
-  geom_label_repel(data=aggregate(year~breeding_entry, data=dat[dat$ptt==i & dat$deployment_entry==FALSE,], FUN=function(x){unique(x)})
+  geom_label_repel(data=aggregate(year~breeding_entry, data=dat[dat$ptt==i,], FUN=function(x){unique(x)})
                    , aes(x=-18, y=45,
-                colour=factor(year), label=unique(substr(breeding_entry, 6,10)))) + scale_colour_manual(values = cols)
+                colour=factor(year), label=substr(breeding_entry, 6,10))) + scale_colour_manual(values = cols)
 

### 2018-09-17 stopover_rainfall_vis.Rmd

@@
 
-int_out<-read.csv('data/spring_rainfall_NDVI_by_stopover_2018_dead.csv', h=T)
+int_out<-read.csv('data/spring_rainfall_NDVI_GRIMMS_by_stopover_2018_dead.csv', h=T)
 
@@
 
-for( i in 2012:2017)
+for( i in 2012:2018)
   {
@@
 geom_rect(aes(xmin=SO_startDOY, xmax=SO_endDOY, ymin=0, ymax=500), fill='grey')+
-    geom_bar(data=t2, aes(x=variable, y=value2*500, group=1), stat='identity', fill='red', linetype=0, alpha=1/3)+
+    geom_line(data=t2, aes(x=variable, y=value2*500), colour='red')+
+  geom_line(data=t2, aes(x=variable, y=aquaNDVI*500), colour='cyan')+
+  geom_line(data=t2, aes(x=variable, y=terrNDVI*500), colour='orange')+
   geom_bar(stat='identity', colour='black')+

### 2018-09-17 paper_wireframe_knit.Rmd
Edits to remove leaflet sections and knit to word document

@@
 date: "10 July 2018"
-output: html_document
+output:
+  word_document: default
+  html_document: default
 ---
@@
   setwd("C:/cuckoo_tracking")}else{
-    setwd("N:/cuckoo_tracking")}
+    setwd("~/BTO/cuckoo_tracking")}
 
 #summary data
-dat<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)
+dat<- read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead.csv', h=T)
 
 #stopover coord data (median points)
-sovers<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_BOTH.csv', h=T)
+sovers<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig_BOTH.csv', h=T)
 
 ## landcover and river data
-lc<-read.csv("C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", h=T)
+lc<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_detailcoords_landcov_ext_rivers.csv", h=T)
 
@@
 
-extraz<- read.csv('C:/cuckoo_tracking/data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv', h=T)
+extraz<- read.csv('data/stopover_bestofday_1daymin_recalc_spring_mig_summary_extras.csv', h=T)
 
@@
 
-env<-read.csv('C:/cuckoo_tracking/data/spring_rainfall_NDVI_GRIMMS_TAMSAT_emodis_by_stopover_detailcoords_2018_dead.csv', h=T)
+env<-read.csv('data/spring_rainfall_NDVI_GRIMMS_TAMSAT_emodis_by_stopover_detailcoords_2018_dead.csv', h=T)
 
@@
 
-dmod<-read.csv('C:/cuckoo_tracking/data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_attrib_modelready.csv', h=T)
+dmod<-read.csv('data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_attrib_modelready.csv', h=T)
 
@@
 Explore the dataset. Each point gives the final stopover in West Africa before migration. Colours denote different years.
-
-```{r, warning=F, message=F, echo=F}
-
-dat_sp<-st_as_sf(na.omit(dat[,c(1,2,6,23,24)]), coords = c("dep_long","dep_lat"), crs="+proj=longlat +datum=WGS84")
-
-dat_sp$year<-factor(dat_sp$year)
-
-pal <- colorFactor(c("navy", "blue", 'cyan', 'green', 'yellow', 'orange', 'red'), domain = c("2012", "2013", '2014', '2015', '2016', '2017', '2018'))
-
-m <- leaflet(dat_sp) %>%
-  addTiles() %>% 
-  addCircleMarkers( popup= paste("ptt:", dat_sp$ptt, "<br>", "year:", dat_sp$year,"<br>", "Departure date:", dat_sp$DEPwestAF), color=~pal(year), stroke=F, radius=6, fillOpacity = 0.5)
-
-m
-
-```
 
@@
 ```{r, warning=F, message=F, echo=F}
-setwd('C:/cuckoo_tracking/sourced_data')
+setwd('~/BTO/cuckoo_tracking/sourced_data')
 
@@
 
-Rather than looking at arrival of first rains we can concentrate on arrival of the monsoon, in particular its temporal and spatial variation.
-In the visualisation below we define monsoon arrival as the day of year when there has been cumulatively > 100ml of rain(we can modify this to any value) for each pixel.
-Turn of all the yearly tick boxes on the right then explore each year in turn to see temporal variation.
-Pan and zoom to explore spatial variation and how birds within each year time their stopovers to co-incide with arrival of monsoon.
-*Note how in 2016 rains are late across the entire region, but timing of departures from West Africa do not seem to be affected*
-
-```{r, warning=F, message=F, echo=F}
-
-r2012_mon<-raster('C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2012.tif')
-r2013_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2013.tif')
-r2014_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2014.tif')
-r2015_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2015.tif')
-r2016_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2016.tif')
-r2017_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2017.tif')
-r2018_mon<-raster( 'C:/cuckoo_tracking/sourced_data/PERSIANN_rainfall/mons_arr2018.tif')
-
-setwd('C:/cuckoo_tracking/sourced_data')
-
-wrld<-readOGR(layer='TM_WORLD_BORDERS-0.3', dsn='country_borders', verbose = FALSE)
-
-dat$year<-as.integer(dat$year)
- soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
-
- forsp<-na.omit(soversWA1[,c(1,2,7,8,9,17,18,22)])
- forsp<-forsp %>% group_by(ptt, year) %>% mutate(nzz=n()) %>% filter(nzz>1) # drops Peckham in 2016 that cocks up linestring
- dat_sp<-st_as_sf(forsp, coords = c("SO_median_long","SO_median_lat"), crs="+proj=longlat +datum=WGS84")
- dat_ls<-dat_sp %>% group_by(ptt, year, DEPwestAF) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
- #dat_ls
- #setwd('C:/cuckoo_tracking/data/spatial');st_write(dat_ls,'spring_linesWA.shp')
-
-pal <- colorBin(c('#9e0142','#d53e4f','#f46d43','#fdae61','#fee08b','#e6f598','#abdda4','#66c2a5','#3288bd','#5e4fa2'), c(NA, 0:125), na.color = "transparent")
-
- m <- leaflet() %>%
-  addTiles() %>%
-
-  addRasterImage(r2012_mon, colors = pal, opacity = 0.8, group='2012') %>%
-     addRasterImage(r2013_mon, colors = pal, opacity = 0.8, group='2013') %>%
-     addRasterImage(r2014_mon, colors = pal, opacity = 0.8, group='2014') %>%
-     addRasterImage(r2015_mon, colors = pal, opacity = 0.8, group='2015') %>%
-     addRasterImage(r2016_mon, colors = pal, opacity = 0.8, group='2016') %>%
-     addRasterImage(r2017_mon, colors = pal, opacity = 0.8, group='2017') %>%
-     addRasterImage(r2018_mon, colors = pal, opacity = 0.8, group='2018') %>%
-   addLegend(pal = pal, values = c(NA, 0:125),
-    title = "Monsoon arrival day") %>%
-
-   addPolylines(data=wrld[wrld$REGION==2,], color=1, weight=2, opacity=1) %>%
-
-  addPolylines(data=dat_ls[dat_ls$year==2012,], weight=1, color=1, group='2012', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2012,], popup= paste("ptt:", dat_sp[dat_sp$year==2012,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2012,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2012,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2012,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2012') %>%
-
-     addPolylines(data=dat_ls[dat_ls$year==2013,], weight=1, color=1, group='2013', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2013,], popup= paste("ptt:", dat_sp[dat_sp$year==2013,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2013,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2013,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2013,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2013') %>%
-
-     addPolylines(data=dat_ls[dat_ls$year==2014,], weight=1, color=1, group='2014', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2014,], popup= paste("ptt:", dat_sp[dat_sp$year==2014,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2014,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2014,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2014,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2014') %>%
-
-     addPolylines(data=dat_ls[dat_ls$year==2015,], weight=1, color=1, group='2015', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2015,], popup= paste("ptt:", dat_sp[dat_sp$year==2015,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2015,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2015,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2015,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2015') %>%
-
-     addPolylines(data=dat_ls[dat_ls$year==2016,], weight=1, color=1, group='2016', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2016,], popup= paste("ptt:", dat_sp[dat_sp$year==2016,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2016,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2016,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2016,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2016') %>%
-
-     addPolylines(data=dat_ls[dat_ls$year==2017,], weight=1, color=1, group='2017', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2017,], popup= paste("ptt:", dat_sp[dat_sp$year==2017,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2017,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2017,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2017,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2017') %>%
-
-     addPolylines(data=dat_ls[dat_ls$year==2018,], weight=1, color=1, group='2018', dashArray='4,2') %>%
-
-  addCircleMarkers(data=dat_sp[dat_sp$year==2018,], popup= paste("ptt:", dat_sp[dat_sp$year==2018,]$ptt, "<br>", "name:", dat_sp[dat_sp$year==2018,]$name,"<br>", "SO start:", dat_sp[dat_sp$year==2018,]$SO_startDOY,"<br>", "depart WA:", dat_sp[dat_sp$year==2018,]$DEPwestAF), color='#219607', stroke=T, radius=3, group='2018') %>%
-
-     addLayersControl(overlayGroups = c('2012', '2013', '2014', '2015', '2016', '2017', '2018'), options=layersControlOptions(collapsed=F))
-
-m
-
-```
-
-OK so are there any explanatory environmental variables that can explain the variation in cuckoo departure from West Africa
-
-```{r, warning=F, message=F, echo=F} 
- # make two gradient metrics
- dmod$ndvi_last_ang<-dmod$ndviTdiff_last*dmod$ndvidiff_last
-dmod$rain_first_ang<-(dmod$monsTdiff_first+25)*dmod$raindiff2_first
- m0<-lmer(DEPwestAF~(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m1<-lmer(DEPwestAF~dep_country+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m2<-lmer(DEPwestAF~firstCdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m3<-lmer(DEPwestAF~departureGarr+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m4<-lmer(DEPwestAF~nSOdepartureG+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m5<-lmer(DEPwestAF~monsTdiff_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m6<-lmer(DEPwestAF~raindiff2_first+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m7<-lmer(DEPwestAF~emodisNDVI_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m8<-lmer(DEPwestAF~ndviTdiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m9<-lmer(DEPwestAF~ndvidiff_last+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m10<-lmer(DEPwestAF~emodisANOM_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m11<-lmer(DEPwestAF~Trees_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m12<-lmer(DEPwestAF~Croplands_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
-m13<-lmer(DEPwestAF~d_river_mean+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
- m14<-lmer(DEPwestAF~rain_first_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
- m15<-lmer(DEPwestAF~ndvi_last_ang+(1|ptt)+(1|year), data=dmod[dmod$DEPwestAF %in% 85:120,])
- anova(m0,m1,m2,m3,m4,m5,m6,m7,m8,m9,m10, m11, m12,m13, m14, m15)
- moddat<-dmod[dmod$DEPwestAF %in% 85:120,]
- mfin<-lmer(DEPwestAF~departureGarr+nSOdepartureG+monsTdiff_first*raindiff2_first+emodisNDVI_last+ndviTdiff_last*ndvidiff_last+emodisANOM_mean+Trees_mean+Croplands_mean+d_river_mean+(1|ptt)+(1|year), data=moddat,  na.action = na.fail)
- mdr<-dredge(mfin)
- get.models(mdr, subset = delta < 4, method = "REML")
- #qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=nSOdepartureG)
- #qplot(data=dmod[dmod$DEPwestAF %in% 85:120,], y=DEPwestAF, x=ndviTdiff_last)
- #qplot(data=dmod[dmod$DEPwestAF %in% 85:120 & dmod$emodisANOM_mean>-0.05,], y=DEPwestAF, x=emodisANOM_mean)
-```
-
- So what do these early leaving birds do in their runup to departing Cote d'Ivoire that is different to other birds?
-The tracks of early departing birds are grouped with blue lines and late departing with orange lines.
-Each stopover is coloured either red (indicating below average NDVI) or green (above average NDVI) to see if early birds chain together 'greener stopovers.
-
-```{r, warning=F, message=F, echo=F} 
- dat$year<-as.integer(dat$year)
- soversWA1<-left_join(soversWA, dat[,c(1,2,6)], by=c('ptt', 'year'))
- soversWA2<-left_join(soversWA1, dat2.2[,c(1,2, 3,9)], by=c('ptt', 'year', 'SO_startDOY'))
- forsp<-na.omit(soversWA2[,c(1,2,7,8,9,17,18,22,23)])
- forsp<-forsp %>% group_by(ptt, year) %>% mutate(nzz=n()) %>% filter(nzz>1) # drops Peckham in 2016 that cocks up linestring
- dat_sp<-st_as_sf(forsp, coords = c("SO_median_long","SO_median_lat"), crs="+proj=longlat +datum=WGS84")
- dat_ls<-dat_sp %>% group_by(ptt, year, DEPwestAF) %>% summarise(do_union=F) %>% st_cast("LINESTRING")
- #dat_ls
- #setwd('C:/cuckoo_tracking/data/spatial');st_write(dat_ls,'spring_linesWA.shp')
- d2<-dat_sp[dat_sp$DEPwestAF<101,]
-d3<-dat_sp[dat_sp$DEPwestAF>100,]
- m <- leaflet() %>%
-  addTiles() %>%
-
-  addPolylines(data=dat_ls[dat_ls$DEPwestAF<101,], weight=1, color='blue', group='Early departure') %>%
-
-    addPolylines(data=dat_ls[dat_ls$DEPwestAF>100,], weight=1, color='orange', group='Late departure') %>%
-
-  addCircleMarkers(data=d2, popup= paste("ptt:", d2$ptt, "<br>", "year:", d2$year,"<br>", "NDVIanom:", d2$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5, opacity=0.2, group='Early departure') %>%
-
-    addCircleMarkers(data=d3, popup= paste("ptt:", d3$ptt, "<br>", "year:", d3$year,"<br>", "NDVIanom:", d3$aquaANOM), color=~ifelse(aquaANOM>0, 'green', 'red'), stroke=F, radius=5,opacity=0.2, group='Late departure') %>%
-
-  addLayersControl(overlayGroups = c('Early departure', 'Late departure'), options=layersControlOptions(collapsed=F))
-
-m           
-```
-
 Habitat Variables

### 2018-09-20 paper_wireframe_knit.Rmd
Edits from linux PC, aminly to make docx knitr document of paper_wireframe

@@
 library('dplyr')
+library('reshape2')
 
@@
 
+# read in final dataset
+datfinal<-read.csv("data/stopover_bestofday_2018_1daymin_recalc_spring_mig_summary_dead_final.csv", h=T)
+
 ```
@@
 
-ggplot(data=dat, aes(y=arrive_breeding))+
+ggplot(data=datfinal, aes(y=arrive_breeding))+
   geom_point(aes(x=arrive_uk), col=1, shape=1)+
-  geom_point(aes(x=DEPeurope), col=2, shape=1)+
   geom_point(aes(x=DEPnorthAF), col=3, shape=1)+
@@
 
-Stats confirm that date of departure from West Africa significantly impacts subsequent date of arrival at Uk breeding grounds, but prior stages of spring migration do not.
+Stats confirm that date of departure from West Africa significantly impacts subsequent date of arrival at UK breeding grounds, but prior stages do not.
+*Note after reanalyses of departure from central africa we see that it is significant upon arrival at UK breeding grounds, but not highly so*
 
 ```{r, echo=F}
-d1<-dat
-  names(d1)[10]<-'arrive.breeding.uk'
-  names(d1)[4]<-'depart.wintering'
-  names(d1)[5]<-'depart.central.africa'
-  names(d1)[6]<-'depart.west.africa'
+m1<-lmer(arrive_breeding~arrive_uk+(1|breeding_hab/ptt)+(1|year), data=datfinal)
+m2<-lmer(arrive_breeding~DEPnorthAF+(1|breeding_hab/ptt)+(1|year), data=datfinal)
+m3<-lmer(arrive_breeding~DEPwestAF+(1|breeding_hab/ptt)+(1|year), data=datfinal)
 
-m1<-lmer(arrive.breeding.uk~depart.wintering+depart.central.africa+depart.west.africa+(1|year)+(1|ptt), data=d1)
+m5<-lmer(arrive_breeding~DEPcentralAF+(1|breeding_hab/ptt)+(1|year), data=datfinal)
+m6<-lmer(arrive_breeding~depart_winterSO+(1|breeding_hab/ptt)+(1|year), data=datfinal)
 
+'Arrival at UK border'
 Anova(m1, test.statistic='F')
+'Arrival in Europe'
+Anova(m2, test.statistic='F')
+'Depart West Africa'
+Anova(m3, test.statistic='F')
+'Depart Central Africa'
+Anova(m5, test.statistic='F')
+'Depart wintering grounds'
+Anova(m6, test.statistic='F')
 
@@
 
-For every extra day departure from West Africa is delayed cuckoos arrive 0.78 days later back at UK breeding grounds.
+Looking at explanatory power (r2), we can see that the there is a large increase in explanatory power between departing Central Africa and departing West AFrica
 
 ```{r, warning=F, message=F, echo=F}
+'Arrival at UK border'
+r.squaredGLMM(m1)
+'Arrival in Europe'
+r.squaredGLMM(m2)
+'Depart West Africa'
+r.squaredGLMM(m3)
+'Depart Central Africa'
+r.squaredGLMM(m5)
+'Depart wintering grounds'
+r.squaredGLMM(m6)
+
+```
+
+Explanatory power of models does not scale linearly with time, more an exponential relationship
+
+```{r, warning=F, message=F, echo=F}
+
+lin_r2<-data.frame(day=1:120, simr2=seq(0,1, length.out=120))
+
+dpred<-rbind(data.frame(t(data.frame(r.squaredGLMM(m1)))),
+             data.frame(t(data.frame(r.squaredGLMM(m2)))),
+             data.frame(t(data.frame(r.squaredGLMM(m3)))),
+             data.frame(t(data.frame(r.squaredGLMM(m5)))),
+             data.frame(t(data.frame(r.squaredGLMM(m6)))))
+dpred$mod<-c('Arrival at UK border',
+             'Arrival in Europe',
+             'Depart West Africa',
+             'Depart Central Africa',
+             'Depart wintering grounds')
+
+dpred$day=c(median(datfinal$arrive_uk, na.rm=T),
+            median(datfinal$DEPnorthAF, na.rm=T),
+            median(datfinal$DEPwestAF, na.rm=T),
+            median(datfinal$DEPcentralAF, na.rm=T),
+            median(datfinal$depart_winterSO, na.rm=T))
+
+qplot(data=lin_r2, x=day, y=simr2, geom='line')+
+  geom_point(data=dpred, aes(x=day, y=R2m), colour='red')+
+  geom_text(data=dpred, aes(x=day, y=R2m, label=mod),
+            nudge_y=-0.05)+ylab('Explanatory power (r2)')
+```
+
+Exploring the residuals for each of migration stage against arrival at breeding ground models. As expected residual variance increases further back in time,
+
+```{r, warning=F, message=F, echo=F}
+dfresid<-rbind(data.frame(residuals=resid(m1, type='pearson'),
+                    mig_region='UK arrive'), 
+               data.frame(residuals=resid(m2, type='pearson'),
+                          mig_region='Europe arrive'), 
+               data.frame(residuals=resid(m3, type='pearson'),
+                          mig_region='West Africa depart'), 
+               data.frame(residuals=resid(m5, type='pearson'),
+                          mig_region='Central Africa depart'), 
+               data.frame(residuals=resid(m6, type='pearson'),
+                          mig_region='Wintering depart'))
+
+ggplot(data=dfresid, aes(x=mig_region, y=residuals))+geom_jitter(shape=1, width=0.05)+
+  geom_hline(yintercept=0, colour='red', linetype=2)+
+  labs(x='Migration milestone', y='Residuals')+
+  scale_y_continuous(breaks=c(-15, -10, -5, 0, 5, 10, 15, 20, 25))+theme(axis.text.x=element_text(angle=90))
+```
+
+Partitioning the residual variance into certain regions allows us to see if different factors explain variance in certain regions, for example upland birds seem to travel faster through Europe than lowland birds.
+
+```{r, warning=F, message=F, echo=F}
+datfinal_nona<-na.omit(datfinal)
+
+m1<-lmer(arrive_breeding~arrive_uk+(1|breeding_hab/ptt)+(1|year), data=datfinal_nona)
+m2<-lmer(arrive_breeding~DEPnorthAF+(1|breeding_hab/ptt)+(1|year), data=datfinal_nona)
+m3<-lmer(arrive_breeding~DEPwestAF+(1|breeding_hab/ptt)+(1|year), data=datfinal_nona)
+
+m5<-lmer(arrive_breeding~DEPcentralAF+(1|breeding_hab/ptt)+(1|year), data=datfinal_nona)
+m6<-lmer(arrive_breeding~depart_winterSO+(1|breeding_hab/ptt)+(1|year), data=datfinal_nona)
+
+datfinal_nona$r.arriveUK<-resid(m1, type='pearson')
+datfinal_nona$r.arriveEU<-resid(m2, type='pearson')
+datfinal_nona$r.departWA<-resid(m3, type='pearson')
+datfinal_nona$r.departCA<-resid(m5, type='pearson')
+datfinal_nona$r.departWI<-resid(m6, type='pearson')
+
+#attrib variance to regions
+
+datfinal_nona$var.UK<-datfinal_nona$r.arriveUK
+datfinal_nona$var.europe<-datfinal_nona$r.arriveEU-datfinal_nona$r.arriveUK
+datfinal_nona$var.Sahara<-datfinal_nona$r.departWA-datfinal_nona$r.arriveEU
+datfinal_nona$var.Wafrica<-datfinal_nona$r.departCA-datfinal_nona$r.departWA
+datfinal_nona$var.Cafrica<-datfinal_nona$r.departWI-datfinal_nona$r.departCA
+
+var_yr<-melt(datfinal_nona[,c(2,12,19:23)], id.vars=c('year', 'breeding_hab'))
+
+qplot(data=var_yr, x=variable, y=value, colour=breeding_hab, geom='boxplot')+labs(y='migration speed (residual variance)', x='region')+theme(axis.text.x=element_text(angle=90))
+
+```
+
+Year can also explain some variance, for example in 2012 birds seem to move through all regions comparatbely faster than in other years.
+
+```{r, warning=F, message=F, echo=F}
+qplot(data=var_yr, x=variable, y=value, colour=factor(year), geom='boxplot')+labs(y='migration speed (residual variance)', x='region')+theme(axis.text.x=element_text(angle=90))
+```
+
+Back to the relationship between west africa and uk breeding: For every extra day departure from West Africa is delayed cuckoos arrive 0.78 days later back at UK breeding grounds.
+
+```{r, warning=F, message=F, echo=F}
+
 m4<-lmer(arrive_breeding~DEPwestAF+
@@
 newdat$p1<-predict(m4, newdata=newdat, re.form=~0)
-
  predmat<-model.matrix(arrive_breeding~DEPwestAF, data=newdat)
@@
 
-Cuckoos that departed West Africa early had longer migrations back to the breeding grounds but those that departed after median departure date (day 102) could not compensate for delayed departure with faster migration.
+cuckoos that departed West Africa early had longer migrations back to the breeding grounds but those that departed after median departure date (day 102) could not compensate for delayed departure with faster migration.
 
 ```{r, warning=F, message=F, echo=F}
-d1$migration.length<-d1$arrive.breeding.uk-d1$depart.west.africa
+
+dat$migration.length<-dat$arrive_breeding-dat$DEPwestAF
 
@@
 
-qplot(data=d1, x=depart.west.africa, y=migration.length)+
+qplot(data=dat, x=DEPwestAF, y=migration.length)+
   xlab("Departure date West Africa (DOY)")+
@@
 # warning ok
-
-qplot(data=dmod2, y=arrive_breeding, x=mean_diff, colour=breeding_loc)
 
@@
 ```
-
-Habitat Variables
-Habitat of the final stopover before departure from West Africa does not seem to have a huge imapact.
-The percentage of tree cover does not affect departure date.
-
-```{r, warning=F, message=F, echo=F} 
- m1<-lmer(DEPwestAF~Tree+(1|year)+(1|ptt), data=dat)
- #Anova(m1, test.statistic = 'F')
- ggplot(data=dat, aes(x=Tree, y=DEPwestAF ))+xlab("% tree cover of final stopover")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.27, p=0.609', x = -Inf, y = Inf), hjust = 0, vjust = 1)
- ```
- The percentage of grass cover does appear to affect departure date. *Not sure how believeable this is*
- ```{r, warning=F, message=F, echo=F} 
-m1<-lmer(DEPwestAF~Grass+(1|year)+(1|ptt), data=dat)
- #Anova(m1, test.statistic = 'F')
- newdat<-data.frame(DEPwestAF=1, Grass=1:75)
-newdat$p1<-predict(m1, newdata=newdat, re.form=~0)
- predmat<-model.matrix(DEPwestAF~Grass, data=newdat)
-vcv<-vcov(m1)
-semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
-newdat$lc<-newdat$p1-semod*1.96
-newdat$uc<-newdat$p1+semod*1.96
- qplot(data=dat, y=DEPwestAF, x=Grass)+
-  geom_line(data=newdat, aes(x=Grass, y=p1), colour='red')+
-  geom_line(data=newdat, aes(x=Grass, y=lc), colour='red', linetype='dashed')+
-  geom_line(data=newdat, aes(x=Grass, y=uc), colour='red', linetype='dashed')+
-  ylab("Departure date West Africa (DOY)")+
-  xlab("% grass cover of habitat")+
-  theme_bw()+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=7.54, p=0.011', x = -Inf, y = Inf), hjust = 0, vjust = 1)
-```
-
-Distance to rivers does not affect departure date.
-
-```{r, warning=F, message=F, echo=F} 
- m1<-lmer(DEPwestAF~D_riv+(1|year)+(1|ptt), data=dat)
- #Anova(m1, test.statistic = 'F')
- ggplot(data=dat, aes(x=D_riv, y=DEPwestAF ))+xlab("Distance to nearest river (m)")+ylab("Departure date West Africa (DOY)")+geom_point(shape=1)+theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=0.39, p=0.531', x = -Inf, y = Inf), hjust = 0, vjust = 1)
-```
-
-Dynamic environmental variables
-NDVI anomoly appears important. Greener final stopover sites allow earlier departure from West. *Now to test rainfall and greening in more detail*
-
-```{r, warning=F, message=F, echo=F}
- dat2.3<-dat2.2 %>% group_by(year, ptt) %>%
-  summarise_all(last)
- dat$year<-as.integer(dat$year)
- dat3<-full_join(dat, dat2.3, by=c('ptt', 'year'))
- m6<-lmer(DEPwestAF~aquaANOM+
-           (1|ptt)+(1|year), data=dat3)
- #summary(m6)
- # plot
- newdat<-data.frame(aquaANOM=seq(-0.1,0.15, 0.005), DEPwestAF=1)
-newdat$p1<-predict(m6, newdata=newdat, re.form=~0)
- predmat<-model.matrix(DEPwestAF~aquaANOM, data=newdat)
-vcv<-vcov(m6)
-semod<-sqrt(diag(predmat%*%vcv%*%t(predmat)))
-newdat$lc<-newdat$p1-semod*1.96
-newdat$uc<-newdat$p1+semod*1.96
- qplot(data=dat3, x=aquaANOM, y=DEPwestAF)+
-  geom_line(data=newdat, aes(x=aquaANOM, y=p1), colour='red')+
-  geom_line(data=newdat, aes(x=aquaANOM, y=lc), colour='red', linetype='dashed')+
-  geom_line(data=newdat, aes(x=aquaANOM, y=uc), colour='red', linetype='dashed')+
-  xlab("NDVI anomoly")+
-  ylab("Departure date West Africa (DOY)")+
-  theme_bw()+geom_label(data=data.frame(), aes(label = 'Anova F=4.65, p=0.038', x = -Inf, y = Inf), hjust = 0, vjust = 1)+theme(axis.text.x=element_text(angle=90))
- #Anova(m6, test.statistic = 'F')
-```

### 2018-09-20 paper_analyses.R

@@ -483,7 +483,30 @@ r.squaredGLMM(m6)
 #just little idea to see if explanatory power is a simple function of time
 # or actually per region
 
-lin_r2<-seq(0.07,0.88, length.out=120)
+lin_r2<-data.frame(day=1:120, simr2=seq(0,1, length.out=120))
+
+dpred<-rbind(data.frame(t(data.frame(r.squaredGLMM(m1)))),
+             data.frame(t(data.frame(r.squaredGLMM(m2)))),
+             data.frame(t(data.frame(r.squaredGLMM(m3)))),
+             data.frame(t(data.frame(r.squaredGLMM(m5)))),
+             data.frame(t(data.frame(r.squaredGLMM(m6)))))
+dpred$mod<-c('Arrival at UK border',
+             'Arrival in Europe',
+             'Depart West Africa',
+             'Depart Central Africa',
+             'Depart wintering grounds')
+
+dpred$day=c(median(datfinal$arrive_uk, na.rm=T),
+            median(datfinal$DEPnorthAF, na.rm=T),
+            median(datfinal$DEPwestAF, na.rm=T),
+            median(datfinal$DEPcentralAF, na.rm=T),
+            median(datfinal$depart_winterSO, na.rm=T))
+            
+
+qplot(data=lin_r2, x=day, y=simr2, geom='line')+
+  geom_point(data=dpred, aes(x=day, y=R2m), colour='red')+
+  geom_text(data=dpred, aes(x=day, y=R2m, label=mod),
+            nudge_y=-0.05)
 
 median(dmod$DEPcentralAF, na.rm=T)
 lin_r2[70]
@@ -513,7 +536,7 @@ dmod_nona$var.Sahara<-dmod_nona$r.departWA-dmod_nona$r.arriveEU
 dmod_nona$var.Wafrica<-dmod_nona$r.departCA-dmod_nona$r.departWA
 dmod_nona$var.Cafrica<-dmod_nona$r.departWI-dmod_nona$r.departCA
 
-var_yr<-melt(dmod_nona[,c(2,19:23)], id.vars=c('year'))
+var_yr<-melt(dmod_nona[,c(2,12,19:23)], id.vars=c('year', 'breeding_hab'))
 
 qplot(data=var_yr, x=variable, y=value, colour=factor(year), geom='boxplot')
 
